---
title: "P8131 Homework 10"
author: "Yongyan Liu (yl6107)"
date: "2026-04-25"
output:
  pdf_document:
    latex_engine: xelatex
header-includes:
  - \usepackage{amsmath}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 6, fig.height = 4)
library(survival)
library(KMsurv)
```

# Problem 1

**Suppose we have the following data:**

- **Group 1: 4, 12+, 15, 21+, 23**
- **Group 2: 2, 6+, 8+, 10, 19**

**($+$ denotes right-censoring.)
Test $H_{0}\!:\,h_{1}(t)=h_{2}(t)\;\forall t$ versus
$H_{1}\!:\,h_{1}(t)\neq h_{2}(t)$ for some $t$, using the log-rank test.**

**Hypotheses (Lec 22, slide 6).**
$$
H_{0}: S_{1}(t)=S_{2}(t)\ \text{for all }t
\quad\Bigl(\Leftrightarrow\ h_{1}(t)=h_{2}(t)\Bigr)
\qquad\text{vs.}\qquad
H_{1}: S_{1}(t)\neq S_{2}(t)\ \text{for some }t.
$$

At each *uncensored* pooled event time $t_{i}$, expected count and hypergeometric variance under $H_{0}$ are
$$
E(d_{1,i}) \;=\; \frac{m_{D,i}\,n_{1,i}}{n_{i}},
\qquad
\mathrm{Var}(d_{1,i}) \;=\; \frac{n_{1,i}\,n_{2,i}\,m_{D,i}\,m_{\bar D,i}}
                                  {n_{i}^{2}\,(n_{i}-1)}.
$$

The pooled uncensored event times are $\{2,4,10,15,19,23\}$.
The per-event-time table is

\begin{center}
\begin{tabular}{ccccccccc}
\hline
$t_{i}$ & $n_{1,i}$ & $n_{2,i}$ & $n_{i}$ & $d_{1,i}$ & $d_{2,i}$ & $m_{D,i}$ & $E(d_{1,i})$ & $\mathrm{Var}(d_{1,i})$ \\
\hline
 2 & 5 & 5 & 10 & 0 & 1 & 1 & 0.5000 & 0.2500 \\
 4 & 5 & 4 &  9 & 1 & 0 & 1 & 0.5556 & 0.2469 \\
10 & 4 & 2 &  6 & 0 & 1 & 1 & 0.6667 & 0.2222 \\
15 & 3 & 1 &  4 & 1 & 0 & 1 & 0.7500 & 0.1875 \\
19 & 2 & 1 &  3 & 0 & 1 & 1 & 0.6667 & 0.2222 \\
23 & 1 & 0 &  1 & 1 & 0 & 1 & 1.0000 & 0.0000 \\
\hline
\textbf{Sum} & & & & \textbf{3} & & & \textbf{4.1389} & \textbf{1.1288} \\
\hline
\end{tabular}
\end{center}

(The variance at $t=23$ is $0$ because only one subject is at risk; $n_{i}-1=0$ in the
denominator, but $n_{1,i}n_{2,i}=0$ in the numerator dominates and the contribution is $0$.)

The log-rank statistic is
$$
Z^{2} \;=\; \frac{\bigl(\sum_{i}d_{1,i}-\sum_{i}E(d_{1,i})\bigr)^{2}}
                  {\sum_{i}\mathrm{Var}(d_{1,i})}
\;=\; \frac{(3-4.1389)^{2}}{1.1288}
\;=\; \frac{1.2972}{1.1288}
\;\approx\; 1.149,
\qquad Z^{2}\sim\chi^{2}_{1}.
$$

The p-value is $P(\chi^{2}_{1}>1.149)\approx 0.284$.


# Problem 2

**Consider the `kidtran` dataset in the `KMsurv` package, where the death
times of kidney transplant patients are reported. The patients can be
classified by race (1 = white, 2 = black) and gender (1 = male, 2 = female)
into one of four groups. Compare the survival curves for different races
in each sex group.**

Per the question, the analysis is run *separately within each sex* — a
two-sample race comparison among males, and another among females. For
each sex we plot Kaplan-Meier curves for the two race groups and run a
log-rank test.

```{r p2-load}
data(kidtran)
str(kidtran)
male   <- subset(kidtran, gender == 1)
female <- subset(kidtran, gender == 2)
```

## Males: KM curves and log-rank by race

```{r p2-male-km, fig.cap="Kaplan-Meier survival by race, males only. Tick marks denote right-censored observations."}
fit_m <- survfit(Surv(time, delta) ~ race, data = male)
plot(fit_m, mark.time = TRUE, lty = 1, col = c("black", "red"),
     xlab = "Time (days)", ylab = "Survival probability",
     main = "Males: KM by race")
legend("bottomleft", c("Race 1 (white)", "Race 2 (black)"),
       lty = 1, col = c("black", "red"), bty = "n")
```

```{r p2-male-lr}
survdiff(Surv(time, delta) ~ race, data = male)
```

**Hypotheses:** $H_{0}: S_{\text{white}}(t)=S_{\text{black}}(t)$ for all
$t$ within males vs $H_{1}$: not equal for some $t$.

Test statistic $\chi^{2}=0.10$ on $\mathrm{df}=1$, $p\approx 0.76$.
**Do not reject $H_{0}$**. Among males, there is no evidence of a
race difference in survival; the two KM curves track each other closely.

## Females: KM curves and log-rank by race

```{r p2-female-km, fig.cap="Kaplan-Meier survival by race, females only. Tick marks denote right-censored observations."}
fit_f <- survfit(Surv(time, delta) ~ race, data = female)
plot(fit_f, mark.time = TRUE, lty = 1, col = c("black", "red"),
     xlab = "Time (days)", ylab = "Survival probability",
     main = "Females: KM by race")
legend("bottomleft", c("Race 1 (white)", "Race 2 (black)"),
       lty = 1, col = c("black", "red"), bty = "n")
```

```{r p2-female-lr}
survdiff(Surv(time, delta) ~ race, data = female)
```

**Hypotheses:** $H_{0}: S_{\text{white}}(t)=S_{\text{black}}(t)$ for all
$t$ within females vs $H_{1}$: not equal for some $t$.

Test statistic $\chi^{2}=4.8$ on $\mathrm{df}=1$, $p\approx 0.028$.
**Reject $H_{0}$** at the $5\%$ level. Among females, survival differs
significantly by race: black female recipients had more observed deaths
than expected under $H_{0}$ ($14$ vs $8.21$), so they have *worse*
survival than white female recipients, consistent with the KM plot in
which the race-2 curve drops below the race-1 curve.


# Problem 3

**Consider the `larynx` data in the `KMsurv` package. Let**

- **$Z_{1}=1$ if the patient is in stage II, $0$ otherwise;**
- **$Z_{2}=1$ if the patient is in stage III, $0$ otherwise;**
- **$Z_{3}=1$ if the patient is in stage IV, $0$ otherwise;**
- **$Z_{4}=$ age (in years).**

**Fit a proportional-hazards model with $Z_{1},Z_{2},Z_{3},Z_{4}$ and the
interaction $Z_{1}\!\cdot\!Z_{4}$. Explain the results. What is the
relative risk of dying for a stage-II patient of age $50$ compared to a
stage-I patient of the same age?**

Stage I is the reference category.

## Cox PH fit

```{r p3-fit}
data(larynx)
larynx$Z1 <- as.numeric(larynx$stage == 2)  # stage II indicator
larynx$Z2 <- as.numeric(larynx$stage == 3)  # stage III indicator
larynx$Z3 <- as.numeric(larynx$stage == 4)  # stage IV indicator
larynx$Z4 <- larynx$age

fit <- coxph(Surv(time, delta) ~ Z1 + Z2 + Z3 + Z4 + Z1:Z4,
             data = larynx, ties = "breslow")
summary(fit)
```

The fitted model is
$$
h(t \mid Z) \;=\; h_{0}(t)\,
\exp\!\bigl\{\beta_{1}Z_{1}+\beta_{2}Z_{2}+\beta_{3}Z_{3}
            +\beta_{4}Z_{4}+\beta_{14}Z_{1}Z_{4}\bigr\}.
$$

## Interpretation

Recall that hazard ratios are constant in time under the PH assumption
(Lec 23, slide 6) but here vary across covariate combinations because
of the interaction. From the coefficient table:

- **$\hat\beta_{2}=0.622$, $\exp(\hat\beta_{2})=1.86$** ($p=0.08$).
  Stage III versus stage I patients have $1.86$ times the hazard of
  death at *any* age (no interaction with age was fit for stage III).
- **$\hat\beta_{3}=1.753$, $\exp(\hat\beta_{3})=5.77$** ($p<0.001$).
  Stage IV versus stage I patients have $5.77$ times the hazard at any
  age — stage IV is highly significantly worse than stage I.
- **$\hat\beta_{4}=0.0060$, $\exp(\hat\beta_{4})=1.006$** ($p=0.69$).
  Because $Z_{1}\!\cdot\!Z_{4}$ is in the model, $\beta_{4}$ is the
  per-year log hazard ratio for *non-stage-II* patients (stages I, III,
  IV); a one-year increase in age multiplies their hazard by $1.006$,
  not significantly different from $1$.
- **$\hat\beta_{1}=-7.382$, $\exp(\hat\beta_{1})=6.2\times 10^{-4}$**
  ($p=0.03$). This is *not* the stage II vs stage I hazard ratio in
  general; it is the HR at age $0$, which is far outside the range of
  the data. Because of the $Z_{1}\!\cdot\!Z_{4}$ interaction, the stage
  II vs stage I log hazard ratio at age $a$ equals
  $\beta_{1}+a\,\beta_{14}$.
- **$\hat\beta_{14}=0.1117$, $\exp(\hat\beta_{14})=1.118$** ($p=0.02$).
  Each additional year of age multiplies the stage-II-vs-stage-I hazard
  ratio by $1.12$. Equivalently, for stage II patients, the per-year
  log hazard ratio is $\beta_{4}+\beta_{14}=0.118$, so a one-year
  increase in age multiplies a stage-II patient's hazard by
  $\exp(0.118)\approx 1.12$. The age effect is therefore much stronger
  in stage II than in the other stages, and the stage-II-vs-stage-I
  gap *widens* with increasing age.

The omnibus likelihood-ratio test ($24.1$ on $5$ df, $p=2\times 10^{-4}$)
indicates that the covariate set is jointly highly significant.

## Relative risk: stage II at age 50 vs stage I at age 50

Both patients have $Z_{2}=Z_{3}=0$, $Z_{4}=50$. Their covariate vectors are

- Stage I: $(Z_{1},Z_{2},Z_{3},Z_{4},Z_{1}Z_{4})=(0,0,0,50,0)$;
- Stage II: $(Z_{1},Z_{2},Z_{3},Z_{4},Z_{1}Z_{4})=(1,0,0,50,50)$.

The log hazard ratio is therefore
$$
\log\mathrm{RR} \;=\; \beta_{1}\,(1-0)+\beta_{14}\,(50-0)
                 \;=\; \beta_{1}+50\,\beta_{14}.
$$
Plugging in the estimates,
$$
\widehat{\log\mathrm{RR}}
\;=\; -7.3820 + 50\,(0.1117)
\;=\; -1.7986,
$$
$$
\widehat{\mathrm{RR}}\;=\;\exp(-1.7986)\;\approx\;0.166.
$$

```{r p3-rr}
b1  <- coef(fit)["Z1"]
b14 <- coef(fit)["Z1:Z4"]
logRR <- b1 + 50 * b14
RR    <- exp(logRR)
c(logRR = unname(logRR), RR = unname(RR))
```

**Answer.** A stage II patient aged $50$ has an estimated hazard about
$\widehat{\mathrm{RR}}\approx 0.17$ times that of a stage I patient
aged $50$ — i.e., at age $50$ a stage II patient has *lower* mortality
hazard than a stage I patient. (Because $\hat\beta_{14}>0$, this gap
shrinks at older ages and reverses around age $\beta_{1}/(-\beta_{14})
\approx 66$ years, beyond which stage II patients have higher hazard
than stage I patients.)
