---
title: "Data Science II Homework 3"
author: "Yongyan Liu (yl6107)"
date: "2026-03-25"
output:
  pdf_document:
    latex_engine: xelatex
header-includes:
- \usepackage{fancyhdr}
- \usepackage{lipsum}
- \pagestyle{fancy}
- \fancyhead[R]{\thepage}
- \fancypagestyle{plain}{\pagestyle{fancy}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

Split the dataset into two parts: training data (70%) and test data (30%).

```{r}
library(tidyverse)
library(rsample)
library(caret)
library(earth)
library(MASS)
library(pROC)
library(pdp)
library(vip)
library(corrplot)

auto <- read.csv("auto.csv") %>%
  mutate(mpg_cat = factor(mpg_cat, levels = c("low", "high")),
         origin = factor(origin))

set.seed(2026)
data_split <- initial_split(auto, prop = 0.7, strata = mpg_cat)
train_data <- training(data_split)
test_data <- testing(data_split)

ctrl <- trainControl(method = "repeatedcv", number = 10, repeats = 5,
                     summaryFunction = twoClassSummary,
                     classProbs = TRUE)
```

# (a) Logistic Regression

```{r}
glm.fit <- glm(mpg_cat ~ ., data = train_data, family = binomial(link = "logit"))
summary(glm.fit)
```

```{r}
# Fit logistic regression using caret for later comparison
set.seed(2026)
model.glm <- train(mpg_cat ~ .,
                   data = train_data,
                   method = "glm",
                   metric = "ROC",
                   trControl = ctrl)
```

We examine the correlation structure among numeric predictors:

```{r, fig.height=5, fig.width=6}
num_vars <- train_data[, sapply(train_data, is.numeric)]
corrplot(cor(num_vars), method = "circle", type = "full", addCoef.col = "black",
         tl.cex = 0.8, number.cex = 0.7)
```

Yes, there are redundant predictors. The correlation plot shows that `cylinders`, `displacement`, `horsepower`, and `weight` are highly correlated with one another. In the logistic regression summary, `cylinders`, `displacement`, `horsepower`, and `acceleration` have large p-values (> 0.05), indicating they do not contribute significant additional information beyond the other predictors. These variables are redundant due to multicollinearity — they carry overlapping information about engine size and vehicle performance.

# (b) MARS Model

```{r, fig.height=4, fig.width=6}
mars_grid <- expand.grid(degree = 1:3, nprune = 2:20)

set.seed(2026)
model.mars <- train(mpg_cat ~ .,
                    data = train_data,
                    method = "earth",
                    tuneGrid = mars_grid,
                    metric = "ROC",
                    trControl = ctrl,
                    glm = list(family = binomial))

ggplot(model.mars) + theme_bw()
model.mars$bestTune
coef(model.mars$finalModel)
```

```{r, fig.height=4, fig.width=6}
vip(model.mars$finalModel, type = "nsubsets")
```

```{r}
# Compare CV performance
res_ab <- resamples(list(GLM = model.glm, MARS = model.mars))
summary(res_ab)
bwplot(res_ab, metric = "ROC")
```

```{r}
# Compare test set performance
glm.pred.b <- predict(model.glm, newdata = test_data, type = "prob")[, 2]
mars.pred.b <- predict(model.mars, newdata = test_data, type = "prob")[, 2]
roc.glm.b <- roc(test_data$mpg_cat, glm.pred.b)
roc.mars.b <- roc(test_data$mpg_cat, mars.pred.b)
auc.glm.b <- roc.glm.b$auc[1]
auc.mars.b <- roc.mars.b$auc[1]
cat("Test AUC - GLM:", round(auc.glm.b, 4), "\n")
cat("Test AUC - MARS:", round(auc.mars.b, 4), "\n")
```

Based on both the cross-validation and test set AUC, the MARS model achieves a comparable (or slightly higher) AUC relative to logistic regression.

# (c) Linear Discriminant Analysis

```{r}
set.seed(2026)
model.lda <- train(mpg_cat ~ .,
                   data = train_data,
                   method = "lda",
                   metric = "ROC",
                   trControl = ctrl)
model.lda$finalModel
```

Since there are $K = 2$ classes, there is only one linear discriminant. We plot the histogram of the discriminant scores for each class:

```{r, fig.height=6, fig.width=6}
x_train <- model.matrix(mpg_cat ~ ., data = train_data)[, -1]
lda.scores <- predict(model.lda$finalModel, x_train)$x[, 1]
ldahist(lda.scores, g = train_data$mpg_cat)
```

# (d) Model Comparison and ROC Curve

```{r}
# Compare all three models via CV
res_all <- resamples(list(GLM = model.glm, MARS = model.mars, LDA = model.lda))
summary(res_all)
bwplot(res_all, metric = "ROC")
```

```{r, fig.height=5, fig.width=6}
# Test set predictions
glm.pred <- predict(model.glm, newdata = test_data, type = "prob")[, 2]
mars.pred <- predict(model.mars, newdata = test_data, type = "prob")[, 2]
lda.pred <- predict(model.lda, newdata = test_data, type = "prob")[, 2]

# ROC curves
roc.glm <- roc(test_data$mpg_cat, glm.pred)
roc.mars <- roc(test_data$mpg_cat, mars.pred)
roc.lda <- roc(test_data$mpg_cat, lda.pred)

auc <- c(roc.glm$auc[1], roc.mars$auc[1], roc.lda$auc[1])
modelNames <- c("GLM", "MARS", "LDA")

ggroc(list(roc.glm, roc.mars, roc.lda), legacy.axes = TRUE) +
  scale_color_discrete(labels = paste0(modelNames, " (", round(auc, 3), ")"),
                       name = "Models (AUC)") +
  geom_abline(intercept = 0, slope = 1, color = "grey") +
  theme_bw()
```

```{r}
cat("Selected model:", modelNames[which.max(auc)],
    "with test AUC:", round(max(auc), 4), "\n")

# Number of original predictors used by each model
n_glm <- ncol(model.glm$finalModel$model) - 1
n_mars <- length(evimp(model.mars$finalModel)[, 1])
n_lda <- nrow(model.lda$finalModel$scaling)
cat("\nNumber of predictors used:\n")
cat("GLM:", n_glm, "\nMARS:", n_mars, "\nLDA:", n_lda, "\n")
```

The three models have comparable test AUC values, but differ in complexity. GLM and LDA use all predictors, while MARS performs automatic variable selection and retains fewer terms. Since the MARS model achieves the highest test AUC with a simpler model structure, it is preferred on the grounds of both predictive accuracy and parsimony. Therefore, we select MARS as our final model.

# (e) Confusion Matrix

We use Youden's J statistic to select the optimal classification threshold, which maximizes the sum of sensitivity and specificity. 

```{r}
threshold <- coords(roc.mars, "best", ret = "threshold",
                    best.method = "youden")$threshold
cat("Optimal threshold (Youden's J):", round(threshold, 3), "\n")

test.pred <- factor(ifelse(mars.pred > threshold, "high", "low"),
                    levels = c("low", "high"))

confusionMatrix(data = test.pred,
                reference = test_data$mpg_cat,
                positive = "high")
```

The confusion matrix indicates that our selected MARS model has strong predictive power. By applying the Youden-derived threshold, we achieved a high rate of correct classifications for both 'high' and 'low' mpg cars. The minimal number of misclassifications suggests the model generalizes well to new data, and the balanced accuracy proves it is reliable across both categories.
