---
title: "P8131 Homework 2"
author: "Yongyan Liu (yl6107)"
date: "2026-02-19"
output:
  pdf_document:
    latex_engine: xelatex
    extra_dependencies: ["amsmath", "amssymb", "amsthm"]
header-includes:
  - \usepackage{bm}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(knitr)
```

# Problem 1: Bioassay Study

**The table below gives the data collected from a bioassay study in
which** $X$ variable (treated as continuous variable) is the
concentration level. At each of five different dose levels (0-4), 30
animals are tested and the number of dying are recorded.

|   Dose ($X$)    |  0  |  1  |  2  |  3  |  4  |
|:---------------:|:---:|:---:|:---:|:---:|:---:|
| Number of dying |  2  |  8  | 15  | 23  | 27  |

**Fit the model** $g(P(\text{dying})) = \alpha + \beta X$, with logit,
probit, and complementary log-log links.

```{r}
# Input data
dose <- 0:4
dying <- c(2, 8, 15, 23, 27)
n <- rep(30, 5)

bioassay <- data.frame(dose = dose, dying = dying, n = n, surviving = n - dying)

# Fit models with different link functions
fit_logit <- glm(cbind(dying, surviving) ~ dose, 
                 family = binomial(link = "logit"), data = bioassay)
fit_probit <- glm(cbind(dying, surviving) ~ dose, 
                  family = binomial(link = "probit"), data = bioassay)
fit_cloglog <- glm(cbind(dying, surviving) ~ dose, 
                   family = binomial(link = "cloglog"), data = bioassay)
```

## (a) Fill out the table and give comments

```{r}
# Function to extract results
get_results <- function(fit, link_name) {
  beta <- coef(fit)[2]
  se_beta <- summary(fit)$coefficients[2, 2]
  ci_beta <- confint.default(fit)[2, ]  # Wald CI
  dev <- fit$deviance

  # Prediction at x = 0.01
  newdata <- data.frame(dose = 0.01)
  p_pred <- predict(fit, newdata = newdata, type = "response")

  return(c(
    Model = link_name,
    Estimate_beta = round(beta, 4),
    CI_lower = round(ci_beta[1], 4),
    CI_upper = round(ci_beta[2], 4),
    Deviance = round(dev, 4),
    p_dying = round(p_pred, 4)
  ))
}

# Extract results for all models
results_logit <- get_results(fit_logit, "logit")
results_probit <- get_results(fit_probit, "probit")
results_cloglog <- get_results(fit_cloglog, "c-log-log")

# Combine into a table
results_table <- rbind(results_logit, results_probit, results_cloglog)
results_df <- as.data.frame(results_table)
colnames(results_df) <- c("Model", "Estimate of $\\beta$", "95% CI Lower", "95% CI Upper",
                          "Deviance", "$\\hat{p}$(dying|x=0.01)")
rownames(results_df) <- NULL

kable(results_df, caption = "Model Comparison for Bioassay Data")
```

**Comments:**

1.  **Coefficient estimates:** All three models show positive
    $\hat{\beta}$, indicating that higher dose increases the probability
    of dying. The logit model has the largest $\hat{\beta}$ (1.1619),
    followed by probit (0.6864), and c-log-log (0.7468). This difference
    is expected due to different scales of the link functions.

2.  **Deviance:** The deviance measures the goodness of fit (smaller is
    better). All three models fit reasonably well:

    -   Logit: 0.3787
    -   Probit: 0.3137
    -   C-log-log: 2.2305

    With $n - p = 5 - 2 = 3$ degrees of freedom and
    $\chi^2_{0.95}(3) = 7.81$, all deviances are well below the critical
    value, suggesting adequate fit. The probit model has the smallest
    deviance, followed by logit. The c-log-log model has noticeably
    larger deviance.

3.  **Predicted probability at** $x = 0.01$: The three models give
    similar predictions at low doses:

    -   Logit: 0.0729
    -   Probit: 0.0711
    -   C-log-log: 0.0533

    The c-log-log model predicts a slightly lower probability,
    reflecting its asymmetric nature.

4.  **Model selection:** Based on deviance, the probit and logit models
    fit better than c-log-log for this symmetric dose-response data. The
    logit model is often preferred due to interpretability (odds ratio).

------------------------------------------------------------------------

## (b) LD50 Estimation with 90% Confidence Interval

**Suppose that the dose level is in natural logarithm scale, estimate
LD50 with 90% confidence interval based on the three models.**

The LD50 is the dose at which $P(\text{dying}) = 0.5$.

For a model $g(\pi) = \alpha + \beta X$:

-   **Logit:**
    $\log\frac{0.5}{0.5} = 0 = \alpha + \beta \cdot \text{LD50} \Rightarrow \text{LD50} = -\frac{\alpha}{\beta}$

-   **Probit:**
    $\Phi^{-1}(0.5) = 0 = \alpha + \beta \cdot \text{LD50} \Rightarrow \text{LD50} = -\frac{\alpha}{\beta}$

-   **C-log-log:**
    $\log(-\log(0.5)) = -0.3665 = \alpha + \beta \cdot \text{LD50} \Rightarrow \text{LD50} = \frac{-0.3665 - \alpha}{\beta}$

Since dose is on log scale, the actual LD50 in original scale is
$e^{\text{LD50}}$.

```{r}
# Function to calculate LD50 and CI using Delta method
calc_LD50 <- function(fit, link_type, alpha_level = 0.10) {
  coefs <- coef(fit)
  vcov_mat <- vcov(fit)

  a <- coefs[1]  # intercept
  b <- coefs[2]  # slope

  if (link_type == "cloglog") {
    c_val <- log(-log(0.5))  # ≈ -0.3665
    LD50 <- (c_val - a) / b
    # Gradient for delta method: d(LD50)/d(a) = -1/b, d(LD50)/d(b) = -(c-a)/b^2
    grad <- c(-1/b, -(c_val - a)/b^2)
  } else {
    # For logit and probit, g(0.5) = 0
    LD50 <- -a / b
    # Gradient: d(LD50)/d(a) = -1/b, d(LD50)/d(b) = a/b^2
    grad <- c(-1/b, a/b^2)
  }

  # Variance using delta method
  var_LD50 <- t(grad) %*% vcov_mat %*% grad
  se_LD50 <- sqrt(var_LD50)

  # CI for log(dose)
  z_val <- qnorm(1 - alpha_level/2)
  ci_lower <- LD50 - z_val * se_LD50
  ci_upper <- LD50 + z_val * se_LD50

  # Transform to original scale (exponentiate)
  LD50_orig <- exp(LD50)
  ci_lower_orig <- exp(ci_lower)
  ci_upper_orig <- exp(ci_upper)

  return(list(
    LD50_log = LD50,
    SE_log = se_LD50,
    CI_log = c(ci_lower, ci_upper),
    LD50_orig = LD50_orig,
    CI_orig = c(ci_lower_orig, ci_upper_orig)
  ))
}

# Calculate LD50 for each model
LD50_logit <- calc_LD50(fit_logit, "logit")
LD50_probit <- calc_LD50(fit_probit, "probit")
LD50_cloglog <- calc_LD50(fit_cloglog, "cloglog")

# Create summary table
LD50_summary <- data.frame(
  Model = c("Logit", "Probit", "C-log-log"),
  LD50_log_scale = c(LD50_logit$LD50_log, LD50_probit$LD50_log, LD50_cloglog$LD50_log),
  SE = c(LD50_logit$SE_log, LD50_probit$SE_log, LD50_cloglog$SE_log),
  CI_lower_log = c(LD50_logit$CI_log[1], LD50_probit$CI_log[1], LD50_cloglog$CI_log[1]),
  CI_upper_log = c(LD50_logit$CI_log[2], LD50_probit$CI_log[2], LD50_cloglog$CI_log[2]),
  LD50_original = c(LD50_logit$LD50_orig, LD50_probit$LD50_orig, LD50_cloglog$LD50_orig),
  CI_lower_orig = c(LD50_logit$CI_orig[1], LD50_probit$CI_orig[1], LD50_cloglog$CI_orig[1]),
  CI_upper_orig = c(LD50_logit$CI_orig[2], LD50_probit$CI_orig[2], LD50_cloglog$CI_orig[2])
)

kable(LD50_summary, digits = 4,
      col.names = c("Model", "LD50 (log)", "SE", "90% CI Lower (log)", "90% CI Upper (log)",
                    "LD50 (original)", "90% CI Lower (orig)", "90% CI Upper (orig)"),
      caption = "LD50 Estimates with 90% Confidence Intervals")
```

**Interpretation:**

The logit and probit models give very similar LD50 estimates, while the
c-log-log model gives a lower estimate. This reflects the asymmetric
nature of the c-log-log link function.

# Problem 2: MPH Enrollment Data

**The table below contains the enrollment data of some MPH program in a
year.**

```{r}
# Input enrollment data
enrollment <- data.frame(
  amount = c(10, 15, 20, 25, 30, 35, 40, 45, 50, 55, 60, 65, 70, 75, 80, 85, 90),
  offers = c(4, 6, 10, 12, 39, 36, 22, 14, 10, 12, 8, 9, 3, 1, 5, 2, 1),
  enrolls = c(0, 2, 4, 2, 12, 14, 10, 7, 5, 5, 3, 5, 2, 0, 4, 2, 1)
)
enrollment$not_enrolls <- enrollment$offers - enrollment$enrolls

kable(enrollment[, 1:3], caption = "MPH Enrollment Data",
      col.names = c("Amount (thousand $)", "Offers", "Enrolls"))
```

**Fit a logistic regression model:**

```{r}
# Fit logistic regression
fit_enroll <- glm(cbind(enrolls, not_enrolls) ~ amount,
                  family = binomial(link = "logit"), data = enrollment)
summary(fit_enroll)
```

## (a) How does the model fit the data?

```{r}
# Goodness of fit
dev <- fit_enroll$deviance
df <- fit_enroll$df.residual
p_value_dev <- 1 - pchisq(dev, df)

# Pearson chi-squared
pearson_resid <- residuals(fit_enroll, type = "pearson")
pearson_chisq <- sum(pearson_resid^2)
p_value_pearson <- 1 - pchisq(pearson_chisq, df)

# Hosmer-Lemeshow test (more appropriate for small group sizes)
library(ResourceSelection)
enroll_individual <- enrollment[rep(seq_len(nrow(enrollment)), enrollment$offers), ]
enroll_individual$y <- unlist(
  mapply(function(e, n) c(rep(1, e), rep(0, n - e)),
         enrollment$enrolls, enrollment$offers))
enroll_individual$fitted <- predict(
  fit_enroll, type = "response")[rep(seq_len(nrow(enrollment)), enrollment$offers)]
hl_test <- hoslem.test(enroll_individual$y, enroll_individual$fitted, g = 10)

cat("Deviance Chi-squared:", round(qchisq(0.95, df), 4), "\n")
cat("Deviance P-value:", round(p_value_dev, 4), "\n")
cat("Pearson Chi-squared:", round(pearson_chisq, 4), "\n")
cat("Pearson P-value:", round(p_value_pearson, 4), "\n")
cat("Hosmer-Lemeshow Chi-squared:", round(hl_test$statistic, 4), "\n")
cat("Hosmer-Lemeshow P-value:", round(hl_test$p.value, 4), "\n")
```

**Assessment of model fit:**

1.  **Deviance test:** The residual deviance is `r round(dev, 2)` with
    `r df` degrees of freedom. The critical value
    $\chi^2_{0.95}(15) = 25.00$. Since the deviance (`r round(dev, 2)`)
    is less than the critical value, we fail to reject the null
    hypothesis that the model fits adequately. The p-value is
    `r round(p_value_dev, 4)`, further supporting good fit.

2.  **Pearson's** $\chi^2$: The Pearson chi-squared statistic is
    `r round(pearson_chisq, 2)` with p-value
    `r round(p_value_pearson, 4)`, also indicating adequate fit.

3.  **Estimated dispersion parameter:**
    $\hat{\phi} = \frac{G}{n-p} = \frac{`r round(pearson_chisq, 2)`}{15} = `r round(pearson_chisq/df, 4)`$,
    which is less than 1, suggesting no significant overdispersion.

4.  **Hosmer-Lemeshow test:** Since several scholarship amount groups have
very few observations, we also did the HL test, whose statistic is
    `r round(hl_test$statistic, 2)` with p-value
    `r round(hl_test$p.value, 4)`, which also suggests adequate model
    fit.
    
**Visual Inspection**.

```{r, fig.height=4, fig.width=6}
# Diagnostic plot
enrollment$fitted_prob <- fitted(fit_enroll)
enrollment$observed_prob <- enrollment$enrolls / enrollment$offers

ggplot(enrollment, aes(x = amount)) +
  geom_point(aes(y = observed_prob), color = "blue", size = 2) +
  geom_line(aes(y = fitted_prob), color = "red") +
  labs(x = "Scholarship Amount (thousand $)", y = "Enrollment Rate",
       title = "Observed vs Fitted Enrollment Rates") +
  theme_minimal()
```

**Conclusion:** The logistic regression model fits the data adequately.
This conclusion is supported by multiple assessments: (1) formal tests
(Deviance, Pearson, and Hosmer-Lemeshow) all yield non-significant
p-values, and (2) visual inspection of the fitted curve against observed
proportions (see plot below) shows reasonable agreement between the
model and data.


------------------------------------------------------------------------

## (b) Relationship between scholarship amount and enrollment rate

```{r}
# Extract coefficient for amount
beta_hat <- coef(fit_enroll)[2]
se_beta <- summary(fit_enroll)$coefficients[2, 2]

# 95% CI for beta
ci_beta <- confint.default(fit_enroll)[2, ]

# Odds ratio and its CI
OR <- exp(beta_hat)
ci_OR <- exp(ci_beta)

cat("Coefficient (beta):", round(beta_hat, 6), "\n")
cat("95% CI for beta:", round(ci_beta, 6), "\n")
cat("Odds Ratio (per $1,000 increase):", round(OR, 4), "\n")
cat("95% CI for Odds Ratio:", round(ci_OR, 4), "\n")
```

**Relationship between scholarship amount and enrollment rate:**

There is a **statistically significant positive relationship** between scholarship amount and enrollment rate (p-value = `r round(summary(fit_enroll)$coefficients[2, 4], 4)`).

For every \$1,000 increase in scholarship amount, the odds of enrollment increase by a factor of $e^{`r round(beta_hat, 4)`} = `r round(OR, 4)`$, or approximately `r round((OR - 1) * 100, 2)`%.

**95% Confidence Interval:**

- 95% CI for $\beta$: $(`r round(ci_beta[1], 4)`, `r round(ci_beta[2], 4)`)$
- 95% CI for Odds Ratio: $(`r round(ci_OR[1], 4)`, `r round(ci_OR[2], 4)`)$

Since the 95% CI for the odds ratio does not include 1, the positive effect of scholarship amount on enrollment is statistically significant at the 0.05 level.

------------------------------------------------------------------------

## (c) Scholarship amount for 40% yield rate

We want to find the scholarship amount $X^*$ such that $\pi = 0.4$.

From the logistic model:
$$\log\left(\frac{0.4}{0.6}\right) = \alpha + \beta X^*$$
$$X^* = \frac{\log(0.4/0.6) - \alpha}{\beta} = \frac{-0.4055 - \alpha}{\beta}$$

```{r}
# Calculate scholarship amount for 40% yield rate
alpha_hat <- coef(fit_enroll)[1]
beta_hat <- coef(fit_enroll)[2]
vcov_mat <- vcov(fit_enroll)

target_prob <- 0.4
eta_target <- log(target_prob / (1 - target_prob))  # = log(0.4/0.6) ≈ -0.4055

X_star <- (eta_target - alpha_hat) / beta_hat

# Delta method for CI
# X* = (c - alpha) / beta where c = eta_target
# dX*/d(alpha) = -1/beta
# dX*/d(beta) = -(c - alpha)/beta^2
grad <- c(-1/beta_hat, -(eta_target - alpha_hat)/beta_hat^2)
var_X_star <- t(grad) %*% vcov_mat %*% grad
se_X_star <- sqrt(var_X_star)

# 95% CI
z_val <- qnorm(0.975)
ci_lower <- X_star - z_val * se_X_star
ci_upper <- X_star + z_val * se_X_star

cat("Target yield rate:", target_prob, "\n")
cat("Required scholarship amount: $", round(X_star * 1000, 0), "\n")
cat("Standard error: $", round(se_X_star * 1000, 0), "\n")
cat("95% CI: ($", round(ci_lower * 1000, 0), ", $", round(ci_upper * 1000, 0), ")\n")
```

**Answer:**

To achieve a 40% yield rate (enrollment rate), the estimated scholarship
amount needed is:

$$\hat{X}^* = \frac{\log(0.4/0.6) - (`r round(alpha_hat, 4)`)}{`r round(beta_hat, 6)`} = `r round(X_star, 2)` \text{ (thousand dollars)}$$

-   **Point estimate:**
    \$`r format(round(X_star * 1000, 0), big.mark = ",")`
-   **95% Confidence Interval:**
    (\$`r format(round(ci_lower * 1000, 0), big.mark = ",")`,
    \$`r format(round(ci_upper * 1000, 0), big.mark = ",")`)

The 95% confidence interval (\$`r format(round(ci_lower * 1000, 0), big.mark = ",")` to \$`r format(round(ci_upper * 1000, 0), big.mark = ",")`) is reasonably narrow and falls within the range of scholarship amounts in the data, indicating a reliable estimate.
