---
title: "P8131 Homework 1"
author: "Yongyan Liu (yl6107)"
date: "2026-02-09"
output:
  pdf_document:
    latex_engine: xelatex
    extra_dependencies: ["amsmath", "amssymb", "amsthm"]
header-includes:
  - \usepackage{bm}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Problem 1

**Show that the following distributions belong to the exponential family. Find the natural parameter $\theta$, scale parameter $\phi$ and convex function $b(\theta)$. Also find the $E(Y)$ and $\text{Var}(Y)$ as functions of the natural parameter. Specify the canonical link functions.**

The exponential family (with dispersion) has the form:
$$f(y|\theta, \phi) = \exp\left\{\frac{y\theta - b(\theta)}{\phi} + c(y, \phi)\right\}$$

Properties: $E(Y) = b'(\theta)$, $\text{Var}(Y) = \phi b''(\theta)$.

The canonical link function is $g = (b')^{-1}$, which satisfies $g(\mu) = \theta$.

## (a) Exponential Distribution $\text{Exp}(\lambda)$

$$f(y; \lambda) = \lambda e^{-\lambda y}, \quad y > 0$$

Rewrite:
$$f(y; \lambda) = \exp\{\log \lambda - \lambda y\} = \exp\{-\lambda y - (-\log \lambda)\}$$

Comparing with the exponential family form:

- **Natural parameter:** $\theta = -\lambda$ (so $\lambda = -\theta$)
- **Dispersion parameter:** $\phi = 1$
- **Convex function:** $b(\theta) = -\log(-\theta)$

**Verification:**

- $b'(\theta) = -\frac{1}{\theta} = \frac{1}{\lambda}$
- $b''(\theta) = \frac{1}{\theta^2} = \frac{1}{\lambda^2}$

**Mean and Variance:**

- $E(Y) = b'(\theta) = -\frac{1}{\theta} = \frac{1}{\lambda}$
- $\text{Var}(Y) = \phi b''(\theta) = \frac{1}{\theta^2} = \frac{1}{\lambda^2}$

**Canonical link function:** $g(\mu) = \theta = -\lambda = -\frac{1}{\mu}$ (negative reciprocal link)

---

## (b) Binomial Distribution $\text{Bin}(n, \pi)$, $n$ known

$$f(y; \pi) = \binom{n}{y} \pi^y (1-\pi)^{n-y}, \quad y = 0, 1, \ldots, n$$

Rewrite:
\begin{align*}
f(y; \pi) &= \binom{n}{y} \exp\left\{y \log \pi + (n-y) \log(1-\pi)\right\} \\
&= \binom{n}{y} \exp\left\{y \log \frac{\pi}{1-\pi} + n \log(1-\pi)\right\}
\end{align*}

Comparing with the exponential family form:

- **Natural parameter:** $\theta = \log \frac{\pi}{1-\pi}$ (logit of $\pi$), so $\pi = \frac{e^\theta}{1 + e^\theta}$
- **Dispersion parameter:** $\phi = 1$
- **Convex function:** $b(\theta) = n \log(1 + e^\theta)$

**Verification:**

- $b'(\theta) = \frac{n e^\theta}{1 + e^\theta} = n\pi$
- $b''(\theta) = \frac{n e^\theta}{(1 + e^\theta)^2} = n\pi(1-\pi)$

**Mean and Variance:**

- $E(Y) = b'(\theta) = n\pi = \frac{n e^\theta}{1 + e^\theta}$
- $\text{Var}(Y) = \phi b''(\theta) = n\pi(1-\pi) = \frac{n e^\theta}{(1 + e^\theta)^2}$

**Canonical link function:** $g(\mu) = \log \frac{\mu/n}{1 - \mu/n} = \log \frac{\mu}{n - \mu}$ (logit link)

*Note: If $Y^* = Y/n$ represents the proportion, then $g(\mu^*) = \log \frac{\mu^*}{1-\mu^*}$.*

---

## (c) Poisson Distribution $\text{Pois}(\lambda)$

$$f(y; \lambda) = \frac{1}{y!} \lambda^y e^{-\lambda}, \quad y = 0, 1, 2, \ldots$$

Rewrite:
$$f(y; \lambda) = \frac{1}{y!} \exp\{y \log \lambda - \lambda\}$$

Comparing with the exponential family form:

- **Natural parameter:** $\theta = \log \lambda$, so $\lambda = e^\theta$
- **Dispersion parameter:** $\phi = 1$
- **Convex function:** $b(\theta) = e^\theta$

**Verification:**

- $b'(\theta) = e^\theta = \lambda$
- $b''(\theta) = e^\theta = \lambda$

**Mean and Variance:**

- $E(Y) = b'(\theta) = e^\theta = \lambda$
- $\text{Var}(Y) = \phi b''(\theta) = e^\theta = \lambda$

**Canonical link function:** $g(\mu) = \log \mu$ (log link)

---

## (d) Negative Binomial Distribution $\text{NB}(m, \beta)$, $m$ known

$$f(y; \beta) = \binom{y + m - 1}{m - 1} \beta^m (1-\beta)^y, \quad y = 0, 1, 2, \ldots$$

Rewrite:
\begin{align*}
f(y; \beta) &= \binom{y + m - 1}{m - 1} \exp\{m \log \beta + y \log(1-\beta)\}
\end{align*}

Comparing with the exponential family form:

- **Natural parameter:** $\theta = \log(1 - \beta)$, so $1 - \beta = e^\theta$ and $\beta = 1 - e^\theta$
- **Dispersion parameter:** $\phi = 1$
- **Convex function:** $b(\theta) = -m \log(1 - e^\theta) = -m \log \beta$

**Verification:**

- $b'(\theta) = \frac{m e^\theta}{1 - e^\theta} = \frac{m(1-\beta)}{\beta}$
- $b''(\theta) = \frac{m e^\theta}{(1 - e^\theta)^2} = \frac{m(1-\beta)}{\beta^2}$

**Mean and Variance:**

- $E(Y) = b'(\theta) = \frac{m(1-\beta)}{\beta} = \frac{m e^\theta}{1 - e^\theta}$
- $\text{Var}(Y) = \phi b''(\theta) = \frac{m(1-\beta)}{\beta^2} = \frac{m e^\theta}{(1 - e^\theta)^2}$

**Canonical link function:** $g(\mu) = \log\left(\frac{\mu}{\mu + m}\right)$ (derived from $\theta = \log(1-\beta) = \log\frac{\mu}{\mu+m}$)

---

## (e) Gamma Distribution $\text{Gamma}(\alpha, \beta)$, $\alpha$ known

$$f(y; \beta) = \frac{\beta^\alpha}{\Gamma(\alpha)} y^{\alpha - 1} e^{-\beta y}, \quad y > 0$$

Rewrite:
$$f(y; \beta) = \exp\{-\beta y + \alpha \log \beta + (\alpha - 1) \log y - \log \Gamma(\alpha)\}$$

Comparing with the exponential family form:

- **Natural parameter:** $\theta = -\beta$ (so $\beta = -\theta$)
- **Dispersion parameter:** $\phi = 1/\alpha$
- **Convex function:** $b(\theta) = -\log(-\theta)$

To match the standard form with dispersion:
$$f(y|\theta, \phi) = \exp\left\{\frac{y\theta - b(\theta)}{\phi} + c(y, \phi)\right\}$$

We have $\phi = 1/\alpha$, $b(\theta) = -\log(-\theta)$, where $c(y, \phi) = (\frac{1}{\phi} - 1)\log y - \log\Gamma(1/\phi) - \frac{1}{\phi}\log\phi$.

**Verification:**

- $b'(\theta) = -\frac{1}{\theta} = \frac{1}{\beta}$
- $b''(\theta) = \frac{1}{\theta^2} = \frac{1}{\beta^2}$

**Mean and Variance:**

- $E(Y) = b'(\theta) = \frac{1}{\beta} = -\frac{1}{\theta} = \frac{\alpha}{\beta}$ (when $\phi = 1/\alpha$)
- $\text{Var}(Y) = \phi b''(\theta) = \frac{1}{\alpha} \cdot \frac{1}{\beta^2} = \frac{\alpha}{\beta^2}$

**Canonical link function:** $g(\mu) = -\frac{1}{\mu}$ (negative reciprocal link)

\newpage

# Problem 2

**Assume $Y_1, Y_2, \ldots, Y_n$ are independent and follow a binomial distribution where $Y_i \sim \text{Bin}(m, \pi_i)$ and $m$ is known. Furthermore, assume $\log \frac{\pi_i}{1-\pi_i} = X_i\beta$. What are the expressions of deviance residuals and Pearson residuals respectively (use $\hat{\beta}$ to represent the MLE)? What are the expressions of the deviance and Pearson's $\chi^2$ statistic?**

Let $\hat{\pi}_i = \frac{e^{X_i\hat{\beta}}}{1 + e^{X_i\hat{\beta}}}$ be the fitted probability, where $\hat{\beta}$ is the MLE.

## Pearson Residuals

The Pearson residual is defined as:
$$r_{P_i} = \frac{Y_i - \hat{\mu}_i}{\sqrt{V(\hat{\mu}_i)}}$$

For binomial data with $\hat{\mu}_i = m\hat{\pi}_i$ and $V(\hat{\mu}_i) = m\hat{\pi}_i(1 - \hat{\pi}_i)$:

$$r_{P_i} = \frac{Y_i - m\hat{\pi}_i}{\sqrt{m\hat{\pi}_i(1 - \hat{\pi}_i)}}$$

## Pearson's $\chi^2$ Statistic

$$G = \sum_{i=1}^{n} r_{P_i}^2 = \sum_{i=1}^{n} \frac{(Y_i - m\hat{\pi}_i)^2}{m\hat{\pi}_i(1 - \hat{\pi}_i)}$$

Under the assumed model, $G \stackrel{\text{approx}}{\sim} \chi^2(n - p)$ for large $m$.

## Deviance Residuals

The deviance residual is defined as:
$$r_{D_i} = \text{sign}(Y_i - \hat{\mu}_i) \sqrt{d_i}$$

where $d_i = D(Y_i, \hat{\mu}_i)$ is the individual contribution to the deviance.

For binomial data, the deviance contribution is:
$$d_i = 2\left[Y_i \log\frac{Y_i}{m\hat{\pi}_i} + (m - Y_i) \log\frac{m - Y_i}{m(1 - \hat{\pi}_i)}\right]$$

Therefore:
$$r_{D_i} = \text{sign}(Y_i - m\hat{\pi}_i) \sqrt{2\left[Y_i \log\frac{Y_i}{m\hat{\pi}_i} + (m - Y_i) \log\frac{m - Y_i}{m(1 - \hat{\pi}_i)}\right]}$$

*Note: The convention $0 \cdot \log 0 = 0$ is used when $Y_i = 0$ or $Y_i = m$.*

## Deviance

$$D = \sum_{i=1}^{n} d_i = 2\sum_{i=1}^{n}\left[Y_i \log\frac{Y_i}{m\hat{\pi}_i} + (m - Y_i) \log\frac{m - Y_i}{m(1 - \hat{\pi}_i)}\right]$$

Under the assumed model, $D \stackrel{\text{approx}}{\sim} \chi^2(n - p)$ for large $m$.

\newpage

# Problem 3

**Consider the binary response variable $Y \sim \text{Bernoulli}$ with $P(Y = 1) = \pi$ and $P(Y = 0) = 1 - \pi$. Observations $Y_i$, $i = 1, \ldots, n$, are independent and identically distributed as $Y$.**

## (a) Test Statistics for $H_0: \pi = \pi_0$

The log-likelihood function is:
$$\ell(\pi) = \sum_{i=1}^{n} \{Y_i \log \pi + (1 - Y_i) \log(1 - \pi)\} = n\bar{Y} \log \pi + n(1 - \bar{Y}) \log(1 - \pi)$$

where $\bar{Y} = \frac{1}{n}\sum_{i=1}^{n} Y_i$.

**Score function:**
$$s(\pi) = \frac{\partial \ell}{\partial \pi} = \frac{n\bar{Y}}{\pi} - \frac{n(1 - \bar{Y})}{1 - \pi} = \frac{n(\bar{Y} - \pi)}{\pi(1 - \pi)}$$

**Fisher information:**
$$\iota(\pi) = E\left[-\frac{\partial^2 \ell}{\partial \pi^2}\right] = \frac{n}{\pi(1 - \pi)}$$

**MLE:** $\hat{\pi} = \bar{Y}$

### Wald Test Statistic

$$TS_W = (\hat{\pi} - \pi_0)^T \iota(\hat{\pi}) (\hat{\pi} - \pi_0) = \frac{n(\bar{Y} - \pi_0)^2}{\bar{Y}(1 - \bar{Y})}$$

### Score Test Statistic

$$TS_S = s(\pi_0)^T \iota(\pi_0)^{-1} s(\pi_0) = \frac{n(\bar{Y} - \pi_0)^2}{\pi_0(1 - \pi_0)}$$

### Likelihood Ratio Test Statistic

$$TS_{LR} = 2\{\ell(\hat{\pi}) - \ell(\pi_0)\} = 2n\left[\bar{Y} \log\frac{\bar{Y}}{\pi_0} + (1 - \bar{Y}) \log\frac{1 - \bar{Y}}{1 - \pi_0}\right]$$

Under $H_0$, all three statistics asymptotically follow $\chi^2(1)$.

---

## (b) Numerical Calculations

Data: $(0, 1, 0, 0, 1, 0, 0, 0, 1, 0)$, so $n = 10$, $\sum Y_i = 3$, $\bar{Y} = 0.3$.

```{r}
# Data
n <- 10
y_bar <- 0.3

# Function to calculate test statistics
calc_tests <- function(pi0, y_bar, n) {
  # Wald test
  if (y_bar == 0 | y_bar == 1) {
    wald <- NA
  } else {
    wald <- n * (y_bar - pi0)^2 / (y_bar * (1 - y_bar))
  }

  # Score test
  score <- n * (y_bar - pi0)^2 / (pi0 * (1 - pi0))

  # LRT
  if (y_bar == 0) {
    lrt <- 2 * n * log((1 - y_bar) / (1 - pi0))
  } else if (y_bar == 1) {
    lrt <- 2 * n * log(y_bar / pi0)
  } else {
    lrt <- 2 * n * (y_bar * log(y_bar / pi0) + (1 - y_bar) * log((1 - y_bar) / (1 - pi0)))
  }

  # p-values
  p_wald <- 1 - pchisq(wald, 1)
  p_score <- 1 - pchisq(score, 1)
  p_lrt <- 1 - pchisq(lrt, 1)

  return(c(Wald = wald, Score = score, LRT = lrt,
           p_Wald = p_wald, p_Score = p_score, p_LRT = p_lrt))
}

# Critical value
alpha <- 0.05
crit_val <- qchisq(1 - alpha, 1)
cat("Chi-squared critical value (alpha = 0.05):", round(crit_val, 4), "\n\n")

# (i) pi0 = 0.1
cat("(i) H0: pi = 0.1\n")
res1 <- calc_tests(0.1, y_bar, n)
print(round(res1, 4))

# (ii) pi0 = 0.3
cat("\n(ii) H0: pi = 0.3\n")
res2 <- calc_tests(0.3, y_bar, n)
print(round(res2, 4))

# (iii) pi0 = 0.5
cat("\n(iii) H0: pi = 0.5\n")
res3 <- calc_tests(0.5, y_bar, n)
print(round(res3, 4))
```


---

## (c) Comparison of Tests

```{r}
results <- data.frame(
  pi0 = c(0.1, 0.3, 0.5),
  Wald = c(res1[1], res2[1], res3[1]),
  Score = c(res1[2], res2[2], res3[2]),
  LRT = c(res1[3], res2[3], res3[3]),
  Wald_decision = ifelse(c(res1[1], res2[1], res3[1]) > crit_val, "Reject", "FTR"),
  Score_decision = ifelse(c(res1[2], res2[2], res3[2]) > crit_val, "Reject", "FTR"),
  LRT_decision = ifelse(c(res1[3], res2[3], res3[3]) > crit_val, "Reject", "FTR")
)
knitr::kable(results, digits = 4, caption = "Test Statistics and Decisions")
```

**Conclusions:**

- For $\pi_0 = 0.1$: Tests disagree—Score rejects while Wald and LRT fail to reject.
- For $\pi_0 = 0.3$: All tests fail to reject (expected since $\hat{\pi} = \pi_0$).
- For $\pi_0 = 0.5$: All tests fail to reject.

With only $n = 10$, statistical power is low—even though $\hat{\pi} = 0.3$, we lack sufficient evidence to reject $\pi_0 = 0.1$ (for Wald and LRT) or $\pi_0 = 0.5$ (for all tests). Asymptotically (large $n$), all three tests are equivalent and would have greater power to detect such differences.

# Problem 4

**Assume $Y_i \sim \text{Pois}(\lambda)$, $i = 1, \ldots, n$. We are interested in testing $H_0: \log \lambda = \log \lambda_0$.**

Since log is a monotonic function, $H_0: \log \lambda = \log \lambda_0 \Leftrightarrow H_0: \lambda = \lambda_0$. Thus the test statistics are identical regardless of parameterization. These results are the same as Lecture 4, slide 10-11.

**Log-likelihood:**
$$\ell(\lambda) = \sum_{i=1}^{n} \{Y_i \log \lambda - \lambda - \log Y_i!\} = n\bar{Y} \log \lambda - n\lambda - \sum_{i=1}^{n} \log Y_i!$$

**Score function:**
$$s(\lambda) = \frac{\partial \ell}{\partial \lambda} = \frac{n\bar{Y}}{\lambda} - n = \frac{n(\bar{Y} - \lambda)}{\lambda}$$

**Fisher information:**
$$\iota(\lambda) = E\left[-\frac{\partial^2 \ell}{\partial \lambda^2}\right] = E\left[\frac{n\bar{Y}}{\lambda^2}\right] = \frac{n}{\lambda}$$

**MLE:** $\hat{\lambda} = \bar{Y}$

## Wald Test Statistic

$$TS_W = (\hat{\lambda} - \lambda_0)^T \iota(\hat{\lambda}) (\hat{\lambda} - \lambda_0) = \frac{n(\bar{Y} - \lambda_0)^2}{\bar{Y}}$$

## Score Test Statistic

$$TS_S = s(\lambda_0)^T \iota(\lambda_0)^{-1} s(\lambda_0) = \frac{[n(\bar{Y} - \lambda_0)/\lambda_0]^2}{n/\lambda_0} = \frac{n(\bar{Y} - \lambda_0)^2}{\lambda_0}$$

$$TS_S = \frac{n(\bar{Y} - \lambda_0)^2}{\lambda_0}$$

## Likelihood Ratio Test Statistic

$$TS_{LR} = 2\{\ell(\hat{\lambda}) - \ell(\lambda_0)\} = 2\{n\bar{Y} \log \bar{Y} - n\bar{Y} - n\bar{Y} \log \lambda_0 + n\lambda_0\}$$

$$TS_{LR} = 2n\left\{\bar{Y} \log\frac{\bar{Y}}{\lambda_0} - (\bar{Y} - \lambda_0)\right\}$$

Under $H_0: \lambda = \lambda_0$, all three test statistics asymptotically follow $\chi^2(1)$.

