---
title: "COVID-19 Severity: Predictive Modeling and Risk-Score Analysis"
author: "Yongyan Liu (yl6107) — P8106 Data Science II Final Project"
date: "2026-05-11"
output:
  pdf_document:
    latex_engine: xelatex
fontsize: 12pt
geometry: margin=1in
mainfont: "Times New Roman"
header-includes:
  - \usepackage{setspace}\onehalfspacing
  - \usepackage{caption}\captionsetup{font=small,labelfont=bf}
  - \usepackage{float}
  - \usepackage{booktabs}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.pos = "H", out.extra = "")
library(tidyverse); library(pROC); library(gridExtra)
library(knitr); library(kableExtra)

## results.RData is produced by p8106_final_yl6107_code.Rmd and contains
## every fitted model, the resamples object, test-set performance, plus
## all pre-computed render inputs (pdp_data, vi_tab, rocs, calib_df,
## auc_ci_df). This report only reads from it; it does not refit, predict,
## bootstrap, or recompute PDPs.
load("results.RData")

cv_means  <- summary(resamp)$statistics$ROC[, "Mean"]
get_auc   <- function(m) test_perf$AUC[test_perf$Model == m]
get_cv    <- function(m) cv_means[m]
```

# 1. Introduction

The goal is to identify demographic and clinical factors that predict
severe COVID-19 illness and to derive a patient-level risk score. From
the 10,000-participant study, 1,000 were drawn at random
(`set.seed(6107)`) and partitioned 80 / 20 into training (n = 800) and
test (n = 200) sets. The outcome is binary (1 = severe, 0 = not
severe). All 13 candidate predictors in the codebook were retained —
seven continuous (age, height, weight, BMI, SBP, LDL, depression) and
six categorical, including the multi-level factors `race` and
`smoking`.

# 2. Exploratory analysis

Severe outcomes occur in `r round(mean(train_d$severity == "Severe") * 100)`\%
of the training sample. There is no missingness and no obvious outliers.
Summary statistics by outcome are in **Table 1** and the marginal
patterns in **Figure 1**. Three observations drive the modelling choices
that follow. **First**, severe patients are older (median age
`r round(median(train_d$age[train_d$severity == "Severe"]))` vs
`r round(median(train_d$age[train_d$severity == "NotSev"]))`) and have
higher BMI and SBP. **Second**, the empirical severity rate is
`r round(100*mean(train_d$severity[train_d$vaccine=="No"]=="Severe"))`\%
in unvaccinated vs
`r round(100*mean(train_d$severity[train_d$vaccine=="Yes"]=="Severe"))`\%
in vaccinated patients — by far the largest single-variable gap (the
next, hypertension, is
`r round(100*mean(train_d$severity[train_d$hypertension=="Yes"]=="Severe"))`\%
vs
`r round(100*mean(train_d$severity[train_d$hypertension=="No"]=="Severe"))`\%).
**Third**, `height`, `weight` and `bmi` are mechanically related
(BMI = weight / height\textsuperscript{2}, with r = 0.70 between weight
and BMI and r = −0.50 between height and BMI); `SBP` and `hypertension`
correlate at r = 0.80. We kept all 13 codebook predictors and let
regularised / tree-based models handle the redundancy.

# 3. Model training

Nine candidates were trained with `caret::train` on the 800-row training
set using **shared 10-fold CV folds** (`trainControl(index =
createFolds(...))`) so that every model is compared on identical
resamples. Tuning and ranking metric: area under the ROC curve
(`metric = "ROC"`). `caret` dummy-coded factors automatically; the two
SVMs additionally centred and scaled the predictors. The seed (6107)
was set before the split, before fold construction, and before every
`train` call. **Every complexity hyperparameter** — `lambda`, `nprune`,
`min.node.size`, `n.trees`, `interaction.depth`, `shrinkage`, $C$,
$\sigma$ — **was selected strictly inside its grid**, so the
cross-model comparison is apple-to-apple. Three secondary picks land
at a family's natural endpoint (GLMNET $\alpha=1$, MARS `degree=1`,
RF `mtry=1`); these are not signs of an under-extended grid but
themselves substantive findings — the data prefer the simplest /
most-regularised variant within each family.

The candidate set spans interpretable and flexible families:
**logistic regression** (`glm`); **penalised logistic regression**
(`glmnet`, $\alpha\in[0,1]$ over 11 values, $\lambda$ over 30 log-spaced
values $e^{-8}$–$e^{-1}$); **GAM** with thin-plate smooths on the
continuous predictors; **MARS** (`earth`, `degree` 1–3, `nprune` 2–18);
**LDA**; **random forest** (`ranger`, `mtry`$\in\{1,\ldots,8\}$,
`min.node.size`$\in\{2,4,\ldots,16\}$); **AdaBoost** (`gbm`,
`distribution = "adaboost"`,
`n.trees`$\in\{1{,}000,2{,}000,3{,}000,4{,}000,5{,}000\}$,
`interaction.depth` 1–4,
`shrinkage`$\in\{0.0005, 0.001, 0.003, 0.01, 0.03\}$,
`n.minobsinnode = 10`); and two SVMs — **linear** (`svmLinear`, $C$ on
a 30-point log grid from $e^{-3}$ to $e^{8}$) and **radial-basis**
(`svmRadialSigma`, $C$ on $e^{-5}$–$e^{5}$ and $\sigma$ on $e^{-8}$–$e^{-2}$,
both with 10–15 grid points). CV tuning curves are in **Figure 2**;
selected hyperparameters are in **Table 2**.

# 4. Results

## 4.1 Predictive performance — and the boosting / SVM question

CV ROC and held-out test-set AUC, accuracy, sensitivity, specificity,
and Brier are in **Table 3** and **Figure 3**; CV ROC distributions are
in **Figure 2 (panel e)**. **The answer to the researchers' first
question is no — boosting and SVM do not beat the simpler methods.**
On 10-fold CV, MARS and GLMNET tie at the top (mean ROC 0.877);
AdaBoost is essentially tied
(`r sprintf("%.3f", get_cv("AdaBoost"))`), and
SVM linear (`r sprintf("%.3f", get_cv("SVM_lin"))`),
SVM radial, GLM, LDA, and GAM all sit within $\le 0.01$ of the top;
**random forest is the only model meaningfully weaker on CV**
(`r sprintf("%.3f", get_cv("RF"))`).
On the test set every AUC falls in the narrow band
[`r sprintf("%.3f", min(test_perf$AUC))`,
`r sprintf("%.3f", max(test_perf$AUC))`], a spread of only
`r sprintf("%.3f", max(test_perf$AUC) - min(test_perf$AUC))` — the 95\%
bootstrap CIs (Table 3) overlap completely and are each ~0.09 wide.
**The test-AUC ranking is therefore essentially within sampling noise
at $n_{\text{test}}=200$**: RF rises from worst on CV to nominal first
on test, while MARS slips from tied-first to last, with neither move
exceeding ~1\%. The CV ranking is the more reliable summary because
it averages over ten folds rather than a single 200-patient draw.

Why don't the flexible methods help? With the four dominant predictors
entering the response in nearly additive, threshold-linear form
(see §4.3), there is no nonlinear-interaction signal for AdaBoost or
the radial SVM to exploit; the extra flexibility chases noise. RF is
penalised more — `mtry`=1 with `min.node.size`=10 is the most
regularised configuration on its CV curve, and even that loses ~0.02
to the simple methods. Boosting's lower test sensitivity at the 0.5
cutoff (Sens = `r sprintf("%.2f", test_perf$Sens[test_perf$Model=="AdaBoost"])`)
reflects miscalibrated probabilities under the exponential loss, not a
shortfall in discrimination — its test AUC is in the same band as the
others.

## 4.2 Predictive risk score

MARS and GLMNET are tied at the top of CV ROC, but only MARS yields a
closed-form, additive risk-score expression that a clinician can read
off; we therefore use MARS to define $\hat{p}(\text{Severe}\mid x)$.
After CV the model reduces to **five terms in four predictors**:
\begin{equation*}
\text{logit}\bigl(\hat{p}_{\text{Severe}}\bigr) =
-0.96 + 0.171\,h(\text{age}-60) + 0.353\,h(\text{bmi}-28.7)
+ 0.089\,h(\text{SBP}-118) - 3.32\cdot\mathbf{1}[\text{vaccinated}],
\end{equation*}
where $h(u)=\max(u,0)$, so each continuous risk factor is flat below
its breakpoint and rises linearly above it. This is unusually
transparent for a "machine-learning" model. The score spans
$[0.01, 0.99]$ on the test set. At the default 0.5 cutoff, test
sensitivity is
`r sprintf("%.2f", test_perf$Sens[test_perf$Model=="MARS"])`
and specificity
`r sprintf("%.2f", test_perf$Spec[test_perf$Model=="MARS"])`
(test Brier `r sprintf("%.3f", test_perf$Brier[test_perf$Model=="MARS"])`);
clinicians can re-thresh along the ROC curve to trade the two off.
Calibration on the held-out test set (**Figure 4**) tracks the diagonal
in the low- and mid-risk range; the highest two deciles deviate by
about 0.10, but each contains only 20 patients, so the apparent
miscalibration is comparable to the per-bin Monte Carlo noise.

## 4.3 Key predictors and their effects

Three importance measures — MARS term selection, AdaBoost relative
influence (**Figure 5a**), and random-forest permutation importance —
agree on the top predictors and on their order: **vaccination status,
SBP, age, and BMI**, which together account for about 98% of
AdaBoost's relative influence. Hypertension is a near-duplicate of
SBP (r = 0.80) and falls out of MARS once SBP enters; race, gender,
smoking, LDL, and depression each contribute below 1% to AdaBoost and
are never selected by MARS.

The partial dependence plots (**Figure 5b**) show *how predicted
risk varies with* each predictor in the AdaBoost model (model-implied
associations in this observational sample, not causal effects):

- **Vaccination:** by far the strongest single signal. The PDP gap
  between unvaccinated and vaccinated is
  `r sprintf("%.2f", 0.561 - 0.126)`
  in $\hat{p}(\text{Severe})$ (0.56 vs 0.13), echoing the empirical-rate
  gap (62\% vs 10\%) and the $-3.32$ MARS vaccine coefficient.
- **Age:** flat below ~58, rising sharply between 58 and 65 to a
  plateau near 0.40 (consistent with the MARS hinge at 60).
- **BMI:** flat below ~28, then rising steeply between 28 and 33
  (PDP climbing from ~0.25 to ~0.40), aligning with the MARS hinge
  at 28.7.
- **SBP:** rising monotonically from ~120 mmHg, with the steepest
  segment between 125 and 140. This captures the same signal as
  binary `hypertension`, which is why hypertension drops out of MARS
  once SBP is in the model.

These directions are biologically plausible and consistent across MARS
hinge coefficients, AdaBoost PDPs, and the GAM smooths (not shown).
They should be read as predictive associations rather than causal
effects: vaccinated and unvaccinated patients differ systematically in
age and comorbidities, so unadjusted comparisons overstate any causal
effect of vaccination.

# 5. Conclusion

For this 1,000-patient sample, **boosting and SVM provide no advantage
over simpler models** — the answer to the researchers' first question.
On 10-fold CV, MARS and GLMNET tie at the top (mean ROC 0.877), with
AdaBoost, the two SVMs, GLM, LDA, and GAM all within $\le 0.01$ of
them; only random forest is meaningfully weaker. Test AUCs all sit in
[`r sprintf("%.3f", min(test_perf$AUC))`,
`r sprintf("%.3f", max(test_perf$AUC))`] with completely overlapping
95\% bootstrap CIs, so the test ranking is dominated by the noise of a
200-patient draw. We select **MARS** as the final model: tied for best
on CV and uniquely yielding a five-term, four-predictor closed-form
risk-score formula clinicians can compute at the bedside.
**The dominant predictors of severe COVID-19 are vaccination status
(strongly inversely associated), age, BMI, and systolic blood
pressure**, entering the response in a nearly additive, threshold-linear
form — which is precisely why the simple methods suffice. The risk
score is well-calibrated below the median and re-thresholdable along the
ROC curve for any sensitivity/specificity target. **Limitations.**
The estimates rest on a single 1,000-patient observational sample
without external validation, so the coefficients are predictive
associations rather than causal effects; calibration in the top two
deciles (Figure 4) is uncertain at $n_{\text{bin}}=20$ and would need
re-checking on a larger or higher-risk validation cohort.

\newpage

# Appendix — Tables and Figures

```{r tab1}
n_sev    <- sum(train_d$severity == "Severe")
n_notsev <- sum(train_d$severity == "NotSev")

# Continuous block: mean (SD) by severity
cont_block <- train_d |>
  pivot_longer(c(age, bmi, SBP, height, weight, LDL, depression),
               names_to = "Variable", values_to = "v") |>
  group_by(severity, Variable) |>
  summarise(val = sprintf("%.1f (%.1f)", mean(v), sd(v)),
            .groups = "drop") |>
  pivot_wider(names_from = severity, values_from = val) |>
  mutate(Variable = factor(Variable,
                            levels = c("age", "bmi", "SBP",
                                       "height", "weight", "LDL", "depression"))) |>
  arrange(Variable) |>
  mutate(Variable = as.character(Variable))

# Categorical block: n (column %) by severity
cat_levels <- list(
  vaccine      = c("Yes", "No"),
  gender       = c("Male", "Female"),
  race         = c("White", "Asian", "Black", "Hispanic"),
  smoking      = c("Never", "Former", "Current"),
  hypertension = c("Yes", "No"),
  diabetes     = c("Yes", "No")
)

cat_block <- map_dfr(names(cat_levels), function(v) {
  lv <- cat_levels[[v]]
  tibble(
    Variable = paste0("  ", lv),
    Severe   = vapply(lv, function(l) {
      n <- sum(train_d[[v]] == l & train_d$severity == "Severe")
      sprintf("%d (%.0f%%)", n, 100 * n / n_sev)
    }, character(1)),
    NotSev   = vapply(lv, function(l) {
      n <- sum(train_d[[v]] == l & train_d$severity == "NotSev")
      sprintf("%d (%.0f%%)", n, 100 * n / n_notsev)
    }, character(1)),
    .group   = v
  )
})

cont_section <- cont_block |>
  transmute(Variable = Variable, Severe = Severe, NotSev = NotSev, .group = "continuous")

t1 <- bind_rows(cont_section, dplyr::select(cat_block, Variable, Severe, NotSev, .group))

cont_n <- nrow(cont_section)
cat_starts <- cont_n + cumsum(c(0, head(rle(cat_block$.group)$lengths, -1))) + 1
cat_ends   <- cont_n + cumsum(rle(cat_block$.group)$lengths)
cat_labels <- rle(cat_block$.group)$values

tab1 <- kable(dplyr::select(t1, Variable, Severe, NotSev),
              booktabs = TRUE, align = "lrr", escape = FALSE,
              col.names = c("",
                            sprintf("Severe (n = %d)", n_sev),
                            sprintf("Not severe (n = %d)", n_notsev)),
              caption = "Patient characteristics by COVID-19 severity (training set, n = 800). Continuous variables: mean (SD); categorical variables: n (column \\%).") |>
  kable_styling(latex_options = c("hold_position"), font_size = 10) |>
  pack_rows("Continuous, mean (SD)", 1, cont_n,
            latex_gap_space = "0.3em") |>
  pack_rows("Categorical, n (column %)", cont_n + 1, nrow(t1),
            latex_gap_space = "0.4em")

for (i in seq_along(cat_labels)) {
  tab1 <- tab1 |>
    pack_rows(cat_labels[i], cat_starts[i], cat_ends[i],
              latex_gap_space = "0.15em",
              bold = FALSE, italic = TRUE,
              hline_before = FALSE, hline_after = FALSE)
}

tab1
```

```{r fig1, fig.cap="EDA — top: continuous predictors by severity (boxplots, training set, n = 800). Bottom: empirical severity rate by level for categorical predictors; bars above the overall rate of 32\\% (dashed line) are red, those below are blue.", fig.width=8.5, fig.height=6, out.width="92%", fig.align='center'}
num_vars <- c("age", "bmi", "SBP", "LDL", "depression")
p_num <- train_d |>
  pivot_longer(all_of(num_vars), names_to = "Var", values_to = "v") |>
  ggplot(aes(severity, v, fill = severity)) +
  geom_boxplot(alpha = 0.6, outlier.size = 0.5) +
  facet_wrap(~ Var, scales = "free_y", nrow = 1) +
  scale_fill_manual(values = c("Severe" = "firebrick", "NotSev" = "steelblue")) +
  labs(x = NULL, y = NULL) +
  theme_bw(base_size = 10) +
  theme(legend.position = "none",
        strip.background = element_rect(fill = "grey95"))

cat_vars <- c("vaccine", "hypertension", "diabetes", "gender", "race", "smoking")
overall_rate <- mean(train_d$severity == "Severe")
p_cat <- train_d |>
  pivot_longer(all_of(cat_vars), names_to = "Var", values_to = "lvl") |>
  group_by(Var, lvl) |>
  summarise(rate = mean(severity == "Severe"), n = n(), .groups = "drop") |>
  mutate(Var = factor(Var, levels = cat_vars),
         above = rate > overall_rate) |>
  ggplot(aes(lvl, rate, fill = above)) +
  geom_col(alpha = 0.85) +
  geom_hline(yintercept = overall_rate,
             linetype = "dashed", colour = "grey30") +
  scale_fill_manual(values = c(`TRUE` = "firebrick", `FALSE` = "steelblue")) +
  facet_wrap(~ Var, scales = "free_x", nrow = 2) +
  labs(x = NULL, y = "P(Severe)") +
  theme_bw(base_size = 10) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 35, hjust = 1),
        strip.background = element_rect(fill = "grey95"))

grid.arrange(p_num, p_cat, nrow = 2, heights = c(1, 1.6))
```

```{r fig2, fig.cap="CV tuning curves and the resamples ROC distribution. (a) GLMNET ROC vs $\\log\\lambda$, coloured by mixing $\\alpha$. (b) MARS ROC vs nprune, coloured by interaction degree. (c) AdaBoost ROC vs n.trees, coloured by interaction depth, faceted by shrinkage. (d) Radial SVM ROC heatmap on $(\\log\\sigma,\\log C)$; the cross marks the CV-selected pair. (e) 10-fold CV ROC distribution by model.", fig.width=8.5, fig.height=10, out.width="92%", fig.align='center'}
p_glmn <- ggplot(fits$GLMNET, xTrans = log, highlight = TRUE) +
  ggtitle("(a) GLMNET") +
  theme_bw(base_size = 9) +
  theme(legend.position = "right", legend.key.size = unit(0.35, "cm"))
p_mars <- ggplot(fits$MARS, highlight = TRUE) +
  ggtitle("(b) MARS") +
  theme_bw(base_size = 9) +
  theme(legend.position = "right", legend.key.size = unit(0.35, "cm"))
p_gbm  <- ggplot(fits$AdaBoost, highlight = TRUE) +
  ggtitle("(c) AdaBoost") +
  theme_bw(base_size = 9) +
  theme(legend.position = "right", legend.key.size = unit(0.35, "cm"))

svr_perf <- fits$SVM_rbf$results
svr_best <- fits$SVM_rbf$bestTune
roc_lim  <- range(svr_perf$ROC)
p_svr <- ggplot(svr_perf, aes(log(sigma), log(C), fill = ROC)) +
  geom_tile() +
  scale_fill_viridis_c(option = "C", direction = 1,
                       limits = c(max(roc_lim[1], 0.78), roc_lim[2]),
                       oob = scales::squish, name = "CV ROC") +
  annotate("point", x = log(svr_best$sigma), y = log(svr_best$C),
           shape = 4, size = 4, stroke = 1.5, colour = "white") +
  ggtitle("(d) Radial SVM") +
  theme_bw(base_size = 9) +
  theme(legend.key.size = unit(0.4, "cm"))

roc_long <- resamp$values |>
  dplyr::select(Resample, ends_with("~ROC")) |>
  pivot_longer(-Resample, names_to = "Model", values_to = "ROC") |>
  mutate(Model = sub("~ROC", "", Model),
         Model = factor(Model,
                        levels = names(sort(cv_means))))
p_box <- ggplot(roc_long, aes(Model, ROC)) +
  geom_boxplot(fill = "grey80") +
  geom_jitter(width = 0.1, alpha = 0.5, size = 1, colour = "firebrick") +
  coord_flip() +
  ggtitle("(e) 10-fold CV ROC distribution") +
  theme_bw(base_size = 9)

grid.arrange(grobs = list(p_glmn, p_mars, p_gbm, p_svr, p_box),
             layout_matrix = rbind(c(1, 2), c(3, 4), c(5, 5)),
             heights = c(1, 1, 0.9))
```

\clearpage

```{r tab-best}
bt <- function(m) paste(sapply(names(m), function(nm)
        sprintf("%s=%s", nm, format(unlist(m[[nm]]), digits = 3))),
        collapse = ", ")
tune_tab <- tibble(
  Model = c("GLM", "GLMNET", "GAM", "MARS", "LDA",
            "RF (ranger)", "AdaBoost (gbm)", "SVM linear", "SVM radial"),
  `Tuning grid (10-fold CV)` = c(
    "—",
    "alpha: 11 pts in [0,1]; lambda: 30 log-pts in exp(-8..-1)",
    "default thin-plate smooths on continuous predictors",
    "degree: 1--3; nprune: 2--18",
    "—",
    "mtry: 1--8; min.node.size: {2,4,...,16}",
    "n.trees: {1000,...,5000}; depth: 1--4; shrinkage: {5e-4, 1e-3, 3e-3, 0.01, 0.03}",
    "C: 30 log-pts in exp(-3..8)",
    "C: 15 log-pts in exp(-5..5); sigma: 10 log-pts in exp(-8..-2)"),
  `Best params` = c(
    "—",
    bt(fits$GLMNET$bestTune),
    "—",
    bt(fits$MARS$bestTune),
    "—",
    bt(fits$RF$bestTune),
    bt(fits$AdaBoost$bestTune),
    bt(fits$SVM_lin$bestTune),
    bt(fits$SVM_rbf$bestTune)
  )
)
kable(tune_tab, booktabs = TRUE, align = "lll",
      caption = "Hyper-parameter tuning grids and CV-selected values. Continuous predictors are centred/scaled only for the two SVMs.") |>
  kable_styling(latex_options = c("hold_position"),
                font_size = 8.5) |>
  column_spec(2, width = "8.0cm") |>
  column_spec(3, width = "5.0cm")
```

```{r tab-perf}
test_perf_pretty <- test_perf %>%
  mutate(`CV ROC` = cv_means[Model]) %>%
  left_join(auc_ci_df, by = "Model") %>%
  arrange(desc(AUC)) %>%
  dplyr::select(Model, `CV ROC`, AUC_CI, Accuracy, Sens, Spec, Brier) %>%
  mutate(`CV ROC`  = sprintf("%.3f", `CV ROC`),
         Accuracy = sprintf("%.3f", Accuracy),
         Sens     = sprintf("%.3f", Sens),
         Spec     = sprintf("%.3f", Spec),
         Brier    = sprintf("%.3f", Brier))

kable(test_perf_pretty, booktabs = TRUE, align = "lrlrrrr",
      col.names = c("Model", "CV ROC (mean)", "Test AUC (95% CI)",
                    "Test Acc.", "Sens.", "Spec.", "Brier"),
      caption = "Cross-validated and test-set performance for all models. Models sorted by test AUC. 95\\% CIs are 2{,}000-replicate stratified bootstrap percentiles (\\texttt{pROC::ci.auc}). Brier = mean squared error of predicted probabilities; lower is better.") |>
  kable_styling(latex_options = c("hold_position"), font_size = 10)
```

```{r fig3, fig.cap="Test-set ROC curves for the nine fitted models with their AUCs.", fig.width=6.5, fig.height=5.5}
ord <- order(sapply(rocs, function(r) -as.numeric(r$auc)))
rocs_ord <- rocs[ord]
cols <- RColorBrewer::brewer.pal(9, "Set1")
plot(rocs_ord[[1]], col = cols[1], legacy.axes = TRUE, lwd = 2,
     main = "Test-set ROC curves")
for (i in seq_along(rocs_ord)[-1])
  plot(rocs_ord[[i]], add = TRUE, col = cols[i], lwd = 2)
legend("bottomright",
       legend = sprintf("%s (AUC=%.3f)", names(rocs_ord),
                        sapply(rocs_ord, function(r) as.numeric(r$auc))),
       col = cols, lwd = 2, cex = 0.7, bty = "n")
```

```{r fig4, fig.cap="Calibration of the MARS risk score on the held-out test set. Patients are split into 10 equal-size bins by predicted P(Severe); each dot is a bin's mean predicted vs. observed severity rate (point area is bin size). Closer to the dashed diagonal is better.", fig.width=5, fig.height=4}
ggplot(calib_df, aes(p_mean, obs)) +
  geom_abline(linetype = "dashed", colour = "grey50") +
  geom_point(aes(size = n), colour = "firebrick") +
  geom_line(colour = "firebrick", alpha = 0.5) +
  scale_size_continuous(range = c(2, 5), guide = "none") +
  labs(x = "Mean predicted P(Severe)", y = "Observed P(Severe)",
       title = sprintf("Calibration on test set - %s", best_name)) +
  theme_bw()
```

```{r fig5, fig.cap="Boosting interpretation. (a) Variable importance (relative influence, \\%) from the AdaBoost final model — vaccine, SBP, age, and BMI together account for $\\sim$98\\% of the predictive signal. (b) Partial dependence of the predicted P(Severe) on the four dominant predictors (vaccine, age, BMI, SBP), holding all other predictors at their training distribution.", fig.width=8.5, fig.height=8, out.width="92%", fig.align='center'}
vi_df <- as_tibble(vi_tab) |>
  arrange(rel.inf) |>
  mutate(var = factor(var, levels = var))

p_vi <- ggplot(vi_df, aes(rel.inf, var)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = sprintf("%.1f", rel.inf)), hjust = -0.1, size = 3) +
  scale_x_continuous(expand = expansion(mult = c(0, .15))) +
  labs(title = "(a) AdaBoost variable importance",
       x = "Relative influence (%)", y = NULL) +
  theme_bw(base_size = 11)

cont_pdp <- function(v) {
  ggplot(pdp_data[[v]], aes(.data[[v]], yhat)) +
    geom_line() +
    geom_rug(data = train_d, aes(.data[[v]]), sides = "b",
             alpha = 0.3, inherit.aes = FALSE) +
    labs(title = paste0("PDP - ", v), x = v, y = "P(Severe)") +
    theme_bw(base_size = 11)
}
p_age <- cont_pdp("age")
p_bmi <- cont_pdp("bmi")
p_sbp <- cont_pdp("SBP")
p_vac <- ggplot(pdp_data$vaccine,
                aes(vaccine, yhat, fill = vaccine)) +
  geom_col(width = 0.6, alpha = 0.85) +
  geom_text(aes(label = sprintf("%.2f", yhat)),
            vjust = -0.4, size = 3.2) +
  scale_fill_manual(values = c("No" = "firebrick", "Yes" = "steelblue")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title = "PDP - vaccine", x = "vaccine", y = "P(Severe)") +
  theme_bw(base_size = 11) +
  theme(legend.position = "none")

p_pdp_grid <- arrangeGrob(p_vac, p_age, p_bmi, p_sbp, nrow = 2)
grid.arrange(p_vi, p_pdp_grid, nrow = 2, heights = c(1, 1.4))
```
