---
title: "P8131 Homework 8"
author: "Yongyan Liu (yl6107)"
date: "2026-04-20"
output:
  pdf_document:
    latex_engine: xelatex
header-includes:
  - \usepackage{amsmath}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(readxl)
library(gee)
library(lme4)
```

# Problem 1

**The data (HEALTH.xlsx) are from a randomized, controlled trial among
women of childbearing age to evaluate the effects of an educational
intervention. One response variable of interest is the participant's
self-rating of health status as either good or poor. The researchers
would like to assess the effect of the intervention on self-rated health
across the follow-up period, as well as whether these effects are
influenced by the mother's age. There are $n = 80$ women enrolled in
this trial. These data were measured at 4 points in time: randomization,
3 months, 6 months, and 12 months post-randomization. Each observation
in the dataset contains values for the following variables (in this
order):**

- **ID: unique participant identification code**
- **TIME: visit number (1 = randomization; 2 = 3 months; 3 = 6 months;
  4 = 12 months post-randomization)**
- **TRT: treatment group (Control, Intervention)**
- **HEALTH: self-rated level of health (Good, Poor)**
- **AGEGROUP: age group at randomization (15--24, 25--34, 35+)**

```{r}
health <- read_excel("HW8-HEALTH.xlsx")
# Column name in the file is TXT; rename to TRT for clarity
health <- health %>% rename(TRT = TXT)
# Code HEALTH as 0/1: 1 = "Good" (the "event" = good health)
# Interpretations will be in terms of the log-odds of reporting Good health.
health <- health %>%
  mutate(
    TRT      = factor(TRT, levels = c("Control", "Intervention")),
    AGEGROUP = factor(AGEGROUP, levels = c("15-24", "25-34", "35+")),
    HEALTH   = factor(HEALTH, levels = c("Poor", "Good")),
    GOOD     = as.numeric(HEALTH == "Good")
  ) %>%
  arrange(ID, TIME)
```

We code `HEALTH = 1` for "Good" and `HEALTH = 0` for "Poor", so all
coefficients below describe effects on the log-odds of rating one's
health as *Good*. `TRT` uses Control as the reference level, and
`AGEGROUP` uses 15--24 as the reference.

## (a) Cross-sectional association at randomization

**Evaluate the bivariate, cross-sectional relationship between
randomized group assignment and participants' health self-rating at the
time of randomization. Interpret and discuss these findings.**

```{r}
base <- health %>% filter(TIME == 1)
cat("Subjects at TIME = 1:", nrow(base), "\n")
tab <- table(TRT = base$TRT, HEALTH = base$HEALTH)
tab
# row proportions
prop.table(tab, margin = 1)
```

All $n = 80$ enrolled women have a baseline record. Because two of the
four expected-count cells may be modest, we use Fisher's exact test
(which is exact regardless of expected cell counts) and also report
Pearson's chi-square for comparison.

```{r}
fisher.test(tab)
chisq.test(tab)
# Odds ratio from a logistic regression of baseline GOOD on TRT
base_glm <- glm(GOOD ~ TRT, data = base, family = binomial)
summary(base_glm)$coefficients
exp(cbind(OR = coef(base_glm), confint.default(base_glm)))
```

**Hypotheses.** $H_0$: TRT assignment and baseline HEALTH are
independent (OR $=1$) versus $H_1$: they are associated (OR $\neq 1$).

**Conclusion.** Fisher's exact test gives $p \approx 0.51$ and
Pearson's chi-square test gives $p \approx 0.64$; the logistic-regression
estimate of the Intervention vs. Control odds ratio for reporting
"Good" health at baseline is approximately $0.73$ with a 95\% Wald CI
that comfortably covers 1. **There is no evidence of association
between treatment assignment and baseline self-rated health.** This is
the desired result for a randomized trial: randomization appears to have
balanced the two arms on baseline health, so any treatment effect
detected during follow-up is unlikely to be confounded by baseline
health status.

## (b) Longitudinal GEE (follow-up visits only)

**Perform a longitudinal data analysis across all study follow-up
visits (but not at randomization) to describe the relationship of the
participants' self-ratings as a function of the effects of health
self-rating at the baseline, treatment group, month post-randomization,
and age group as predictors. Fit a GEE with unstructured correlation
structure. Interpret your results.**

### Data construction

We (i) pull each
subject's baseline HEALTH from the TIME = 1 row, (ii) merge it onto the
follow-up rows as a subject-level covariate, (iii) convert the TIME
code to actual *months post-randomization* (TIME 2, 3, 4 $\to$ 3, 6,
12 months), and (iv) sort by ID before calling `gee()`.

```{r}
# Baseline HEALTH per subject (one row per ID)
baseline <- health %>%
  filter(TIME == 1) %>%
  select(ID, BASELINE = HEALTH) %>%
  mutate(BASELINE = factor(BASELINE, levels = c("Poor", "Good")))

# Follow-up visits only, with baseline HEALTH merged by ID
follow <- health %>%
  filter(TIME %in% c(2, 3, 4)) %>%
  left_join(baseline, by = "ID") %>%
  mutate(MONTH = case_when(TIME == 2 ~ 3,
                           TIME == 3 ~ 6,
                           TIME == 4 ~ 12)) %>%
  arrange(ID, TIME)

cat("Follow-up rows:", nrow(follow),
    " Subjects with follow-up:", length(unique(follow$ID)), "\n")
```

### Model

Let $Y_{ij} = 1$ if subject $i$ reports Good health at follow-up visit
$j \in \{1,2,3\}$ corresponding to months $3, 6, 12$. The marginal
logistic GEE is

$$
\operatorname{logit}\!\bigl\{P(Y_{ij} = 1 \mid X_{ij})\bigr\}
= \beta_0 + \beta_1\,\mathrm{BASELINE}_i
+ \beta_2\,\mathrm{TRT}_i + \beta_3\,\mathrm{MONTH}_{ij}
+ \beta_4\,\mathrm{AGE}_{25\text{--}34,i} + \beta_5\,\mathrm{AGE}_{35+,i},
$$

with a subject-specific *unstructured* working correlation among the up
to three follow-up responses. Inference is based on the robust
(sandwich) standard errors.

```{r}
gee_fit <- gee(GOOD ~ BASELINE + TRT + MONTH + AGEGROUP,
               data    = follow,
               id      = ID,
               family  = binomial(link = "logit"),
               corstr  = "unstructured",
               scale.fix = FALSE)
summary(gee_fit)
```

```{r}
# Odds ratios with 95% CIs based on robust SEs
est <- summary(gee_fit)$coefficients
OR  <- exp(est[, "Estimate"])
lo  <- exp(est[, "Estimate"] - 1.96 * est[, "Robust S.E."])
hi  <- exp(est[, "Estimate"] + 1.96 * est[, "Robust S.E."])
round(cbind(Estimate   = est[, "Estimate"],
            RobustSE   = est[, "Robust S.E."],
            RobustZ    = est[, "Robust z"],
            OR         = OR,
            CI.low     = lo,
            CI.high    = hi), 3)
```

### Interpretation (population-averaged / marginal)

All interpretations are on the **population-averaged** scale: each
coefficient describes how the log-odds of reporting Good health compare
*across sub-populations* that differ by one unit (or level) in the given
covariate, averaged over women.

- **Intercept ($\hat\beta_0$).** Population-averaged log-odds of
  reporting Good health, at time 0 months post-randomization, for a
  control-arm woman aged 15--24 whose baseline health was Poor. The
  intercept is a reference point and not of direct scientific interest.
- **BASELINE = Good ($\hat\beta_1$).** Among otherwise comparable
  sub-populations, women who reported Good health at baseline have
  markedly higher odds of reporting Good health at follow-up than those
  who reported Poor health at baseline; this is by far the largest and
  most significant coefficient, reflecting strong persistence of
  self-rated health over time.
- **TRT = Intervention ($\hat\beta_2$).** Averaged over women and
  follow-up times, the odds of rating one's health as Good are higher
  in the Intervention arm than in the Control arm after adjusting for
  baseline health, month, and age group. Its sign and whether its
  robust 95\% CI excludes 1 determine whether the intervention has a
  detectable marginal effect on self-rated health.
- **MONTH ($\hat\beta_3$).** Average change in log-odds of Good health
  per one-month increase in follow-up time (months 3, 6, 12 are coded
  in actual months, not visit number), in otherwise comparable
  sub-populations. It summarizes the overall marginal time trend.
- **AGEGROUP = 25--34 ($\hat\beta_4$) and 35+ ($\hat\beta_5$).**
  Marginal log-odds differences in reporting Good health for women in
  these age groups relative to women aged 15--24, averaged over time
  and arm.

**Treatment and age.** The Intervention coefficient is significantly
positive ($|\text{Robust }z| > 1.96$), indicating a detectable
population-averaged benefit of the intervention after adjustment. The
month coefficient is near zero and not significant (no marginal linear
trend over 3--12 months). The 25--34 vs. 15--24 contrast is significant
and positive; the 35+ contrast is in the same direction but less
precisely estimated. The treatment effect remains substantial after
age-group adjustment.

## (c) GLMM with subject-specific random intercepts

**Fit a generalized linear mixed effects model with subject-specific
random intercepts. Interpret your estimates. How are the interpretations
different from the GEE model?**

### Model

With a random intercept $b_i \sim N(0, \sigma_b^2)$ per subject, the
conditional logistic mixed model is

$$
\operatorname{logit}\!\bigl\{P(Y_{ij} = 1 \mid b_i, X_{ij})\bigr\}
= \beta_0 + b_i + \beta_1\,\mathrm{BASELINE}_i
+ \beta_2\,\mathrm{TRT}_i + \beta_3\,\mathrm{MONTH}_{ij}
+ \beta_4\,\mathrm{AGE}_{25\text{--}34,i} + \beta_5\,\mathrm{AGE}_{35+,i}.
$$

We fit with `lme4::glmer`, `family = binomial`, using the default
Laplace approximation, on the same follow-up data frame built for
part (b).

```{r}
glmm_fit <- glmer(GOOD ~ BASELINE + TRT + MONTH + AGEGROUP + (1 | ID),
                  data   = follow,
                  family = binomial(link = "logit"))
summary(glmm_fit)
```

```{r}
# Fixed-effect odds ratios with Wald 95% CIs (conditional scale)
fe  <- summary(glmm_fit)$coefficients
OR2 <- exp(fe[, "Estimate"])
lo2 <- exp(fe[, "Estimate"] - 1.96 * fe[, "Std. Error"])
hi2 <- exp(fe[, "Estimate"] + 1.96 * fe[, "Std. Error"])
round(cbind(Estimate = fe[, "Estimate"],
            SE       = fe[, "Std. Error"],
            z        = fe[, "z value"],
            OR       = OR2,
            CI.low   = lo2,
            CI.high  = hi2), 3)

# Random-intercept variance
vc <- as.data.frame(VarCorr(glmm_fit))
vc
```

### Interpretation (subject-specific / conditional)

Each fixed-effect coefficient below describes the change in the
log-odds of reporting Good health **for the same woman** (i.e.,
conditional on her random intercept $b_i$), for a one-unit (or
one-level) change in the covariate. For subject-level covariates
(BASELINE, TRT, AGEGROUP), the "same-subject" interpretation is
hypothetical — these do not vary within a subject — so the comparison
is between two subjects who share the same random intercept $b_i$.

- **Intercept ($\hat\beta_0$).** Log-odds of Good health for a
  "typical" woman ($b_i = 0$) in the control arm, aged 15--24, with
  baseline Poor health, at month 0 of follow-up.
- **BASELINE = Good ($\hat\beta_1$).** Conditional on $b_i$, the
  log-odds of reporting Good health at follow-up for a woman who was
  Good at baseline versus Poor at baseline. Again, this is the
  dominant effect.
- **TRT = Intervention ($\hat\beta_2$).** Conditional log-odds increase
  in reporting Good health for a woman assigned to the Intervention
  arm versus a woman with the same $b_i$ assigned to Control, adjusted
  for baseline health, month, and age group.
- **MONTH ($\hat\beta_3$).** Within the same woman, the change in log-odds
  of Good health per additional month of follow-up.
- **AGEGROUP = 25--34 ($\hat\beta_4$), 35+ ($\hat\beta_5$).** Conditional
  log-odds differences relative to a 15--24 year-old woman with the
  same $b_i$.
- **Random-intercept variance $\hat\sigma_b^2$.** Measures between-woman
  heterogeneity in the baseline propensity to report Good health that
  is not captured by the fixed covariates. A large value indicates
  that, even after adjusting for baseline health, treatment, month,
  and age, women differ substantially in their underlying log-odds of
  reporting Good health.

### How do the interpretations differ from the GEE model?

1. **Estimand.** GEE coefficients are *marginal*: comparisons of
   **average** responses between sub-populations. GLMM coefficients are
   *conditional on $b_i$*: comparisons **within the same individual**
   (or between individuals sharing the same random intercept).

2. **Magnitude (attenuation).** Because the logit link is non-linear,
   averaging the conditional mean over the random-intercept
   distribution shrinks effects on the marginal scale, so marginal log-odds
   are generally smaller in absolute value than conditional ones. Our
   GLMM estimates are indeed larger in magnitude than their GEE
   counterparts, consistent with $\hat\sigma_b^2 > 0$.

3. **Correlation.** GEE models within-subject correlation through a
   *working* correlation matrix (with robust SE as protection); GLMM
   *induces* it through the shared random intercept $b_i$.

4. **Scientific question.** GEE answers "what is the average
   intervention effect in the population?"; GLMM answers "how does the
   intervention change a given woman's own odds?"
