---
title: "P8131 Homework 5"
author: "Yongyan Liu (yl6107)"
date: "2026-03-24"
output:
  pdf_document:
    latex_engine: xelatex
header-includes:
  - \usepackage{amsmath}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(pscl)
```

# Problem 1: Horseshoe Crabs

In a study of nesting horseshoe crabs, each female horseshoe crab had a
male crab attached to her in her nest. The study investigated factors
that affect whether the female crab had any other males, called
satellites, residing near her. Explanatory variables that are thought to
affect this included the female crab's color (C), spine condition (S),
carapace width (W) and weight (Wt). The response outcome for each
female crab is her number of satellites (Sa). There are 173 females in
this study. Data are provided in the crab.txt.

```{r}
crab <- read.table("HW5-crab.txt", header = TRUE)
```

## (a) Poisson Model M1 with W

**Fit a Poisson model (M1) with log link with W as the single
predictor. Check the goodness of fit and interpret your model.**

Model: $\log(\mu_i) = \beta_0 + \beta_1 W_i$, where
$Y_i \sim \text{Poisson}(\mu_i)$.

```{r}
M1 <- glm(Sa ~ W, family = poisson(link = "log"), data = crab)
summary(M1)
```

**Goodness of fit:**

```{r}
dev_M1 <- deviance(M1)
df_M1 <- M1$df.residual
p_dev_M1 <- 1 - pchisq(dev_M1, df_M1)

G_M1 <- sum(residuals(M1, type = "pearson")^2)
p_G_M1 <- 1 - pchisq(G_M1, df_M1)

cat("Deviance =", round(dev_M1, 2), ", df =", df_M1,
    ", p =", round(p_dev_M1, 4), "\n")
cat("Pearson X² =", round(G_M1, 2), ", df =", df_M1,
    ", p =", round(p_G_M1, 4), "\n")
```

Both the deviance and Pearson $\chi^2$ statistics are large relative to
their degrees of freedom (p < 0.05), indicating that the model does not
fit the data well. There is evidence of lack of fit, possibly due to
over-dispersion or missing predictors.

**Interpretation:** The estimated model is
$\log(\hat{\mu}) = `r round(coef(M1)[1], 4)` + `r round(coef(M1)[2], 4)` \cdot W$.

$\exp(\hat{\beta}_1)$ is the estimated rate ratio for the number of
satellites associated with a one-cm increase in carapace width. With
$\hat{\beta}_1 = `r round(coef(M1)[2], 4)`$ (p < 0.001), the expected
number of satellites for a female crab is
$`r round(exp(coef(M1)[2]), 3)`$ $(= \exp(`r round(coef(M1)[2], 4)`))$
times of that for a crab one cm narrower in width.

## (b) Model M2 with W and Wt

**Fit a model (M2) with W and Wt as predictors. Compare it with the
model in (a). Interpret your results.**

```{r}
M2 <- glm(Sa ~ W + Wt, family = poisson(link = "log"), data = crab)
summary(M2)
```

**Model comparison via deviance analysis (LRT):**

```{r}
test_stat <- deviance(M1) - deviance(M2)
df_test <- df_M1 - M2$df.residual
p_test <- 1 - pchisq(test_stat, df_test)

cat("Delta Deviance =", round(test_stat, 3), ", df =", df_test,
    ", p =", round(p_test, 4), "\n")
```

The test statistic is `r round(test_stat, 3)` on `r df_test` df with
p = `r round(p_test, 4)` < 0.05, so adding Wt leads to a statistically
significant improvement. However, the deviance only decreases from
`r round(dev_M1, 2)` to `r round(deviance(M2), 2)`, both far exceeding
their respective degrees of freedom, indicating that neither M1 nor M2
fits the data adequately (likely due to over-dispersion).

**Interpretation:** In M2, $W$ is no longer significant (p =
`r round(summary(M2)$coefficients["W", "Pr(>|z|)"], 3)`) while $Wt$ is
significant (p =
`r round(summary(M2)$coefficients["Wt", "Pr(>|z|)"], 3)`). This
suggests that weight captures most of the information about satellite
count that width provided, and the two predictors are likely correlated.

$\exp(\hat{\beta}_{Wt})$ is the estimated rate ratio for the number of
satellites associated with a one-kg increase in weight, holding width
fixed. With $\hat{\beta}_{Wt} = `r round(coef(M2)["Wt"], 4)`$, the
expected number of satellites is
$`r round(exp(coef(M2)["Wt"]), 3)`$
$(= \exp(`r round(coef(M2)["Wt"], 4)`))$ times of that for a crab one
kg lighter, holding width constant.

$\exp(\hat{\beta}_{W})$ is the estimated rate ratio for a one-cm
increase in width, holding weight fixed. With
$\hat{\beta}_{W} = `r round(coef(M2)["W"], 4)`$, $W$ is not
significant (p = `r round(summary(M2)$coefficients["W", "Pr(>|z|)"], 3)`),
suggesting width does not provide additional information beyond weight.

## (c) Over-dispersion in M2

**Check over dispersion in M2. Interpret the model after adjusting for
over dispersion.**

```{r}
G_M2 <- sum(residuals(M2, type = "pearson")^2)
phi_hat <- G_M2 / M2$df.residual

cat("Pearson X² =", round(G_M2, 2), "\n")
cat("Estimated dispersion parameter: phi =", round(phi_hat, 4), "\n")
```

Since $\hat{\phi} = `r round(phi_hat, 2)`$ is substantially greater
than 1, there is evidence of over-dispersion.

**Adjusted model using quasi-Poisson:**

```{r}
M2_quasi <- glm(Sa ~ W + Wt, family = quasipoisson(link = "log"),
                data = crab)
summary(M2_quasi)
```

The updated model has the same coefficient estimates as M2. After
adjusting for over-dispersion, the standard errors are inflated by a
factor of $\sqrt{\hat{\phi}} = `r round(sqrt(phi_hat), 3)`$. In the
adjusted model, neither $W$ (p =
`r round(summary(M2_quasi)$coefficients["W", "Pr(>|t|)"], 3)`) nor $Wt$
(p =
`r round(summary(M2_quasi)$coefficients["Wt", "Pr(>|t|)"], 3)`) is
individually significant at $\alpha = 0.05$.

**Negative binomial regression as a comparison:**

```{r}
library(MASS)
M2_nb <- glm.nb(Sa ~ W + Wt, data = crab)
summary(M2_nb)
```

The NB model assumes $\text{Var}(Y) = \mu + \mu^2/\theta$. The
estimated $\hat{\theta} = `r round(M2_nb$theta, 2)`$ with a finite
standard error, confirming over-dispersion. The coefficient estimates
are similar to those from quasi-Poisson, and neither $W$ nor $Wt$ is
individually significant, consistent with the quasi-Poisson results.

\newpage

# Problem 2: Fish Parasites

Researchers examined a large number of fish to determine the prevalence
of parasites. The dataset (parasite.txt) includes the variables
Intensity (i.e., the number of parasites), Area (a categorical
variable), Year (to be treated as categorical), and Length of the fish.

```{r}
parasite <- read.table("HW5-parasite.txt", header = TRUE, sep = "\t")
parasite <- parasite %>%
  dplyr::select(Sample, Intensity, Year, Length, Area)
parasite$Area <- factor(parasite$Area)
parasite$Year <- factor(parasite$Year)
```

## (a) Poisson Model

**Fit a Poisson model with log link to the data with area, year, and
length as predictors. Interpret each model parameter.**

Model:
$\log(\mu_i) = \beta_0 + \beta_1 \text{Area}_2 + \beta_2 \text{Area}_3 + \beta_3 \text{Area}_4 + \beta_4 \text{Year}_{2000} + \beta_5 \text{Year}_{2001} + \beta_6 \text{Length}$

```{r}
fit_pois <- glm(Intensity ~ Area + Year + Length,
                family = poisson(link = "log"), data = parasite)
summary(fit_pois)
```

**Interpretation of parameters:**

```{r}
coefs <- round(coef(fit_pois), 4)
```

$\exp(\hat{\beta}_0) = `r round(exp(coefs[1]), 2)`$
$(= \exp(`r coefs[1]`))$ is the estimated expected parasite count for a
fish in Area 1, Year 1999, with Length = 0.

$\exp(\hat{\beta}_1)$ is the estimated rate ratio comparing Area 2 to
Area 1, holding year and length fixed. With
$\hat{\beta}_1 = `r coefs["Area2"]`$, the expected parasite count for
fish in Area 2 is $`r round(exp(coefs["Area2"]), 3)`$
$(= \exp(`r coefs["Area2"]`))$ times of that in Area 1.

$\exp(\hat{\beta}_2)$ is the estimated rate ratio comparing Area 3 to
Area 1, holding year and length fixed. With
$\hat{\beta}_2 = `r coefs["Area3"]`$, the expected parasite count for
fish in Area 3 is $`r round(exp(coefs["Area3"]), 3)`$
$(= \exp(`r coefs["Area3"]`))$ times of that in Area 1.

$\exp(\hat{\beta}_3)$ is the estimated rate ratio comparing Area 4 to
Area 1, holding year and length fixed. With
$\hat{\beta}_3 = `r coefs["Area4"]`$, the expected parasite count for
fish in Area 4 is $`r round(exp(coefs["Area4"]), 3)`$
$(= \exp(`r coefs["Area4"]`))$ times of that in Area 1.

$\exp(\hat{\beta}_4)$ is the estimated rate ratio comparing Year 2000
to Year 1999, holding area and length fixed. With
$\hat{\beta}_4 = `r coefs["Year2000"]`$, the expected parasite count in
2000 is $`r round(exp(coefs["Year2000"]), 3)`$
$(= \exp(`r coefs["Year2000"]`))$ times of that in 1999.

$\exp(\hat{\beta}_5)$ is the estimated rate ratio comparing Year 2001
to Year 1999, holding area and length fixed. With
$\hat{\beta}_5 = `r coefs["Year2001"]`$, the expected parasite count in
2001 is $`r round(exp(coefs["Year2001"]), 3)`$
$(= \exp(`r coefs["Year2001"]`))$ times of that in 1999.

$\exp(\hat{\beta}_6)$ is the estimated rate ratio for a one-unit
increase in fish length, holding area and year fixed. With
$\hat{\beta}_6 = `r coefs["Length"]`$, the expected parasite count is
$`r round(exp(coefs["Length"]), 4)`$
$(= \exp(`r coefs["Length"]`))$ times of that for a fish one unit
shorter.

## (b) Goodness of Fit

**Test for goodness of fit of the model in (a) and state conclusions.**

```{r}
dev_pois <- deviance(fit_pois)
df_pois <- fit_pois$df.residual
p_dev_pois <- 1 - pchisq(dev_pois, df_pois)

G_pois <- sum(residuals(fit_pois, type = "pearson")^2)
p_G_pois <- 1 - pchisq(G_pois, df_pois)

cat("Deviance =", round(dev_pois, 2), ", df =", df_pois,
    ", p =", round(p_dev_pois, 4), "\n")
cat("Pearson X² =", round(G_pois, 2), ", df =", df_pois,
    ", p =", round(p_G_pois, 4), "\n")
```

Both the deviance and Pearson $\chi^2$ are extremely large relative to
the degrees of freedom (p $\approx$ 0), indicating very poor fit. The
Poisson model does not adequately describe these data. This could be due
to over-dispersion, excess zeros, or both.

## (c) Zero-Inflated Poisson Model

**Researchers suspect that there may be two strains of fish, one that is
susceptible to parasites and one that is not. Without knowing which fish
are susceptible, this could be regarded as a zero-inflated model.
Building on the model in (a) (using the same predictors), fit an
appropriate model to the data that can account for extra zeros. Provide
an interpretation for each model parameter in terms of the problem.**

The Zero-Inflated Poisson (ZIP) model assumes two components. Let
$Z_i$ be a latent binary variable indicating whether fish $i$ is
susceptible ($Z_i = 1$) or non-susceptible ($Z_i = 0$), with
$P(Z_i = 0) = \pi_i$. Then:

- With probability $\pi_i$, $Y_i | (Z_i = 0) = 0$ (structural zero:
  non-susceptible fish always have 0 parasites).
- With probability $1 - \pi_i$,
  $Y_i | (Z_i = 1) \sim \text{Poisson}(\lambda_i)$.

Thus $P(Y_i = 0) = \pi_i + (1-\pi_i)e^{-\lambda_i}$, and for $y > 0$,
$P(Y_i = y) = (1-\pi_i)\frac{e^{-\lambda_i}\lambda_i^y}{y!}$.

We model both components using the same predictors
$\mathbf{x}_i = (1, \text{Area}_i, \text{Year}_i, \text{Length}_i)^T$:
$\log(\lambda_i) = \mathbf{x}_i^T\boldsymbol{\beta}$ (Poisson count
component) and
$\text{logit}(\pi_i) = \mathbf{x}_i^T\boldsymbol{\gamma}$
(zero-inflation component).

```{r}
fit_zip <- zeroinfl(Intensity ~ Area + Year + Length | Area + Year + Length,
                    data = parasite)
summary(fit_zip)
```

**Interpretation of parameters:**

*Count model (Poisson with log link) — for susceptible fish:*

```{r}
coefs_c <- round(coef(fit_zip, "count"), 4)
coefs_z <- round(coef(fit_zip, "zero"), 4)
```

$\exp(\hat{\beta}_0) = `r round(exp(coefs_c[1]), 2)`$
$(= \exp(`r coefs_c[1]`))$ is the estimated expected parasite count for
a susceptible fish in Area 1, Year 1999, at Length = 0.

$\exp(\hat{\beta}_{\text{Area2}})$ is the estimated rate ratio comparing
Area 2 to Area 1 among susceptible fish, holding year and length fixed.
With $\hat{\beta}_{\text{Area2}} = `r coefs_c["Area2"]`$, the expected
parasite count in Area 2 is $`r round(exp(coefs_c["Area2"]), 3)`$
$(= \exp(`r coefs_c["Area2"]`))$ times of that in Area 1.

$\exp(\hat{\beta}_{\text{Area3}})$ is the estimated rate ratio comparing
Area 3 to Area 1 among susceptible fish, holding year and length fixed.
With $\hat{\beta}_{\text{Area3}} = `r coefs_c["Area3"]`$, the expected
parasite count in Area 3 is $`r round(exp(coefs_c["Area3"]), 3)`$
$(= \exp(`r coefs_c["Area3"]`))$ times of that in Area 1.

$\exp(\hat{\beta}_{\text{Area4}})$ is the estimated rate ratio comparing
Area 4 to Area 1 among susceptible fish, holding year and length fixed.
With $\hat{\beta}_{\text{Area4}} = `r coefs_c["Area4"]`$, the expected
parasite count in Area 4 is $`r round(exp(coefs_c["Area4"]), 3)`$
$(= \exp(`r coefs_c["Area4"]`))$ times of that in Area 1.

$\exp(\hat{\beta}_{\text{Year2000}})$ is the estimated rate ratio
comparing Year 2000 to 1999 among susceptible fish, holding area and
length fixed. With
$\hat{\beta}_{\text{Year2000}} = `r coefs_c["Year2000"]`$, the expected
parasite count in 2000 is $`r round(exp(coefs_c["Year2000"]), 3)`$
$(= \exp(`r coefs_c["Year2000"]`))$ times of that in 1999.

$\exp(\hat{\beta}_{\text{Year2001}})$ is the estimated rate ratio
comparing Year 2001 to 1999 among susceptible fish, holding area and
length fixed. With
$\hat{\beta}_{\text{Year2001}} = `r coefs_c["Year2001"]`$, the expected
parasite count in 2001 is $`r round(exp(coefs_c["Year2001"]), 3)`$
$(= \exp(`r coefs_c["Year2001"]`))$ times of that in 1999.

$\exp(\hat{\beta}_{\text{Length}})$ is the estimated rate ratio for a
one-unit increase in length among susceptible fish, holding area and
year fixed. With
$\hat{\beta}_{\text{Length}} = `r coefs_c["Length"]`$, the expected
parasite count is $`r round(exp(coefs_c["Length"]), 4)`$
$(= \exp(`r coefs_c["Length"]`))$ times of that for a fish one unit
shorter.

*Zero-inflation model (logistic with logit link) — probability of being
non-susceptible:*

$\exp(\hat{\gamma}_0) / (1 + \exp(\hat{\gamma}_0)) = `r round(exp(coefs_z[1]) / (1 + exp(coefs_z[1])), 3)`$
is the estimated probability of being non-susceptible for a fish in
Area 1, Year 1999, at Length = 0.

$\exp(\hat{\gamma}_{\text{Area2}})$ is the estimated odds ratio of being
non-susceptible comparing Area 2 to Area 1, holding year and length
fixed. With $\hat{\gamma}_{\text{Area2}} = `r coefs_z["Area2"]`$, the
odds of being non-susceptible in Area 2 is
$`r round(exp(coefs_z["Area2"]), 3)`$
$(= \exp(`r coefs_z["Area2"]`))$ times of that in Area 1.

$\exp(\hat{\gamma}_{\text{Area3}})$ is the estimated odds ratio of being
non-susceptible comparing Area 3 to Area 1, holding year and length
fixed. With $\hat{\gamma}_{\text{Area3}} = `r coefs_z["Area3"]`$, the
odds of being non-susceptible in Area 3 is
$`r round(exp(coefs_z["Area3"]), 3)`$
$(= \exp(`r coefs_z["Area3"]`))$ times of that in Area 1.

$\exp(\hat{\gamma}_{\text{Area4}})$ is the estimated odds ratio of being
non-susceptible comparing Area 4 to Area 1, holding year and length
fixed. With $\hat{\gamma}_{\text{Area4}} = `r coefs_z["Area4"]`$, the
odds of being non-susceptible in Area 4 is
$`r round(exp(coefs_z["Area4"]), 3)`$
$(= \exp(`r coefs_z["Area4"]`))$ times of that in Area 1.

$\exp(\hat{\gamma}_{\text{Year2000}})$ is the estimated odds ratio of
being non-susceptible comparing Year 2000 to 1999, holding area and
length fixed. With
$\hat{\gamma}_{\text{Year2000}} = `r coefs_z["Year2000"]`$, the odds of
being non-susceptible in 2000 is
$`r round(exp(coefs_z["Year2000"]), 3)`$
$(= \exp(`r coefs_z["Year2000"]`))$ times of that in 1999.

$\exp(\hat{\gamma}_{\text{Year2001}})$ is the estimated odds ratio of
being non-susceptible comparing Year 2001 to 1999, holding area and
length fixed. With
$\hat{\gamma}_{\text{Year2001}} = `r coefs_z["Year2001"]`$, the odds of
being non-susceptible in 2001 is
$`r round(exp(coefs_z["Year2001"]), 3)`$
$(= \exp(`r coefs_z["Year2001"]`))$ times of that in 1999.

$\exp(\hat{\gamma}_{\text{Length}})$ is the estimated odds ratio of
being non-susceptible for a one-unit increase in length, holding area
and year fixed. With
$\hat{\gamma}_{\text{Length}} = `r coefs_z["Length"]`$, the odds of
being non-susceptible is
$`r round(exp(coefs_z["Length"]), 4)`$
$(= \exp(`r coefs_z["Length"]`))$ times of that for a fish one unit
shorter. The negative coefficient means longer fish are less likely to
be from the non-susceptible strain (i.e., longer fish are more likely
susceptible to parasites).
