---
title: "P8131 Homework 3"
author: "Yongyan Liu (yl6107)"
date: "2026-02-25"
output:
  pdf_document:
    latex_engine: xelatex
    extra_dependencies: ["amsmath", "amssymb", "amsthm"]
header-includes:
  - \usepackage{bm}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(knitr)
```

# Problem 1: Esophageal Cancer and Alcohol Consumption

**The following table gives data from a retrospective study between
esophageal cancer and daily alcohol consumption adjusted for age.**

| Age   | Case 0-79g | Case 80+g | Control 0-79g | Control 80+g |
|:-----:|:----------:|:---------:|:-------------:|:------------:|
| 25-34 | 0          | 1         | 106           | 9            |
| 35-44 | 5          | 4         | 164           | 26           |
| 45-54 | 21         | 25        | 138           | 29           |
| 55-64 | 34         | 42        | 139           | 27           |
| 65-74 | 36         | 19        | 88            | 18           |
| 75+   | 8          | 5         | 31            | 0            |

```{r}
# Input data
# Create data frame with all cells
age_midpoint <- c(25, 35, 45, 55, 65, 75)

# Case data
case_low <- c(0, 5, 21, 34, 36, 8)   # 0-79g alcohol
case_high <- c(1, 4, 25, 42, 19, 5)  # 80+g alcohol

# Control data
ctrl_low <- c(106, 164, 138, 139, 88, 31)   # 0-79g alcohol
ctrl_high <- c(9, 26, 29, 27, 18, 0)        # 80+g alcohol

# Create long-format data for prospective analysis
# Each row: age, alcohol (0=low, 1=high), case count, control count
data1 <- data.frame(
  age = rep(age_midpoint, each = 2),
  alcohol = rep(c(0, 1), 6),
  case = c(rbind(case_low, case_high)),
  control = c(rbind(ctrl_low, ctrl_high))
)

data1$total <- data1$case + data1$control
```

**(a) Fit a prospective model to the data to study the relation between
alcohol consumption, age, and disease (model age as a continuous
variable taking values 25, 35, 45, 55, 65, and 75). Interpret the
result.**

For a retrospective (case-control) study, although we sample based on
disease status, we can still fit a prospective logistic model to
estimate the odds ratio, since the odds ratio is symmetric (same whether
estimated prospectively or retrospectively).

The model:
$$\log\left(\frac{\pi}{1-\pi}\right) = \beta_0 + \beta_1 \cdot \text{Age} + \beta_2 \cdot \text{Alcohol}$$
where $\pi = P(\text{Disease} = 1 | \text{Age}, \text{Alcohol})$ and
Alcohol = 1 for 80+ g/day, 0 otherwise.

```{r}
# Fit prospective logistic regression model
fit_prospective <- glm(cbind(case, control) ~ age + alcohol,
                       family = binomial(link = "logit"), data = data1)
summary(fit_prospective)
```

```{r}
# Extract coefficients and confidence intervals
coefs <- summary(fit_prospective)$coefficients
ci <- confint.default(fit_prospective)

# Odds ratios
OR_age <- exp(coef(fit_prospective)["age"])
OR_alcohol <- exp(coef(fit_prospective)["alcohol"])

# OR confidence intervals
OR_ci_age <- exp(ci["age", ])
OR_ci_alcohol <- exp(ci["alcohol", ])

cat("Odds Ratio for Age (per 10-year increase):", round(OR_age^10, 4), "\n")
cat("95% CI:", round(OR_ci_age^10, 4), "\n\n")
cat("Odds Ratio for Alcohol (80+g vs 0-79g):", round(OR_alcohol, 4), "\n")
cat("95% CI:", round(OR_ci_alcohol, 4), "\n")
```

**Interpretation:**

1.  **Age effect:** $\hat{\beta}_1 = `r round(coef(fit_prospective)["age"], 5)`$
    (p < 0.001). For each 10-year increase in age, the odds of
    esophageal cancer increase by a factor of
    $e^{10 \times `r round(coef(fit_prospective)["age"], 5)`} = `r round(OR_age^10, 3)`$
    (95% CI: `r round(OR_ci_age[1]^10, 3)` -- `r round(OR_ci_age[2]^10, 3)`),
    holding alcohol consumption constant.

2.  **Alcohol effect:** $\hat{\beta}_2 = `r round(coef(fit_prospective)["alcohol"], 4)`$
    (p < 0.001). The odds of esophageal cancer for high alcohol
    consumption (80+ g/day) are $e^{`r round(coef(fit_prospective)["alcohol"], 4)`} = `r round(OR_alcohol, 3)`$
    times the odds for low alcohol consumption (0-79 g/day), adjusting
    for age (95% CI: `r round(OR_ci_alcohol[1], 3)` -- `r round(OR_ci_alcohol[2], 3)`).

3.  **Model fit:** The residual deviance is `r round(fit_prospective$deviance, 2)`
    on `r fit_prospective$df.residual` degrees of freedom. Comparing to
    $\chi^2_{0.95}(`r fit_prospective$df.residual`) = `r round(qchisq(0.95, fit_prospective$df.residual), 2)`$,
    the deviance exceeds the threshold, suggesting some lack of fit
    (p = `r round(1 - pchisq(fit_prospective$deviance, fit_prospective$df.residual), 4)`).
    This could be due to the assumption that age has a linear effect on
    the log-odds, or potential interaction between age and alcohol.

**Conclusion:** Both age and high alcohol consumption are significantly
associated with increased risk of esophageal cancer.

------------------------------------------------------------------------

**(b) Let $\Psi_j$ be the odds ratio relating alcohol consumption and
disease in the $j^{th}$ age group ($j = 1, ..., 6$). Assume different
age groups have different odds. Compare the following two models:**

-   $M_0: \Psi_j = 1$ for all $j$
-   $M_1: \Psi_j = \Psi$ (where $\Psi$ is an unknown constant)

**Model Formulation:**

For each age group $j$, let $\pi_{j,k}$ be the probability of disease
given alcohol level $k$ (0 = low, 1 = high). The odds ratio in age group
$j$ is:
$$\Psi_j = \frac{\pi_{j,1}/(1-\pi_{j,1})}{\pi_{j,0}/(1-\pi_{j,0})}$$

**Model $M_0$:** $\Psi_j = 1$ for all $j$ means alcohol has no effect on
disease within any age group.
$$\text{logit}(\pi) = \beta_0 + \beta_1 \cdot \text{Age}$$
This model includes only age (as a continuous variable), with no alcohol
effect.

**Model $M_1$:** $\Psi_j = \Psi$ (constant) means the odds ratio for
alcohol is the same across all age groups. For any age group $j$,
$$\text{logit}(\pi_{j,1}) - \text{logit}(\pi_{j,0}) = log(\Psi) = \beta_2$$
which means $\log(\Psi_j) = \beta_2$ is the same for all $j$. Therefore
$\Psi_j = e^{\beta_2} = \Psi$ is constant across all age groups, consistent
with the logistic regression model structure.
$$\text{logit}(\pi) = \beta_0 + \beta_1 \cdot \text{Age} + \beta_2 \cdot \text{Alcohol}$$

$M_0$ is nested within $M_1$ (set $\beta_2 = 0$ in $M_1$ to get $M_0$).

```{r}
# Fit M0: Only age effect (no alcohol effect, implies Psi_j = 1)
fit_M0 <- glm(cbind(case, control) ~ age,
              family = binomial(link = "logit"), data = data1)

# Fit M1: Age effect + constant alcohol effect (Psi_j = Psi for all j)
fit_M1 <- glm(cbind(case, control) ~ age + alcohol,
              family = binomial(link = "logit"), data = data1)

# Deviance analysis
dev_M0 <- fit_M0$deviance
dev_M1 <- fit_M1$deviance
df_M0 <- fit_M0$df.residual
df_M1 <- fit_M1$df.residual

# Test statistic: difference in deviances
dev_diff <- dev_M0 - dev_M1
df_diff <- df_M0 - df_M1
p_value <- 1 - pchisq(dev_diff, df_diff)

cat("Model M0 (Psi_j = 1): Deviance =", round(dev_M0, 4), ", df =", df_M0, "\n")
cat("Model M1 (Psi_j = Psi): Deviance =", round(dev_M1, 4), ", df =", df_M1, "\n")
cat("\nDeviance difference:", round(dev_diff, 4), "\n")
cat("df difference:", df_diff, "\n")
cat("Critical value chi^2_0.95(1):", round(qchisq(0.95, 1), 4), "\n")
cat("P-value:", format(p_value, scientific = TRUE, digits = 4), "\n")
```

```{r}
# Estimate of common odds ratio Psi
Psi_hat <- exp(coef(fit_M1)["alcohol"])
ci_Psi <- exp(confint.default(fit_M1)["alcohol", ])
cat("\nEstimate of common odds ratio Psi:", round(Psi_hat, 4), "\n")
cat("95% CI for Psi:", round(ci_Psi, 4), "\n")
```

**Deviance Analysis:**

Since $M_0$ is nested within $M_1$, we can use the likelihood ratio test:
$$D_{M_0} - D_{M_1} \sim \chi^2(df_{M_0} - df_{M_1}) \text{ under } H_0: \beta = 0$$

-   Test statistic: $`r round(dev_diff, 4)`$
-   Degrees of freedom: $`r df_diff`$
-   Critical value $\chi^2_{0.95}(1)$: $`r round(qchisq(0.95, 1), 4)`$
-   P-value: $`r format(p_value, scientific = TRUE, digits = 4)`$

**Conclusion:** Since the test statistic (`r round(dev_diff, 2)`) far exceeds
the critical value (3.84), we reject $M_0$ in favor of $M_1$ at the 0.05
significance level. There is strong evidence that alcohol consumption is
associated with esophageal cancer risk ($\Psi \neq 1$).

The estimated common odds ratio is $\hat{\Psi} = `r round(Psi_hat, 3)`$
(95% CI: `r round(ci_Psi[1], 3)` -- `r round(ci_Psi[2], 3)`), indicating
that high alcohol consumption (80+ g/day) is associated with
approximately `r round(Psi_hat, 1)` times the odds of esophageal cancer
compared to low consumption, across all age groups.

# Problem 2: Orobanche Seed Germination

**The following table provides data from a study of the germination of
two species of Orobanche seeds. The seeds were grown on 1/125 dilutions
of two different root extract media (cucumber or bean) in a 2 x 2
factorial layout with replicates. The data $(y_i/m_i)$ consist of the
number of seeds, $m_i$, and the number germinating, $y_i$, for each
batch. Interest focuses on the possible differences in germination rates
for the two types of seed and root extract.**

```{r}
# Input germination data
# O. aegyptiaca 75
bean_75 <- data.frame(
  germ = c(10, 23, 23, 26, 17),
  total = c(39, 62, 81, 51, 39),
  seed = "O. aegyptiaca 75",
  extract = "Bean"
)

cucumber_75 <- data.frame(
  germ = c(5, 53, 55, 32, 46, 10),
  total = c(6, 74, 72, 51, 79, 13),
  seed = "O. aegyptiaca 75",
  extract = "Cucumber"
)

# O. aegyptiaca 73
bean_73 <- data.frame(
  germ = c(8, 10, 8, 23, 0),
  total = c(16, 30, 28, 45, 4),
  seed = "O. aegyptiaca 73",
  extract = "Bean"
)

cucumber_73 <- data.frame(
  germ = c(3, 22, 15, 32, 3),
  total = c(12, 41, 30, 51, 7),
  seed = "O. aegyptiaca 73",
  extract = "Cucumber"
)

# Combine all data
germ_data <- rbind(bean_75, cucumber_75, bean_73, cucumber_73)
germ_data$not_germ <- germ_data$total - germ_data$germ
germ_data$seed <- factor(germ_data$seed)
germ_data$extract <- factor(germ_data$extract)

# Display data summary
kable(germ_data[, c("seed", "extract", "germ", "total")],
      col.names = c("Seed Type", "Extract", "Germinated", "Total"),
      caption = "Orobanche Seed Germination Data")
```

**(a) Fit a logistic regression model to study the relation between
germination rates and different types of seed and root extract.
Interpret the result.**

Model:
$$\text{logit}(\pi_i) = \beta_0 + \beta_1 \cdot \text{Seed} + \beta_2 \cdot \text{Extract}$$
where Seed = 1 for O. aegyptiaca 75, Extract = 1 for Cucumber.

```{r}
# Fit logistic regression with main effects
fit_germ <- glm(cbind(germ, not_germ) ~ seed + extract,
                family = binomial(link = "logit"), data = germ_data)
summary(fit_germ)
```

```{r}
# Odds ratios and confidence intervals
OR <- exp(coef(fit_germ))
ci <- exp(confint.default(fit_germ))

results <- data.frame(
  Coefficient = round(coef(fit_germ), 4),
  OR = round(OR, 4),
  CI_lower = round(ci[, 1], 4),
  CI_upper = round(ci[, 2], 4),
  p_value = round(summary(fit_germ)$coefficients[, 4], 4)
)
kable(results, caption = "Logistic Regression Results (Standard Binomial)")
```

**Interpretation:**

1.  **Seed type effect:** The coefficient for seed type (O. aegyptiaca 75
    vs 73) is $\hat{\beta}_1 = `r round(coef(fit_germ)["seedO. aegyptiaca 75"], 4)`$
    with OR = `r round(OR["seedO. aegyptiaca 75"], 3)`. O. aegyptiaca 75
    has `r round(OR["seedO. aegyptiaca 75"], 3)` times the odds of
    germination compared to O. aegyptiaca 73 (p = `r round(summary(fit_germ)$coefficients["seedO. aegyptiaca 75", 4], 4)`).
    Note that the 95% CI (`r round(ci["seedO. aegyptiaca 75", 1], 3)` -- `r round(ci["seedO. aegyptiaca 75", 2], 3)`)
    includes 1, suggesting the effect may not be statistically significant.

2.  **Extract effect:** The coefficient for extract type (Cucumber vs
    Bean) is $\hat{\beta}_2 = `r round(coef(fit_germ)["extractCucumber"], 4)`$
    with OR = `r round(OR["extractCucumber"], 3)`. Seeds grown on
    cucumber extract have `r round(OR["extractCucumber"], 3)` times the
    odds of germination compared to bean extract (p = `r round(summary(fit_germ)$coefficients["extractCucumber", 4], 4)`).

3.  **Model deviance:** Residual deviance = `r round(fit_germ$deviance, 2)`
    on `r fit_germ$df.residual` df. Compared to $\chi^2_{0.95}(`r fit_germ$df.residual`) = `r round(qchisq(0.95, fit_germ$df.residual), 2)`$,
    the deviance is much larger, suggesting lack of fit which could indicate
    overdispersion.

------------------------------------------------------------------------

**(b) Is there overdispersion? If so, what is the estimate of dispersion
parameter? Update your model and reinterpret the result.**

```{r}
# Check for overdispersion
# Pearson chi-squared statistic
pearson_resid <- residuals(fit_germ, type = "pearson")
G <- sum(pearson_resid^2)
n <- nrow(germ_data)
p <- length(coef(fit_germ))

# Estimate dispersion parameter
phi_hat <- G / (n - p)

cat("Pearson chi-squared statistic G =", round(G, 4), "\n")
cat("Degrees of freedom (n - p) =", n - p, "\n")
cat("Estimated dispersion parameter phi_hat = G/(n-p) =", round(phi_hat, 4), "\n")
cat("Alternative estimate using deviance: D/(n-p) =",
    round(fit_germ$deviance / (n - p), 4), "\n")
```

**Assessment of overdispersion:**

Under the standard binomial model, $E[G] \approx n - p = `r n - p`$ when
the model is correct. We observe $G = `r round(G, 2)`$, which is
substantially larger than expected.

The estimated dispersion parameter is:
$$\hat{\phi} = \frac{G}{n - p} = \frac{`r round(G, 2)`}{`r n - p`} = `r round(phi_hat, 3)`$$

Since $\hat{\phi} = `r round(phi_hat, 3)` > 1$, there is evidence of
**overdispersion** in the data.

```{r}
# Adjust for overdispersion by specifying dispersion parameter
summary_adj <- summary(fit_germ, dispersion = phi_hat)
summary_adj
```

```{r}
# Updated results with overdispersion adjustment
# Note: coefficients are the same, only SEs and p-values change
se_adj <- summary_adj$coefficients[, 2]
ci_adj <- cbind(
  coef(fit_germ) - 1.96 * se_adj,
  coef(fit_germ) + 1.96 * se_adj
)

results_adj <- data.frame(
  Coefficient = round(coef(fit_germ), 4),
  SE = round(se_adj, 4),
  OR = round(OR, 4),
  CI_lower = round(exp(ci_adj[, 1]), 4),
  CI_upper = round(exp(ci_adj[, 2]), 4),
  p_value = round(summary_adj$coefficients[, 4], 4)
)
kable(results_adj,
      caption = "Results Adjusted for Overdispersion")
```

**Comparison of standard errors:**

```{r}
# Compare standard errors
se_comparison <- data.frame(
  Parameter = names(coef(fit_germ)),
  SE_binomial = round(summary(fit_germ)$coefficients[, 2], 4),
  SE_adjusted = round(se_adj, 4),
  Ratio = round(se_adj / summary(fit_germ)$coefficients[, 2], 4)
)
kable(se_comparison, caption = "Standard Error Comparison")
```

**Reinterpretation with overdispersion adjustment:**

After accounting for overdispersion ($\hat{\phi} = `r round(phi_hat, 3)`$):

1.  **Seed type effect:** O. aegyptiaca 75 still has higher germination
    odds than O. aegyptiaca 73 (OR = `r round(OR["seedO. aegyptiaca 75"], 3)`),
    but the effect is not statistically significant at $\alpha = 0.05$
    (p = `r round(summary_adj$coefficients["seedO. aegyptiaca 75", 4], 4)`).

2.  **Extract effect:** Cucumber extract is associated with higher
    germination odds than bean extract (OR = `r round(OR["extractCucumber"], 3)`),
    and this effect remains statistically significant at $\alpha = 0.05$
    (p = `r round(summary_adj$coefficients["extractCucumber", 4], 4)`).

3.  **Standard errors:** The standard errors are inflated by a factor of
    $\sqrt{\hat{\phi}} = \sqrt{`r round(phi_hat, 3)`} = `r round(sqrt(phi_hat), 3)`$
    compared to the standard binomial model.

**Conclusion:** After adjusting for overdispersion, extract type remains
statistically significant while seed type does not. Cucumber extract is
associated with higher germination rates than bean extract.

------------------------------------------------------------------------

**(c) What is a plausible cause of the overdispersion?**

1.  **Intra-batch correlation:** Seeds within the same batch share
    environmental conditions, leading to correlated outcomes
    ($\text{Var}(Y) = m\pi(1-\pi)\{1 + (m-1)\rho\} > m\pi(1-\pi)$).

2.  **Batch-to-batch heterogeneity:** Unmeasured factors (seed quality,
    storage conditions) may cause germination probability to vary across
    batches beyond what seed type and extract explain.
