---
title: "P8109 Statistical Inference -- Homework 4"
author: "Yongyan Liu (yl6107)"
date: "2026-04-15"
output:
  pdf_document:
    latex_engine: xelatex
header-includes:
  - \usepackage{amsmath,amssymb,amsthm}
  - \DeclareMathOperator{\Var}{Var}
  - \DeclareMathOperator{\Cov}{Cov}
  - \DeclareMathOperator{\E}{E}
  - \newcommand{\iid}{\overset{\text{iid}}{\sim}}
  - \newcommand{\ind}{\overset{\text{ind}}{\sim}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Problem 1

**Consider sampling from the parent population with density**
$$f(x\mid\theta) = \theta x^{\theta-1}, \qquad 0 < x < 1,\; \theta > 0.$$
**Based on a random sample $X_1$ of size $n=1$ from this parent population, what is the power of the most powerful (MP) test of $H_0: \theta = 1$ vs. $H_1: \theta = 2$ if $P(\text{Type I error}) = \alpha = 0.05$?**

Under $H_0$, $f(x\mid 1) = 1$ (so $X_1 \sim \text{Uniform}(0,1)$); under $H_1$, $f(x\mid 2) = 2x$. By the Neyman--Pearson Lemma, the MP test rejects $H_0$ when the likelihood ratio
$$\Lambda(x_1) \;=\; \frac{f(x_1\mid 2)}{f(x_1\mid 1)} \;=\; 2x_1$$
is large. Since $\Lambda$ is strictly increasing in $x_1$, the rejection region is of the form
$$R = \{x_1 : x_1 \ge c\}$$
for some constant $c \in (0,1)$.

Under $H_0$, $X_1 \sim \text{Uniform}(0,1)$, so
$$\alpha \;=\; P_{\theta=1}(X_1 \ge c) \;=\; 1 - c \;=\; 0.05 \;\Longrightarrow\; c = 0.95.$$
Under $H_1$, $f(x\mid 2) = 2x$, so
$$\text{Power} \;=\; P_{\theta = 2}(X_1 \ge 0.95) \;=\; \int_{0.95}^{1} 2x\,dx \;=\; \bigl[x^2\bigr]_{0.95}^{1} \;=\; 1 - 0.9025 \;=\; 0.0975.$$

# Problem 2

**Consider sampling from the parent population with density**
$$f(x\mid\theta) = \theta x^{\theta - 1}, \qquad 0 < x < 1,\; \theta > 0.$$
**If $X_1$ and $X_2$ constitute a random sample of size $n=2$, show that the rejection region of size $\alpha$ for the MP test of $H_0: \theta = 1$ vs.\ $H_1: \theta = 2$ is $X_1 X_2 \ge k^{*}$, where $k^{*} - k^{*}\log k^{*} = 1 - \alpha$.**

Under independence the joint densities are
$$L(1\mid x_1,x_2) = 1, \qquad L(2\mid x_1,x_2) = 4 x_1 x_2.$$
Hence
$$\Lambda(x_1,x_2) \;=\; \frac{L(2\mid x_1,x_2)}{L(1\mid x_1,x_2)} \;=\; 4\,x_1 x_2,$$
which is a strictly increasing function of the sufficient-like statistic $T = X_1 X_2$. By the Neyman--Pearson Lemma, the MP rejection region has the form
$$R \;=\; \{(x_1,x_2) : x_1 x_2 \ge k^{*}\}.$$

Under $\theta = 1$, $X_1,X_2 \iid \text{Uniform}(0,1)$, so $-\log X_i \iid \text{Exponential}(1)$ and
$$W \;=\; -\log X_1 - \log X_2 \;\sim\; \text{Gamma}(2,1),$$
with density $f_W(w) = w e^{-w}$, $w > 0$, and (by the Erlang tail / Poisson relationship)
$$P(W \ge w) \;=\; e^{-w}(1 + w), \qquad w \ge 0.$$

The event $\{X_1 X_2 \ge k^{*}\}$ is equivalent to $\{W \le -\log k^{*}\}$. Setting the size equal to $\alpha$,
$$\alpha \;=\; P_{H_0}(X_1 X_2 \ge k^{*}) \;=\; 1 - P(W \ge -\log k^{*}) \;=\; 1 - e^{\log k^{*}}\bigl(1 - \log k^{*}\bigr) \;=\; 1 - \bigl(k^{*} - k^{*}\log k^{*}\bigr),$$
so $k^{*}$ satisfies
$$k^{*} - k^{*}\log k^{*} \;=\; 1 - \alpha,$$
which is the stated form.

# Problem 3

**Let $Y_1,\ldots,Y_n$ be a random sample from the density**
$$f(y\mid\theta) \;=\; \tfrac{1}{2}(1 + \theta y), \qquad -1 < y < 1,\; \theta > -1.$$

(a) Under $H_0$, $f(y\mid 0) = 1/2$ on $(-1,1)$ (so the $Y_i$ are $\text{Uniform}(-1,1)$); under $H_1$, $f(y\mid 1) = (1+y)/2$. By the NP Lemma the MP test rejects for large values of
$$\Lambda(\mathbf{y}) \;=\; \frac{\prod_{i=1}^{n}\tfrac{1}{2}(1+y_i)}{\prod_{i=1}^{n}\tfrac{1}{2}} \;=\; \prod_{i=1}^{n}(1+y_i).$$
Because each factor $(1 + y_i) > 0$ on the support, $\Lambda$ is strictly monotone increasing in the statistic $T(\mathbf{Y}) = \prod_{i=1}^{n}(1 + Y_i)$ (equivalently in $\sum_{i=1}^{n}\log(1 + Y_i)$). Hence the MP rejection region is
$$R \;=\; \left\{\mathbf{y} : \prod_{i=1}^{n}(1 + y_i) \ge c\right\},$$
with the constant $c$ determined by $P_{\theta=0}\!\left(\prod_{i=1}^{n}(1+Y_i) \ge c\right) = \alpha$.

(b) With $n = 1$ the region reduces to $\{Y_1 \ge c\}$ (since $1 + y$ is monotone in $y$). Under $H_0$, $Y_1 \sim \text{Uniform}(-1,1)$ with density $1/2$, so
$$0.05 \;=\; P_{\theta = 0}(Y_1 \ge c) \;=\; \int_{c}^{1} \tfrac{1}{2}\,dy \;=\; \tfrac{1 - c}{2}.$$
Solving, $1 - c = 0.10$, so
$$c \;=\; 0.90.$$

(c) Under $H_1$ the density is $f(y\mid 1) = (1+y)/2$, so
$$\text{Power} \;=\; P_{\theta=1}(Y_1 \ge 0.9) \;=\; \int_{0.9}^{1} \tfrac{1+y}{2}\,dy \;=\; \left[\frac{y}{2} + \frac{y^{2}}{4}\right]_{0.9}^{1} \;=\; \left(\tfrac{1}{2} + \tfrac{1}{4}\right) - \left(0.45 + 0.2025\right).$$
That is,
$$\text{Power} \;=\; 0.75 - 0.6525 \;=\; 0.0975.$$

# Problem 4

**Suppose $X \sim \text{Poisson}(\lambda)$. A sample of size $n = 10$ is used to test $H_0: \lambda = 0.1$ vs.\ $H_1: \lambda > 0.1$. Construct the randomized MP test of size $\alpha = 0.05$.**

Let $T = \sum_{i=1}^{10} X_i$. Since $X_1,\ldots,X_{10} \iid \text{Poisson}(\lambda)$ form a one-parameter exponential family with natural statistic $T$, the sum $T$ is sufficient for $\lambda$, and moreover $T \sim \text{Poisson}(n\lambda)$. Under $H_0$,
$$T \;\sim\; \text{Poisson}(n\lambda_0) \;=\; \text{Poisson}(1).$$

For any fixed $\lambda_1 > 0.1$, the joint pmf ratio
$$\frac{L(\lambda_1 \mid \mathbf{x})}{L(\lambda_0 \mid \mathbf{x})} \;=\; \exp\!\bigl(-n(\lambda_1 - \lambda_0)\bigr)\left(\frac{\lambda_1}{\lambda_0}\right)^{T}$$
is strictly increasing in $T$ (since $\lambda_1/\lambda_0 > 1$). Hence the Poisson family has monotone likelihood ratio (MLR) in $T$, and by Theorem 8.1 (Lecture 8) a test that rejects for large values of $T$ is UMP for the composite alternative $H_1: \lambda > 0.1$ at its size. Since $T$ is discrete, an exact size-$\alpha$ test requires randomization at the boundary.

We seek the integer $k$ such that
$$P_{H_0}(T \ge k+1) \;\le\; \alpha \;<\; P_{H_0}(T \ge k).$$
Using $T \sim \text{Poisson}(1)$:

```{r p4-poisson}
alpha <- 0.05
k_vals <- 0:6
tail_ge <- ppois(k_vals - 1, lambda = 1, lower.tail = FALSE)  # P(T >= k)
pmf    <- dpois(k_vals, lambda = 1)                           # P(T = k)
data.frame(k = k_vals,
           P_T_ge_k  = round(tail_ge, 6),
           P_T_eq_k  = round(pmf,    6))
```

From the table,
$$P_{H_0}(T \ge 4) \;\approx\; 0.01899, \qquad P_{H_0}(T \ge 3) \;\approx\; 0.08030,$$
so $P_{H_0}(T \ge 4) < 0.05 < P_{H_0}(T \ge 3)$. Therefore we take $k = 3$ and randomize on the boundary $\{T = 3\}$.

We require
$$P_{H_0}(T > 3) \;+\; \gamma\, P_{H_0}(T = 3) \;=\; \alpha,$$
i.e.,
$$0.018988 + \gamma\cdot 0.061313 \;=\; 0.05 \;\Longrightarrow\; \gamma \;=\; \frac{0.05 - 0.018988}{0.061313} \;\approx\; 0.5058 \in [0,1].$$

The randomized MP (UMP) test is, writing $T = \sum_{i=1}^{10} X_i$,
$$\phi(\mathbf{x}) \;=\; \begin{cases} 1, & T > 3, \\[2pt] \gamma \approx 0.5058, & T = 3, \\[2pt] 0, & T < 3. \end{cases}$$
Operationally: compute $T$; if $T \ge 4$ reject $H_0$; if $T \le 2$ do not reject; if $T = 3$, draw $U \sim \text{Uniform}(0,1)$ independently of the data and reject iff $U < \gamma$.

By construction $\E_{H_0}[\phi(\mathbf X)] = P_{H_0}(T \ge 4) + \gamma\, P_{H_0}(T = 3) = \alpha$, so the test has size exactly $0.05$. By the Neyman--Pearson Lemma and the MLR property of the Poisson family, this randomized test is the uniformly most powerful (and hence MP) test of size $0.05$ for $H_0: \lambda = 0.1$ against $H_1: \lambda > 0.1$.
