---
title: "P8131 Homework 7"
author: "Yongyan Liu (yl6107)"
date: "2026-04-06"
output:
  pdf_document:
    latex_engine: xelatex
header-includes:
  - \usepackage{amsmath}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(nlme)
library(lme4)
```

# Problem 1

**In a pitch study, we are interested in the relationship between pitch
and politeness. There are two levels of politeness (a formal register:
pol, and an informal register: inf). On top of that, we also have an
additional fixed effect, gender. Each subject was tested on several
scenarios (e.g., asking a peer for a favor (informal condition) or
asking a professor for a favor (formal condition)). The pitch
measurements are typically correlated for the same subject and in the
same scenario.**

```{r}
pitch <- read.csv("HW7_politeness_data.csv")
pitch$subject <- factor(pitch$subject)
pitch$gender <- factor(pitch$gender)
pitch$attitude <- factor(pitch$attitude)
pitch$scenario <- factor(pitch$scenario)
```

## (a) Exploratory analysis

**Provide boxplots to show the relation between gender/attitude and
pitch (ignoring different scenarios).**

```{r, fig.width=7, fig.height=4}
ggplot(pitch, aes(x = interaction(gender, attitude), y = frequency,
                  fill = attitude)) +
  geom_boxplot() +
  labs(x = "Gender and Attitude", y = "Pitch (Hz)",
       title = "Pitch by Gender and Attitude",
       fill = "Attitude") +
  scale_x_discrete(labels = c("Female\nInformal", "Male\nInformal",
                               "Female\nPolite", "Male\nPolite")) +
  theme_bw()
```

Females tend to have higher pitch than males across both attitude
conditions. Within each gender, the polite condition tends to have
slightly lower pitch than the informal condition, suggesting that
speakers lower their pitch when being polite.

## (b) Mixed effects model with random intercepts for subjects

**Fit a mixed effects model with random intercepts for different
subjects (gender and attitude being the fixed effects). What is the
covariance matrix for a subject $Y_i$? What is the covariance matrix
for the estimates of fixed effects (Hint: $3 \times 3$ matrix for
intercept, gender and attitude)? What are the BLUPs for
subject-specific intercepts? What are the residuals?**

```{r}
LMM1 <- lme(frequency ~ gender + attitude, random = ~ 1 | subject,
            data = pitch, method = "REML")
summary(LMM1)
```

### Covariance matrix for $Y_i$

This is a random intercept model:
$Y_{ij} = \beta_0 + \beta_1 \cdot \text{gender}_i + \beta_2 \cdot \text{attitude}_{ij} + b_i + \epsilon_{ij}$,
where $b_i \sim N(0, \sigma_b^2)$ and
$\epsilon_{ij} \sim N(0, \sigma^2)$, independently.

From the random intercept model theory (Lecture 17, slide 6), the
marginal covariance of $Y_i$ has compound symmetry structure:

$$\text{Cov}(Y_i) = \begin{pmatrix} \sigma_b^2 + \sigma^2 & \sigma_b^2 & \cdots & \sigma_b^2 \\ \sigma_b^2 & \sigma_b^2 + \sigma^2 & \cdots & \sigma_b^2 \\ \vdots & \vdots & \ddots & \vdots \\ \sigma_b^2 & \sigma_b^2 & \cdots & \sigma_b^2 + \sigma^2 \end{pmatrix}_{14 \times 14}$$

since each subject has $n_i = 14$ observations (7 scenarios $\times$ 2
attitudes).

```{r}
sigma_b <- as.numeric(VarCorr(LMM1)[1, 2]) # SD of random intercept
sigma_e <- as.numeric(VarCorr(LMM1)[2, 2]) # residual SD
sigma_b_sq <- sigma_b^2
sigma_e_sq <- sigma_e^2
cat("sigma_b^2 (random intercept variance):", round(sigma_b_sq, 2), "\n")
cat("sigma^2 (residual variance):", round(sigma_e_sq, 2), "\n")
cat("Diagonal entries:", round(sigma_b_sq + sigma_e_sq, 2), "\n")
cat("Off-diagonal entries:", round(sigma_b_sq, 2), "\n")
```

Plugging in the estimates, the $14 \times 14$ covariance matrix has
diagonal entries $\hat{\sigma}_b^2 + \hat{\sigma}^2$ and off-diagonal
entries $\hat{\sigma}_b^2$, with values shown above.

### Covariance matrix for the fixed effects estimates

The $3 \times 3$ covariance matrix of $\hat{\boldsymbol{\beta}} = (\hat{\beta}_0, \hat{\beta}_1, \hat{\beta}_2)^T$ is obtained via `vcov()`:

```{r}
vcov(LMM1)
```

### BLUPs for subject-specific intercepts

The BLUPs $\hat{b}_i$ represent each subject's predicted deviation from
the population mean intercept:

```{r}
random.effects(LMM1)
```

### Residuals

The within-subject residuals $\hat{\epsilon}_{ij} = Y_{ij} - \hat{Y}_{ij}$, where $\hat{Y}_{ij}$ includes both fixed effects and random intercepts:

```{r}
LMM1$residuals
```

## (c) Model with interaction and likelihood ratio test

**Fit a mixed effects model with intercepts for different subjects
(gender, attitude and their interaction being the fixed effects). Use
likelihood ratio test to compare this model with the model in part (b)
to determine whether the interaction term is significantly associated
with pitch.**

For a valid likelihood ratio test comparing models with different fixed
effects, both models must be fit using maximum likelihood (ML), not
REML (Lecture 17, slide 20).

```{r}
# Refit both models with ML
LMM1.ml <- lme(frequency ~ gender + attitude, random = ~ 1 | subject,
               data = pitch, method = "ML")
LMM2.ml <- lme(frequency ~ gender * attitude, random = ~ 1 | subject,
               data = pitch, method = "ML")
summary(LMM2.ml)
```

```{r}
lrt_result <- anova(LMM1.ml, LMM2.ml)
lrt_result
```

**Hypotheses:**

- $H_0$: $\beta_3 = 0$ (no gender-by-attitude interaction)
- $H_1$: $\beta_3 \neq 0$

The likelihood ratio test statistic is `r round(lrt_result$L.Ratio[2], 4)` with `r lrt_result$df[2] - lrt_result$df[1]` degree of freedom and a p-value of `r round(lrt_result$"p-value"[2], 4)`. Based on the p-value from the test, the interaction between gender and attitude is
not statistically significant at $\alpha = 0.05$, indicating that the
effect of attitude on pitch does not differ significantly between
genders.

## (d) Crossed random effects model

**Write out the mixed effects model with random intercepts for both
subjects and scenarios (gender and attitude being the fixed effects).
Fit the model using `lmer` in the `lme4` package. Write out the
covariance matrix for a subject $Y_i$. What is the interpretation of
the coefficient for the fixed effect term attitude?**

### Model specification

$$Y_{ijk} = \beta_0 + \beta_1 \cdot \text{gender}_i + \beta_2 \cdot \text{attitude}_{k} + b_i + c_j + \epsilon_{ijk},$$

where:

- $i = 1, \ldots, 6$ indexes subjects, $j = 1, \ldots, 7$ indexes
  scenarios, $k$ indexes attitude level,
- $b_i \sim N(0, \sigma_b^2)$ is the random intercept for subject $i$,
- $c_j \sim N(0, \sigma_c^2)$ is the random intercept for scenario $j$,
- $\epsilon_{ijk} \sim N(0, \sigma^2)$ is the residual error,
- $b_i$, $c_j$, and $\epsilon_{ijk}$ are mutually independent.

### Model fitting

```{r}
LMM3 <- lmer(frequency ~ gender + attitude + (1 | subject) + (1 | scenario),
             data = pitch)
summary(LMM3)
```

### Covariance matrix for $Y_i$

For subject $i$, $Y_i$ is a $14 \times 1$ vector (7 scenarios $\times$
2 attitudes). The covariance structure now depends on whether two
observations share the same scenario:

- **Diagonal entries:** $\text{Var}(Y_{ijk}) = \sigma_b^2 + \sigma_c^2 + \sigma^2$.
- **Off-diagonal, same scenario** (same $j$, different attitude):
  $\text{Cov}(Y_{ijk}, Y_{ijk'}) = \sigma_b^2 + \sigma_c^2$ (both
  random effects shared).
- **Off-diagonal, different scenario** (different $j$, same subject):
  $\text{Cov}(Y_{ijk}, Y_{ij'k'}) = \sigma_b^2$ (only subject random
  effect shared).

```{r}
vc <- as.data.frame(VarCorr(LMM3))
sigma_b2 <- vc$vcov[vc$grp == "subject"]
sigma_c2 <- vc$vcov[vc$grp == "scenario"]
sigma_e2 <- vc$vcov[vc$grp == "Residual"]
cat("sigma_b^2 (subject):", round(sigma_b2, 2), "\n")
cat("sigma_c^2 (scenario):", round(sigma_c2, 2), "\n")
cat("sigma^2 (residual):", round(sigma_e2, 2), "\n")
cat("\nDiagonal entries:", round(sigma_b2 + sigma_c2 + sigma_e2, 2), "\n")
cat("Off-diagonal (same scenario):", round(sigma_b2 + sigma_c2, 2), "\n")
cat("Off-diagonal (different scenario):", round(sigma_b2, 2), "\n")
```

If we order the observations within subject $i$ by scenario
(scenario 1: pol, inf; scenario 2: pol, inf; $\ldots$; scenario 7: pol,
inf), the $14 \times 14$ covariance matrix has a block structure. Each
$2 \times 2$ diagonal block (within a scenario) has diagonal
$\sigma_b^2 + \sigma_c^2 + \sigma^2$ and off-diagonal
$\sigma_b^2 + \sigma_c^2$. All entries in off-diagonal blocks (between
different scenarios) equal $\sigma_b^2$.

### Interpretation of the attitude coefficient

```{r}
fixef(LMM3)["attitudepol"]
```

The coefficient for attitude (polite vs. informal as reference) has a
population-average (marginal) interpretation: holding gender constant,
the mean pitch is estimated to change by this amount (in Hz) when
switching from the informal register to the polite register, averaged
across all subjects and scenarios. A negative value indicates that, on
average, speakers use a lower pitch in polite settings compared to
informal settings.
