---
title: "P8109 Statistical Inference -- Homework 1"
author: "yl6107"
date: "2026-02-04"
output:
  pdf_document:
    latex_engine: xelatex
header-includes:
  - \usepackage{amsmath,amssymb,amsthm,bbm}
  - \DeclareMathOperator{\Var}{Var}
  - \DeclareMathOperator{\Cov}{Cov}
  - \DeclareMathOperator{\E}{E}
  - \newcommand{\iid}{\overset{\text{iid}}{\sim}}
  - \newcommand{\dto}{\xrightarrow{d}}
  - \newcommand{\pto}{\xrightarrow{p}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Problem 1

**Let $Z_i$, $i=1,\ldots,6$, be iid $N(0,1)$ random variables. Using a statistical software, calculate**
$$P\!\left(\sum_{i=1}^{6}Z_i^2 < 6\right),\qquad P\!\left(\sum_{i=1}^{4}Z_i^2 > 3\sum_{i=5}^{6}Z_i^2\right).$$

**Solution.**

Since $Z_i \iid N(0,1)$, we have $\sum_{i=1}^6 Z_i^2 \sim \chi^2_6$. Thus
$$P\!\left(\sum_{i=1}^6 Z_i^2 < 6\right) = F_{\chi^2_6}(6).$$

For the second probability, let $U = \sum_{i=1}^4 Z_i^2 \sim \chi^2_4$ and $V = \sum_{i=5}^6 Z_i^2 \sim \chi^2_2$, which are independent. Then
$$P(U > 3V) = P\!\left(\frac{U}{V} > 3\right) = P\!\left(\frac{U/4}{V/2} > \frac{3}{2}\right) = P\!\left(F > \frac{3}{2}\right),$$
where $F = \frac{U/4}{V/2} \sim F_{4,2}$.

```{r}
# First probability
pchisq(6, df = 6)

# Second probability
1 - pf(3/2, df1 = 4, df2 = 2)
```

\newpage

# Problem 2

**Let $X \sim F_{\nu_1,\nu_2}$. Show that $Y=1/X \sim F_{\nu_2,\nu_1}$.**

**Proof.** By definition, $X \sim F_{\nu_1,\nu_2}$ means
$$X = \frac{U/\nu_1}{V/\nu_2},$$
where $U\sim\chi^2_{\nu_1}$, $V\sim\chi^2_{\nu_2}$, and $U\perp V$. Then
$$Y = \frac{1}{X} = \frac{V/\nu_2}{U/\nu_1}.$$
Since $V\sim\chi^2_{\nu_2}$ and $U\sim\chi^2_{\nu_1}$ are independent, this is exactly the definition of $F_{\nu_2,\nu_1}$.
\newpage

# Problem 3

**Let $T = Z/\sqrt{W/\nu}$, where $Z\sim N(0,1)$ and $W\sim\chi^2_\nu$ are independent.**

## (a) Write down the distribution of $T$.

By definition (Eq.\ 1.4), $T\sim t_\nu$ (the Student's $t$-distribution with $\nu$ degrees of freedom).

## (b) Show $\E T = 0$ for $\nu>1$ and $\Var(T)=\nu/(\nu-2)$ for $\nu>2$.

**Given:** $\E(W^\alpha) = \dfrac{2^\alpha\,\Gamma(\nu/2+\alpha)}{\Gamma(\nu/2)}$, valid for $\nu > -2\alpha$.

**i.** Since $Z$ and $W$ are independent,
$$\E T = \E\!\left[\frac{Z}{\sqrt{W/\nu}}\right] = \E[Z]\cdot\E\!\left[\frac{\sqrt{\nu}}{\sqrt{W}}\right] = 0\cdot\E[W^{-1/2}]\cdot\sqrt{\nu}=0,$$
provided $\E[W^{-1/2}]<\infty$. Using the given formula with $\alpha=-1/2$:
$$\E[W^{-1/2}] = \frac{2^{-1/2}\,\Gamma(\nu/2 - 1/2)}{\Gamma(\nu/2)},$$
which is finite when $\nu/2 - 1/2 > 0$, i.e., $\nu > 1$.

**ii.** Since $\E T = 0$ for $\nu > 1$,
$$\Var(T) = \E[T^2] = \E\!\left[\frac{Z^2\,\nu}{W}\right] = \nu\,\E[Z^2]\,\E[W^{-1}] = \nu\,\E[W^{-1}],$$
using independence and $\E[Z^2]=1$. Applying the given formula with $\alpha=-1$:
$$\E[W^{-1}] = \frac{2^{-1}\,\Gamma(\nu/2-1)}{\Gamma(\nu/2)} = \frac{1}{2}\cdot\frac{1}{\nu/2-1} = \frac{1}{\nu-2},$$
where we used $\Gamma(\nu/2)=(\nu/2-1)\,\Gamma(\nu/2-1)$. This requires $\nu/2-1>0$, i.e., $\nu>2$.

Therefore $\Var(T) = \nu/(\nu-2)$ for $\nu > 2$.
\newpage

# Problem 4

**CLT for the Poisson distribution: if $X\sim\text{Poisson}(\lambda)$, then $(X-\lambda)/\sqrt{\lambda}\dto N(0,1)$ as $\lambda\to\infty$.**

## (a) Show $\E e^{tX^*} = e^{-t\sqrt{\lambda}-\lambda+\lambda e^{t/\sqrt{\lambda}}}$, where $X^*=(X-\lambda)/\sqrt{\lambda}$.

**Proof.** The MGF of $X\sim\text{Poisson}(\lambda)$ is $M_X(s) = \E[e^{sX}] = e^{\lambda(e^s-1)}$. Now
$$\E[e^{tX^*}] = \E\!\left[\exp\!\left(\frac{t(X-\lambda)}{\sqrt{\lambda}}\right)\right] = e^{-t\sqrt{\lambda}}\,\E\!\left[\exp\!\left(\frac{t}{\sqrt{\lambda}}\,X\right)\right] = e^{-t\sqrt{\lambda}}\,M_X\!\left(\frac{t}{\sqrt{\lambda}}\right).$$
Substituting the Poisson MGF:
$$= e^{-t\sqrt{\lambda}}\exp\!\left[\lambda\!\left(e^{t/\sqrt{\lambda}}-1\right)\right] = \exp\!\left(-t\sqrt{\lambda} - \lambda + \lambda\,e^{t/\sqrt{\lambda}}\right).
$$

## (b) Show $\E e^{tX^*}\to e^{t^2/2}$ as $\lambda\to\infty$.

**Proof.** Let $u = t/\sqrt{\lambda}\to 0$ as $\lambda\to\infty$. Taylor-expand $e^u$:
$$e^u = 1 + u + \frac{u^2}{2} + \frac{u^3}{6} + \cdots$$
Multiply by $\lambda$:
$$\lambda\,e^{t/\sqrt{\lambda}} = \lambda + t\sqrt{\lambda} + \frac{t^2}{2} + \frac{t^3}{6\sqrt{\lambda}} + \cdots$$
Therefore
$$-t\sqrt{\lambda}-\lambda+\lambda\,e^{t/\sqrt{\lambda}} = \frac{t^2}{2} + \frac{t^3}{6\sqrt{\lambda}}+\cdots \;\longrightarrow\; \frac{t^2}{2}\quad\text{as }\lambda\to\infty.$$
Hence $\E[e^{tX^*}]\to e^{t^2/2}$, which is the MGF of $N(0,1)$. By L\'evy's continuity theorem (Theorem 2.2.4), $X^*\dto N(0,1)$.
\newpage

# Problem 5

**Prove from first principles that if $X\sim\text{Binomial}(n,p)$, then $\dfrac{X-np}{\sqrt{np(1-p)}}\dto N(0,1)$.**

**Proof.** Write $X=_d Y_1+\cdots+Y_n$ where $Y_i\iid\text{Bernoulli}(p)$, with $\E[Y_i]=p$ and $\Var(Y_i)=p(1-p)=:\sigma^2$.

Define $Z_n = \dfrac{X - np}{\sigma\sqrt{n}} = \dfrac{\sum_{i=1}^n (Y_i - p)}{\sigma\sqrt{n}}$. The MGF of $Z_n$ is
$$M_{Z_n}(t) = \left[\E\exp\!\left(\frac{t(Y_1-p)}{\sigma\sqrt{n}}\right)\right]^n$$
by independence. Let $s = t/(\sigma\sqrt{n})$. Then
\begin{align*}
\E[e^{s(Y_1-p)}] &= p\,e^{s(1-p)} + (1-p)\,e^{-sp}\\
&= p\!\left(1+s(1-p)+\frac{s^2(1-p)^2}{2}+O(s^3)\right) \\
&\quad+ (1-p)\!\left(1-sp+\frac{s^2p^2}{2}+O(s^3)\right)\\
&= 1 + s\underbrace{[p(1-p)-(1-p)p]}_{=\,0} + \frac{s^2}{2}\underbrace{[p(1-p)^2+(1-p)p^2]}_{=\,p(1-p)=\sigma^2} + O(s^3)\\
&= 1 + \frac{s^2\sigma^2}{2}+O(s^3).
\end{align*}
Substituting $s^2\sigma^2 = t^2/n$:
$$\E[e^{s(Y_1-p)}] = 1 + \frac{t^2}{2n} + o\!\left(\frac{1}{n}\right).$$
Therefore
$$M_{Z_n}(t) = \left[1+\frac{t^2}{2n}+o\!\left(\frac{1}{n}\right)\right]^n \;\longrightarrow\; e^{t^2/2}\quad\text{as }n\to\infty,$$
using the standard limit $(1+a_n/n)^n\to e^{a_n}$ when $a_n\to a$. Since $e^{t^2/2}$ is the MGF of $N(0,1)$, by L\'evy's continuity theorem, $Z_n\dto N(0,1)$.
\newpage

# Problem 6

**Let $Y_1,\ldots,Y_5\iid N(0,1)$ with sample mean $\bar{Y}_5$. Let $Y_6$ be an independent $N(0,1)$ variable.**

## (a) $\bar{Y}_5$

$\bar{Y}_5 = \frac{1}{5}\sum_{i=1}^5 Y_i$. By Theorem 1.2, a linear combination of independent normals is normal:
$$\E[\bar{Y}_5]=0,\quad \Var(\bar{Y}_5)=\frac{1}{25}\sum_{i=1}^5 1 = \frac{1}{5}.$$
Therefore $\boxed{\bar{Y}_5 \sim N(0,\,1/5)}$.

## (b) $U=\sum_{i=1}^5 Y_i^2$

Each $Y_i\sim N(0,1)$, so $Y_i^2\sim\chi^2_1$. Since they are independent,
$$\boxed{U=\sum_{i=1}^5 Y_i^2 \sim \chi^2_5.}$$

## (c) $V=\sum_{i=1}^5(Y_i-\bar{Y}_5)^2$

Note $V = (n-1)S^2$ with $n=5$ and $\sigma^2=1$. By Theorem 1.4(c),
$$(n-1)S^2/\sigma^2 = V \sim \chi^2_{n-1}.$$
Therefore $\boxed{V \sim \chi^2_4}$.

## (d) $W=\sum_{i=1}^5(Y_i-\bar{Y}_5)^2 + Y_6^2$

We have $V\sim\chi^2_4$ (from part (c)) and $Y_6^2\sim\chi^2_1$. Since $V$ depends only on $Y_1,\ldots,Y_5$ and $Y_6$ is independent of them, $V$ and $Y_6^2$ are independent. By the additivity of independent chi-squared random variables,
$$\boxed{W = V + Y_6^2 \sim \chi^2_5.}$$

\newpage

# Problem 7

**The coefficient of variation is $CV=\sigma/\mu$. The sample coefficient of variation is $\widehat{CV}=S/\bar{Y}$ where $Y_i\sim N(\mu,\sigma^2)$.**

## (a) Find the distribution of $\widehat{CV}$.

By Theorem 1.4, $\bar{Y}$ and $S^2$ are independent with
$$\bar{Y}\sim N\!\left(\mu,\frac{\sigma^2}{n}\right),\qquad \frac{(n-1)S^2}{\sigma^2}\sim\chi^2_{n-1}.$$
Define $Z = \dfrac{\bar{Y}}{\sigma/\sqrt{n}} \sim N\!\left(\frac{\sqrt{n}\,\mu}{\sigma},\,1\right)$ and $V = \dfrac{(n-1)S^2}{\sigma^2}\sim\chi^2_{n-1}$, which are independent. Then
$$\frac{\sqrt{n}}{\widehat{CV}} = \frac{\sqrt{n}\,\bar{Y}}{S} = \frac{\bar{Y}/(\sigma/\sqrt{n})}{S/\sigma} = \frac{Z}{\sqrt{V/(n-1)}},$$
which by definition follows a **noncentral $t$-distribution** with $n-1$ degrees of freedom and noncentrality parameter
$$\delta = \frac{\sqrt{n}\,\mu}{\sigma} = \frac{\sqrt{n}}{CV}.$$
Therefore:
$$\boxed{\frac{\sqrt{n}}{\widehat{CV}} \sim t_{n-1}\!\left(\frac{\sqrt{n}}{CV}\right),}$$
i.e., $\widehat{CV} = \sqrt{n}\big/T$ where $T\sim t_{n-1}(\sqrt{n}/CV)$. In other words, $\widehat{CV}$ follows the distribution of a scaled reciprocal of a noncentral $t$-distributed random variable.

## (b) Find $c$ such that $P(|\widehat{CV}|\le c)=0.95$.

Since $\widehat{CV} = \sqrt{n}/T$,
$$P(|\widehat{CV}|\le c) = P\!\left(|T|\ge\frac{\sqrt{n}}{c}\right) = 0.95.$$
Equivalently, $P\!\left(|T|<\sqrt{n}/c\right)=0.05$, where $T\sim t_{n-1}(\sqrt{n}/CV)$. Denoting the 2.5th percentile of this noncentral $t$ by $t_{0.025}$ and the 97.5th percentile by $t_{0.975}$,
$$c = \frac{\sqrt{n}}{\min(|t_{0.025}|,\,t_{0.975})}.$$

\newpage

# Problem 8

**Let $X_1,\ldots,X_n\iid N(\mu,\sigma^2)$ and $S^2=\frac{1}{n-1}\sum_{i=1}^n(X_i-\bar{X})^2$.**

## (a) Identify the distribution of $S^2$.

By Theorem 1.4(c), $\frac{(n-1)S^2}{\sigma^2}\sim\chi^2_{n-1}$. Therefore
$$\boxed{S^2 \sim \frac{\sigma^2}{n-1}\,\chi^2_{n-1},}$$
i.e., $S^2$ is a scaled chi-squared. Equivalently, $S^2 \sim \text{Gamma}\!\left(\frac{n-1}{2},\,\frac{2\sigma^2}{n-1}\right)$.

## (b) Show $\sqrt{n}\!\left(\dfrac{S^2}{\sigma^2}-1\right)\dto N(0,2)$.

**Proof.** Let $V=\frac{(n-1)S^2}{\sigma^2}\sim\chi^2_{n-1}$. Write $V=\sum_{i=1}^{n-1}W_i$ where $W_i\iid\chi^2_1$ with $\E[W_i]=1$ and $\Var(W_i)=2$.

By the Lindeberg--L\'evy CLT (Theorem 1.3),
$$\frac{V-(n-1)}{\sqrt{2(n-1)}} \dto N(0,1).$$
Now
$$\sqrt{n}\!\left(\frac{S^2}{\sigma^2}-1\right) = \sqrt{n}\cdot\frac{V-(n-1)}{n-1} = \sqrt{\frac{2n}{n-1}}\cdot\frac{V-(n-1)}{\sqrt{2(n-1)}}.$$
Since $\sqrt{2n/(n-1)}\to\sqrt{2}$ as $n\to\infty$ and $\frac{V-(n-1)}{\sqrt{2(n-1)}}\dto N(0,1)$, by Slutsky's theorem (Theorem 2.5):
$$\sqrt{n}\!\left(\frac{S^2}{\sigma^2}-1\right)\dto \sqrt{2}\cdot N(0,1) = N(0,2). $$

\newpage

# Problem 9

**Verify whether each distribution belongs to the exponential family.**

## (a) $\text{Binomial}(n,p)$

The pmf is (with $n$ fixed and $p$ as the parameter):
$$f(x\mid p)=\binom{n}{x}p^x(1-p)^{n-x} = \binom{n}{x}\exp\!\left\{x\log\frac{p}{1-p}+n\log(1-p)\right\}.$$
This has the form $h(x)\exp\{c(p)\,T(x)-A(p)\}$ with:
$$h(x)=\binom{n}{x},\quad T(x)=x,\quad c(p)=\log\frac{p}{1-p},\quad A(p)=-n\log(1-p).$$
Since $c(p)$ is strictly monotone in $p\in(0,1)$, this is a **one-parameter exponential family** with natural parameter $\eta=\log\frac{p}{1-p}$.

## (b) $\text{Gamma}(\alpha,\beta)$

The pdf is ($\alpha,\beta>0$):
$$f(x\mid\alpha,\beta)=\frac{1}{\Gamma(\alpha)\,\beta^\alpha}\,x^{\alpha-1}e^{-x/\beta},\quad x>0.$$
Rewrite as:
$$f(x\mid\alpha,\beta)=\frac{1}{x}\,\mathbbm{I}(x>0)\cdot\exp\!\left\{(\alpha-1)\log x + \left(-\frac{1}{\beta}\right)x - \log\Gamma(\alpha)-\alpha\log\beta\right\}.$$
This has the form $h(x)\exp\!\left\{\sum_{j=1}^2 c_j(\boldsymbol{\theta})\,T_j(x)-A(\boldsymbol{\theta})\right\}$ with:
$$h(x)=\frac{\mathbbm{I}(x>0)}{x},\quad \mathbf{T}(x)=\begin{pmatrix}\log x\\x\end{pmatrix},\quad \mathbf{c}(\boldsymbol{\theta})=\begin{pmatrix}\alpha-1\\-1/\beta\end{pmatrix},\quad A(\boldsymbol{\theta})=\log\Gamma(\alpha)+\alpha\log\beta.$$
Since $\mathbf{c}(\alpha,\beta)$ maps $(\alpha,\beta)\in(0,\infty)^2$ onto an open subset of $\mathbb{R}^2$, this is a **full-rank two-parameter exponential family**.

\newpage

# Problem 10

## (a) Find $h(X)$ such that $h(\mu)+\tfrac{1}{2}h''(\mu)\Var(X)$ is not a good approximation of $\E[h(X)]$.

**Example.** Let $h(x)=e^x$ and $X\sim N(\mu,\sigma^2)$.

- **Exact:** $\E[e^X]=e^{\mu+\sigma^2/2}$ (by the normal MGF).
- **Approximation:** $h(\mu)+\frac{1}{2}h''(\mu)\sigma^2 = e^\mu + \frac{1}{2}e^\mu\sigma^2 = e^\mu\!\left(1+\frac{\sigma^2}{2}\right)$.

The ratio of exact to approximate is $\dfrac{e^{\sigma^2/2}}{1+\sigma^2/2}$, which diverges as $\sigma^2$ grows. For instance, with $\mu=0$ and $\sigma^2=4$:

- Exact: $e^{4/2}=e^2\approx 7.389$.
- Approximation: $1+4/2=3$.

The approximation fails because $h(x)=e^x$ has rapidly growing higher-order derivatives ($h^{(k)}(x)=e^x$ for all $k$), making the neglected higher-order terms in the Taylor expansion significant when $\sigma^2$ is not small.

## (b) Prove the expansion $\E[h(\bar{X})]=h(\mu)+\dfrac{\sigma^2}{2n}h''(\mu)+R_n$ where $R_n=O(1/n^2)$.

**Proof.** Let $X_1,\ldots,X_n$ be i.i.d.\ with mean $\mu$, variance $\sigma^2$, and finite fourth moment $\mu_4=\E[(X_1-\mu)^4]<\infty$. By Taylor's theorem with Lagrange remainder, for some $\xi$ between $\bar{X}$ and $\mu$:
$$h(\bar{X})=h(\mu)+h'(\mu)(\bar{X}-\mu)+\frac{h''(\mu)}{2}(\bar{X}-\mu)^2+\frac{h'''(\mu)}{6}(\bar{X}-\mu)^3+\frac{h^{(4)}(\xi)}{24}(\bar{X}-\mu)^4.$$
Taking expectations and using $\E[\bar{X}-\mu]=0$ and $\E[(\bar{X}-\mu)^2]=\sigma^2/n$:
$$\E[h(\bar{X})]=h(\mu)+\frac{\sigma^2}{2n}h''(\mu)+\underbrace{\frac{h'''(\mu)}{6}\E[(\bar{X}-\mu)^3]+\frac{1}{24}\E\!\left[h^{(4)}(\xi)(\bar{X}-\mu)^4\right]}_{=:\,R_n}.$$

We show $R_n=O(1/n^2)$:

1. **Third-moment term.** By independence and $\E[X_i-\mu]=0$:
$$\E[(\bar{X}-\mu)^3]=\frac{1}{n^3}\E\!\left[\left(\sum_{i=1}^n(X_i-\mu)\right)^3\right]=\frac{1}{n^3}\sum_{i=1}^n\E[(X_i-\mu)^3]=\frac{\mu_3}{n^2},$$
where $\mu_3=\E[(X_1-\mu)^3]$ is finite (since $|\mu_3|\le \mu_4^{3/4}$ by H\"older's inequality, or directly by the existence of the fourth moment). The cross terms vanish by independence and zero mean. Hence this term is $O(1/n^2)$.

2. **Fourth-moment term.** Since $|h^{(4)}(\xi)|\le M$,
$$\left|\frac{1}{24}\E[h^{(4)}(\xi)(\bar{X}-\mu)^4]\right|\le \frac{M}{24}\E[(\bar{X}-\mu)^4].$$
Expanding $(\sum(X_i-\mu))^4$ and using independence:
$$\E\!\left[\left(\sum_{i=1}^n(X_i-\mu)\right)^4\right]=n\mu_4+3n(n-1)\sigma^4.$$
Therefore
$$\E[(\bar{X}-\mu)^4]=\frac{n\mu_4+3n(n-1)\sigma^4}{n^4}=\frac{\mu_4}{n^3}+\frac{3(n-1)\sigma^4}{n^3}=O(1/n^2).$$

Combining, $R_n=O(1/n^2)$.