---
title: "P8131 Homework 4"
author: "Yongyan Liu (yl6107)"
date: "2026-03-01"
output:
  pdf_document:
    latex_engine: xelatex
header-includes:
  - \usepackage{amsmath}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(knitr)
library(nnet)
library(MASS)
```

# Housing Satisfaction in Copenhagen

```{r}
htypes <- c("Tower", "Apartment", "House")
sat_lvl <- c("Low", "Medium", "High")

housing <- data.frame(
  housing_type = factor(rep(htypes, each = 2), levels = htypes),
  contact = factor(rep(c("Low", "High"), 3), levels = c("Low", "High")),
  low_sat  = c(65, 34, 130, 141, 67, 130),
  med_sat  = c(54, 47,  76, 116, 48, 105),
  high_sat = c(100, 100, 111, 191, 62, 104)
)
housing$total <- housing$low_sat + housing$med_sat + housing$high_sat
```

---

## Problem 1

**Summarize the data using appropriate tables of percentages to show the
pair-wise associations between the levels of satisfaction and 1) contact
with other residents and 2) type of housing. Comment on patterns.**

### Satisfaction by Contact Level

```{r}
contact_tbl <- housing %>%
  group_by(contact) %>%
  summarize(Low = sum(low_sat), Medium = sum(med_sat),
            High = sum(high_sat), Total = sum(total)) %>%
  mutate(across(Low:High, ~ sprintf("%.1f%%", 100 * .x / Total),
                .names = "{.col}_pct"))

kable(contact_tbl %>% dplyr::select(contact, Low, Low_pct, Medium, Medium_pct,
                                     High, High_pct, Total),
      col.names = c("Contact", "Low", "(%)", "Medium", "(%)",
                     "High", "(%)", "Total"),
      caption = "Satisfaction by Contact Level")
```

**Pattern:** High-contact residents report higher satisfaction (40.8%
high) compared to low-contact residents (38.3% high). Low satisfaction
is more prevalent among low-contact residents (36.7%) than high-contact
(31.5%). More contact with neighbors is associated with greater
housing satisfaction.

### Satisfaction by Housing Type

```{r}
housing_tbl <- housing %>%
  group_by(housing_type) %>%
  summarize(Low = sum(low_sat), Medium = sum(med_sat),
            High = sum(high_sat), Total = sum(total)) %>%
  mutate(across(Low:High, ~ sprintf("%.1f%%", 100 * .x / Total),
                .names = "{.col}_pct"))

kable(housing_tbl %>% dplyr::select(housing_type, Low, Low_pct, Medium,
                                     Medium_pct, High, High_pct, Total),
      col.names = c("Housing Type", "Low", "(%)", "Medium", "(%)",
                     "High", "(%)", "Total"),
      caption = "Satisfaction by Housing Type")
```

**Pattern:** Tower block residents have the highest proportion of high
satisfaction (50.0%) and lowest low satisfaction (24.8%). House
residents show the most low satisfaction (38.2%) and least high
satisfaction (32.2%). Apartment residents fall in between.

---

## Problem 2

**Use nominal logistic regression to model the associations. Obtain a
model that summarizes the patterns. Describe findings (odds ratios with
95% CIs, goodness-of-fit). Is there interaction?**

We use satisfaction (Low/Medium/High) as the response with "Low" as
reference:
$$\log\frac{\pi_j}{\pi_{\text{Low}}} = \beta_{0j} + \beta_{1j}\,\text{Contact}_{\text{High}} + \beta_{2j}\,\text{Apartment} + \beta_{3j}\,\text{House}, \quad j \in \{\text{Med, High}\}$$

```{r}
# Reshape to long format
dat <- data.frame(
  housing_type = rep(housing$housing_type, 3),
  contact = rep(housing$contact, 3),
  satisfaction = factor(rep(sat_lvl, each = 6), levels = sat_lvl),
  count = c(housing$low_sat, housing$med_sat, housing$high_sat)
)
```

### Test for Interaction

```{r}
fit_full <- multinom(satisfaction ~ contact * housing_type,
                     weights = count, data = dat, trace = FALSE)
fit_main <- multinom(satisfaction ~ contact + housing_type,
                     weights = count, data = dat, trace = FALSE)
summary(fit_full)
summary(fit_main)

dev_diff <- fit_main$deviance - fit_full$deviance
df_diff <- length(coef(fit_full)) - length(coef(fit_main))
p_int <- 1 - pchisq(dev_diff, df_diff)

cat("LRT for interaction: G =", round(dev_diff, 3),
    ", df =", df_diff, ", p =", round(p_int, 3), "\n")
```

The interaction is not significant (p = `r round(p_int, 3)`), so we use
the main effects model.

```{r}
# Odds ratios with 95% CIs
coefs <- coef(fit_main)
se <- summary(fit_main)$standard.errors
OR <- exp(coefs)
OR_lo <- exp(coefs - 1.96 * se)
OR_hi <- exp(coefs + 1.96 * se)

for (j in c("Medium", "High")) {
  cat("\n--- ", j, "vs Low ---\n")
  res <- data.frame(OR = round(OR[j, ], 3),
                    CI_lower = round(OR_lo[j, ], 3),
                    CI_upper = round(OR_hi[j, ], 3))
  print(kable(res))
}
```

### Goodness-of-Fit

```{r}
patterns <- distinct(dat[, c("housing_type", "contact")])
obs <- matrix(dat$count, nrow = 6, ncol = 3)
m <- rowSums(obs)
pred <- predict(fit_main, newdata = patterns, type = "probs")
expected <- pred * m

df_gof <- 6 * 2 - 8  # 6 patterns x (3-1) - 8 parameters = 4

# Pearson chi-square
X2 <- sum((obs - expected)^2 / expected)
p_X2 <- 1 - pchisq(X2, df_gof)

# Deviance (likelihood ratio)
D <- 2 * sum(obs * log(obs / expected))
p_D <- 1 - pchisq(D, df_gof)

cat("Pearson X² =", round(X2, 3), ", df =", df_gof, ", p =", round(p_X2, 3), "\n")
cat("Deviance   =", round(D, 3), ", df =", df_gof, ", p =", round(p_D, 3), "\n")
```

### Interpretation

1. **Contact:** High contact significantly increases odds of both medium
   (OR = `r round(OR["Medium","contactHigh"], 2)`, 95% CI:
   `r round(OR_lo["Medium","contactHigh"], 2)`--`r round(OR_hi["Medium","contactHigh"], 2)`)
   and high satisfaction
   (OR = `r round(OR["High","contactHigh"], 2)`, 95% CI:
   `r round(OR_lo["High","contactHigh"], 2)`--`r round(OR_hi["High","contactHigh"], 2)`)
   vs. low satisfaction.

2. **Housing type:** Compared to tower blocks, apartment residents have
   lower odds of high satisfaction
   (OR = `r round(OR["High","housing_typeApartment"], 2)`, 95% CI:
   `r round(OR_lo["High","housing_typeApartment"], 2)`--`r round(OR_hi["High","housing_typeApartment"], 2)`),
   and house residents have even lower odds
   (OR = `r round(OR["High","housing_typeHouse"], 2)`, 95% CI:
   `r round(OR_lo["High","housing_typeHouse"], 2)`--`r round(OR_hi["High","housing_typeHouse"], 2)`).

3. **Goodness-of-fit:** Pearson $\chi^2$ = `r round(X2, 2)` (p =
   `r round(p_X2, 3)`) and Deviance = `r round(D, 2)` (p =
   `r round(p_D, 3)`) with `r df_gof` df, both indicating adequate fit.

4. **No significant interaction** (p = `r round(p_int, 3)`).

---

## Problem 3

**Fit a proportional odds model with the same variables. What does it
tell?**

The proportional odds model:
$$\log\frac{P(Y \leq j)}{P(Y > j)} = \beta_{0j} - \mathbf{x}^T\boldsymbol{\beta}, \quad j = 1, 2$$

```{r}
dat$satisfaction_ord <- ordered(dat$satisfaction, levels = sat_lvl)
fit_po <- polr(satisfaction_ord ~ contact + housing_type,
               weights = count, data = dat)
summary(fit_po)
```

```{r}
# Cumulative ORs for higher satisfaction
# polr: logit(P(Y<=j)) = beta_0j - x'beta, so exp(beta) = OR for Y > j
b <- coef(fit_po)
se_po <- sqrt(diag(vcov(fit_po)))[1:length(b)]

OR_po <- exp(b)
CI_lo_po <- exp(b - 1.96 * se_po)
CI_hi_po <- exp(b + 1.96 * se_po)

po_res <- data.frame(
  Coefficient = round(b, 4), SE = round(se_po, 4),
  OR = round(OR_po, 3),
  CI_lower = round(CI_lo_po, 3),
  CI_upper = round(CI_hi_po, 3)
)
kable(po_res, col.names = c("Coef", "SE", "OR (higher sat.)",
                             "95% CI Lower", "95% CI Upper"),
      caption = "Proportional Odds Model: Cumulative ORs")
```

```{r}
# Goodness-of-fit
pred_po <- predict(fit_po, newdata = patterns, type = "probs")
exp_po <- pred_po * m
df_po <- 6 * 2 - 5  # 12 - 5 parameters = 7

X2_po <- sum((obs - exp_po)^2 / exp_po)
p_X2_po <- 1 - pchisq(X2_po, df_po)

D_po <- 2 * sum(obs * log(obs / exp_po))
p_D_po <- 1 - pchisq(D_po, df_po)

cat("Pearson X² =", round(X2_po, 3), ", df =", df_po, ", p =", round(p_X2_po, 3), "\n")
cat("Deviance   =", round(D_po, 3), ", df =", df_po, ", p =", round(p_D_po, 3), "\n")
```

### Interpretation

1. **Contact:** High contact increases odds of higher satisfaction by a
   factor of `r round(OR_po["contactHigh"], 2)` (95% CI:
   `r round(CI_lo_po["contactHigh"], 2)`--`r round(CI_hi_po["contactHigh"], 2)`).

2. **Housing type:** Apartment residents have lower odds of higher
   satisfaction vs. tower blocks
   (OR = `r round(OR_po["housing_typeApartment"], 2)`, 95% CI:
   `r round(CI_lo_po["housing_typeApartment"], 2)`--`r round(CI_hi_po["housing_typeApartment"], 2)`).
   House residents have even lower odds
   (OR = `r round(OR_po["housing_typeHouse"], 2)`, 95% CI:
   `r round(CI_lo_po["housing_typeHouse"], 2)`--`r round(CI_hi_po["housing_typeHouse"], 2)`).

3. **Model fit:** Pearson $\chi^2$ = `r round(X2_po, 2)` (p =
   `r round(p_X2_po, 3)`) and Deviance = `r round(D_po, 2)` (p =
   `r round(p_D_po, 3)`) with `r df_po` df, both indicating adequate fit.

4. **Parsimony:** The proportional odds model uses 5 parameters vs. 8
   in the nominal model, giving a more concise summary while fitting
   the data well.

---

## Problem 4

**Calculate Pearson residuals from the proportional odds model. Where
are the largest discrepancies?**

```{r}
resid_po <- (obs - exp_po) / sqrt(exp_po)

rownames(resid_po) <- paste(rep(htypes, each = 2),
                            rep(c("Low", "High"), 3), sep = "-")
colnames(resid_po) <- c("Low Sat", "Med Sat", "High Sat")

kable(round(resid_po, 3), caption = "Pearson Residuals")
```

```{r}
cat("Range of residuals:", round(range(resid_po), 3), "\n")
cat("Largest |residual|:", round(max(abs(resid_po)), 3), "\n")
```

### Interpretation

All Pearson residuals are relatively small ($|r| < 1.5$), confirming
adequate model fit. The largest discrepancies involve house residents:
the model over-predicts high satisfaction for high-contact house
residents (r = -1.48) and under-predicts medium satisfaction for the
same group (r = 1.37). This suggests minor heterogeneity in the
contact-by-housing-type pattern that the main effects model does not
fully capture.
