---
title: "Data Science II Homework 2"
author: "Yongyan Liu (yl6107)"
date: "2026-03-06"
output:
  pdf_document:
    latex_engine: xelatex
header-includes:
- \usepackage{fancyhdr}
- \usepackage{lipsum}
- \pagestyle{fancy}
- \fancyhead[R]{\thepage}
- \fancypagestyle{plain}{\pagestyle{fancy}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

Load the dataset and split the dataset for training and testing.

```{r}
library(tidyverse)
library(rsample)
library(caret)
library(earth)
library(mgcv)
library(pdp)

# Load data (first column is college name, use as row names)
college <- read.csv("College.csv", row.names = 1)

# Train/test split (80/20)
set.seed(2026)
data_split <- initial_split(college, prop = 0.8)
train_data <- training(data_split)
test_data <- testing(data_split)
```

# (a) Smoothing Splines

```{r, fig.height=6, fig.width=8}
# Range of degrees of freedom to explore
df_values <- c(2, 3, 5, 8, 10, 15)

# Set up plot
x_grid <- seq(min(train_data$Top10perc), max(train_data$Top10perc), length = 200)

par(mfrow = c(2, 3), mar = c(4, 4, 2, 1))
for (df_val in df_values) {
  fit_ss <- smooth.spline(train_data$Top10perc, train_data$Outstate, df = df_val)
  pred_ss <- predict(fit_ss, x = x_grid)

  plot(train_data$Top10perc, train_data$Outstate,
       col = rgb(0.5, 0.5, 0.5, 0.4), pch = 19, cex = 0.6,
       xlab = "Top10perc", ylab = "Outstate",
       main = paste("df =", df_val))
  lines(pred_ss, col = "blue", lwd = 2)
}
```

As df increases, the curve becomes more flexible: df = 2 is nearly linear, df = 3--5 captures the increasing trend with moderate curvature, and df = 8--15 overfits with unnecessary wiggles. We select the optimal df using GCV.

```{r, fig.height=4, fig.width=6}
# Fit with GCV (automatic df selection)
fit_opt <- smooth.spline(train_data$Top10perc, train_data$Outstate)
optimal_df <- fit_opt$df

# Plot the optimal fit
pred_opt <- predict(fit_opt, x = x_grid)

plot(train_data$Top10perc, train_data$Outstate,
     col = rgb(0.5, 0.5, 0.5, 0.4), pch = 19, cex = 0.6,
     xlab = "Top10perc", ylab = "Outstate",
     main = paste0("Optimal Smoothing Spline (GCV, df = ", round(optimal_df, 2), ")"))
lines(pred_opt, col = "red", lwd = 2)
```

The GCV-selected degrees of freedom is **`r round(optimal_df, 2)`**, which balances bias and variance automatically.

# (b) MARS Model

```{r, fig.height=4, fig.width=6}
ctrl <- trainControl(method = "cv", number = 10, selectionFunction = "oneSE")

# Tune MARS: degree (1 = additive, 2 = two-way interactions) and nprune
mars_grid <- expand.grid(degree = 1:3,
                         nprune = 2:15)

set.seed(2026)
mars_fit <- train(Outstate ~ .,
                  data = train_data,
                  method = "earth",
                  tuneGrid = mars_grid,
                  trControl = ctrl)

ggplot(mars_fit) + theme_bw()
mars_fit$bestTune
```

```{r}
# Final model coefficients (regression function)
coef(mars_fit$finalModel)
# Show the regression function
summary(mars_fit$finalModel)
```

Using the one-standard-error rule, the selected tuning parameters are degree = `r mars_fit$bestTune$degree` and nprune = `r mars_fit$bestTune$nprune`, which gives the simplest model whose CV error is within one SE of the minimum.

```{r, fig.height=4, fig.width=6}
# Partial dependence plot for Expend
p1 <- pdp::partial(mars_fit, pred.var = "Expend", grid.resolution = 50) %>%
  autoplot(train = train_data, rug = TRUE) +
  ggtitle("PDP: Expend") +
  theme_bw()
p1
```

The PDP for `Expend` shows tuition increases sharply with expenditure up to a point, then levels off.

```{r}
# Test error
mars_pred <- predict(mars_fit, newdata = test_data)
mars_mse <- mean((mars_pred - test_data$Outstate)^2)
cat("MARS Test RMSE:", round(sqrt(mars_mse), 2), "\n")
```

# (c) Generalized Additive Model (GAM)

```{r, fig.height=8, fig.width=10}
# Fit GAM with smooth terms for all predictors
# select = TRUE adds extra shrinkage penalty to drop unimportant terms
gam_fit <- gam(Outstate ~ s(Apps) + s(Accept) + s(Enroll) + s(Top10perc) +
                 s(Top25perc) + s(F.Undergrad) + s(P.Undergrad) +
                 s(Room.Board) + s(Books) + s(Personal) + s(PhD) +
                 s(Terminal) + s(S.F.Ratio) + s(perc.alumni) +
                 s(Expend) + s(Grad.Rate),
               data = train_data, select = TRUE)

summary(gam_fit)
```

```{r, fig.height=10, fig.width=10}
# Identify significant nonlinear terms (edf > 1 and p-value < 0.05)
gam_summary <- summary(gam_fit)
edf <- gam_summary$s.table[, "edf"]
pval <- gam_summary$s.table[, "p-value"]
nonlinear_idx <- which(edf > 1 & pval < 0.05)
nonlinear_names <- names(edf)[nonlinear_idx]

# Plot only the nonlinear smooth terms
par(mfrow = c(ceiling(length(nonlinear_idx) / 3), 3), mar = c(4, 4, 2, 1))
for (idx in nonlinear_idx) {
  plot(gam_fit, shade = TRUE, shade.col = "lightskyblue",
       seWithMean = TRUE, select = idx)
}
```

The plots above show the significant nonlinear smooth terms (edf > 1 and p-value < 0.05). Terms with edf close to 1 (essentially linear) or non-significant p-values are not shown.

```{r}
# Test error
gam_pred <- predict(gam_fit, newdata = test_data)
gam_mse <- mean((gam_pred - test_data$Outstate)^2)
cat("GAM Test RMSE:", round(sqrt(gam_mse), 2), "\n")
```

# (d) MARS vs. Linear Model for This Dataset

```{r}
# Fit a linear model for comparison
set.seed(2026)
lm_fit <- train(Outstate ~ .,
                data = train_data,
                method = "lm",
                trControl = ctrl)

# Compare via resampling
resamp <- resamples(list(MARS = mars_fit, Linear = lm_fit))
summary(resamp)
bwplot(resamp, metric = "RMSE")

# Test errors
lm_pred <- predict(lm_fit, newdata = test_data)
lm_mse <- mean((lm_pred - test_data$Outstate)^2)

cat("Linear Model Test RMSE:", round(sqrt(lm_mse), 2), "\n")
cat("MARS Test RMSE:", round(sqrt(mars_mse), 2), "\n")
```

**Yes, we favor MARS over the linear model.** MARS achieves a lower test RMSE and captures the nonlinear relationships (e.g., `Expend`) that a linear model cannot. It also performs automatic variable selection while remaining interpretable.

# (e) MARS vs. Linear Model in Practice

**Neither is universally superior.** MARS is preferred when nonlinearity is present and the goal is prediction, as it captures threshold effects and performs automatic variable selection. A linear model is preferred when the goal is inference, the sample size is small, or the true relationship is approximately linear. In practice, one should compare both via cross-validation and choose based on the prediction--interpretability trade-off.
