---
title: "P8131 Homework 6"
author: "Yongyan Liu (yl6107)"
date: "2026-04-06"
output:
  pdf_document:
    latex_engine: xelatex
header-includes:
  - \usepackage{amsmath}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(nlme)
```

# Problem 1

**Suppose we have the following statistical model:
$Y_{ij} = \mu + b_i + e_{ij}$, where $i = 1, \ldots, m$ and
$j = 1, \ldots, n$, $b_i \sim N(0, \sigma_b^2)$,
$e_{ij} \sim N(0, \sigma_e^2)$, $b_i$ and $e_{ij}$ are statistically
independent for each $i$ and $j$, and $e_{ij}$ and $e_{ik}$ are
statistically independent for any two values $j, k = 1, \ldots, n$ and
$j \neq k$. Find the variance of $Y_{ij}$ and the covariance and
correlation between any two values $Y_{ij}$ and $Y_{ik}$, $j \neq k$.
Which one of the popular covariance patterns does this correspond to?**

## Variance of $Y_{ij}$

Since $\mu$ is a constant, it contributes zero variance. By the
independence of $b_i$ and $e_{ij}$:

$$\text{Var}(Y_{ij}) = \text{Var}(\mu + b_i + e_{ij}) = \text{Var}(b_i + e_{ij}) = \text{Var}(b_i) + \text{Var}(e_{ij}) = \sigma_b^2 + \sigma_e^2.$$

## Covariance between $Y_{ij}$ and $Y_{ik}$, $j \neq k$

$$\text{Cov}(Y_{ij}, Y_{ik}) = \text{Cov}(b_i + e_{ij},\; b_i + e_{ik}).$$
$$= \text{Cov}(b_i, b_i) + \text{Cov}(b_i, e_{ik}) + \text{Cov}(e_{ij}, b_i) + \text{Cov}(e_{ij}, e_{ik}).$$

By mutual independence: $\text{Cov}(b_i, e_{ik}) = 0$,
$\text{Cov}(e_{ij}, b_i) = 0$, and $\text{Cov}(e_{ij}, e_{ik}) = 0$
(since $j \neq k$). Therefore:

$$\text{Cov}(Y_{ij}, Y_{ik}) = \text{Var}(b_i) = \sigma_b^2.$$

## Correlation between $Y_{ij}$ and $Y_{ik}$, $j \neq k$

Since $\text{Var}(Y_{ij}) = \text{Var}(Y_{ik}) = \sigma_b^2 + \sigma_e^2$
for all $j$:

$$\text{Corr}(Y_{ij}, Y_{ik}) = \frac{\text{Cov}(Y_{ij}, Y_{ik})}{\sqrt{\text{Var}(Y_{ij})\,\text{Var}(Y_{ik})}} = \frac{\sigma_b^2}{\sigma_b^2 + \sigma_e^2}.$$

## Covariance pattern

The correlation $\rho = \sigma_b^2 / (\sigma_b^2 + \sigma_e^2)$ is
constant for all pairs $j \neq k$. The covariance matrix has the form:

$$\text{Cov}(\mathbf{Y}_i) = (\sigma_b^2 + \sigma_e^2)\begin{pmatrix} 1 & \rho & \cdots & \rho \\ \rho & 1 & \cdots & \rho \\ \vdots & \vdots & \ddots & \vdots \\ \rho & \rho & \cdots & 1 \end{pmatrix}.$$

This is the **compound symmetry** (also called exchangeable) covariance
pattern.

# Problem 2

**A study was conducted involving 27 children, 16 boys and 11 girls. On
each child, the distance (mm) from the center of the pituitary to the
pterygomaxillary fissure was made at ages 8, 10, 12, and 14. The goal
was to study how distance is affected by age and gender.**

```{r}
dental <- read.table("HW6-dental.txt", header = TRUE)
dental$Gender <- factor(dental$Gender, levels = c(0, 1),
                        labels = c("Girl", "Boy"))
```

## (a) Spaghetti plot

```{r, fig.width=6, fig.height=4}
ggplot(dental, aes(x = Age, y = Distance,
                   group = Child, color = Gender)) +
  geom_line() +
  geom_point(size = 1) +
  labs(x = "Age (years)", y = "Distance (mm)",
       title = "Spaghetti Plot of Dental Distance by Age and Gender",
       color = "Gender") +
  theme_bw()
```

Distance generally increases with age for both boys and girls. Boys
tend to have slightly larger distances and somewhat steeper growth
trajectories than girls. There is considerable between-subject
variability within each gender group.

## (b) Marginal form of the mixed model

**Assume the model:**
$$Y_{ij} = \beta_0 + a_i + b_0 \cdot I(\text{sex}_i = 0) + b_1 \cdot I(\text{sex}_i = 1) + \beta_1 \cdot \text{age}_{ij} + e_{ij},$$
**where $a_i \sim N(0, \sigma_a^2)$, $b_k \sim N(0, \sigma_b^2)$
($k = 0, 1$) are random coefficients, and
$e_{ij} \sim N(0, \sigma_e^2)$; all mutually independent.**

### Marginal mean

Taking the expectation of both sides:

$$E(Y_{ij}) = \beta_0 + E(a_i) + E(b_0) \cdot I(\text{sex}_i = 0) + E(b_1) \cdot I(\text{sex}_i = 1) + \beta_1 \cdot \text{age}_{ij} + E(e_{ij}).$$

Since $a_i$, $b_0$, $b_1$, and $e_{ij}$ are all zero-mean random
variables:

$$E(Y_{ij}) = \beta_0 + \beta_1 \cdot \text{age}_{ij}.$$

### Marginal variance

For a given subject $i$, exactly one of $I(\text{sex}_i = 0)$ or
$I(\text{sex}_i = 1)$ equals 1. The random components contributing to
$Y_{ij}$ are $a_i$, one of $b_0$ or $b_1$ (depending on sex), and
$e_{ij}$. By mutual independence:

- If $\text{sex}_i = 0$:
  $\text{Var}(Y_{ij}) = \text{Var}(a_i) + \text{Var}(b_0) + \text{Var}(e_{ij}) = \sigma_a^2 + \sigma_b^2 + \sigma_e^2.$

- If $\text{sex}_i = 1$:
  $\text{Var}(Y_{ij}) = \text{Var}(a_i) + \text{Var}(b_1) + \text{Var}(e_{ij}) = \sigma_a^2 + \sigma_b^2 + \sigma_e^2.$

Since $b_0$ and $b_1$ share the same variance $\sigma_b^2$:

$$\text{Var}(Y_{ij}) = \sigma_a^2 + \sigma_b^2 + \sigma_e^2$$

for all subjects regardless of gender.

### Marginal covariance

For two observations $j \neq k$ on the same subject $i$, stripping
the constant terms (which do not contribute to covariance) and noting
that for a given subject exactly one of $b_0$ or $b_1$ applies
(denote it $b_{\text{sex}_i}$):

$$\text{Cov}(Y_{ij}, Y_{ik}) = \text{Cov}(a_i + b_{\text{sex}_i} + e_{ij},\; a_i + b_{\text{sex}_i} + e_{ik}).$$

By mutual independence of $a_i$, $b_{\text{sex}_i}$, $e_{ij}$, and
$e_{ik}$: all cross-terms vanish. Therefore:

$$\text{Cov}(Y_{ij}, Y_{ik}) = \text{Var}(a_i) + \text{Var}(b_{\text{sex}_i}) = \sigma_a^2 + \sigma_b^2, \quad j \neq k.$$

The marginal covariance matrix is therefore:

$$\text{Cov}(\mathbf{Y}_i) = \begin{pmatrix} \sigma_a^2 + \sigma_b^2 + \sigma_e^2 & \sigma_a^2 + \sigma_b^2 & \sigma_a^2 + \sigma_b^2 & \sigma_a^2 + \sigma_b^2 \\ \sigma_a^2 + \sigma_b^2 & \sigma_a^2 + \sigma_b^2 + \sigma_e^2 & \sigma_a^2 + \sigma_b^2 & \sigma_a^2 + \sigma_b^2 \\ \sigma_a^2 + \sigma_b^2 & \sigma_a^2 + \sigma_b^2 & \sigma_a^2 + \sigma_b^2 + \sigma_e^2 & \sigma_a^2 + \sigma_b^2 \\ \sigma_a^2 + \sigma_b^2 & \sigma_a^2 + \sigma_b^2 & \sigma_a^2 + \sigma_b^2 & \sigma_a^2 + \sigma_b^2 + \sigma_e^2 \end{pmatrix},$$

which is a compound symmetry structure with correlation
$\rho = (\sigma_a^2 + \sigma_b^2) / (\sigma_a^2 + \sigma_b^2 + \sigma_e^2)$.

## (c) Fit marginal models with three covariance structures

**Marginal model:**
$E(Y_{ij}) = \beta_0 + \beta_1 \cdot \text{sex}_i + \beta_2 \cdot \text{age}_{ij}$,
$\text{Var}(\mathbf{Y}_i) = \Sigma$.

```{r}
# Recode Gender as numeric for the marginal model
dental$sex <- ifelse(dental$Gender == "Boy", 1, 0)
```

### (i) Compound symmetry

```{r}
fit_cs <- gls(Distance ~ sex + Age,
              correlation = corCompSymm(form = ~ 1 | Child),
              data = dental, method = "REML")
summary(fit_cs)
```

### (ii) Exponential

```{r}
fit_exp <- gls(Distance ~ sex + Age,
               correlation = corExp(form = ~ 1 | Child),
               data = dental, method = "REML")
summary(fit_exp)
```

### (iii) Autoregressive AR(1)

```{r}
fit_ar1 <- gls(Distance ~ sex + Age,
               correlation = corAR1(form = ~ 1 | Child),
               data = dental, method = "REML")
summary(fit_ar1)
```

### Comparison of coefficient estimates

```{r}
coef_table <- data.frame(
  CS   = round(coef(fit_cs), 4),
  Exp  = round(coef(fit_exp), 4),
  AR1  = round(coef(fit_ar1), 4)
)
rownames(coef_table) <- c("Intercept (beta_0)", "Sex (beta_1)",
                           "Age (beta_2)")
knitr::kable(coef_table, caption = "Fixed effect estimates across models")
```

The fixed effect estimates are very similar across all three covariance
structures.

### Comparison of covariance estimates

```{r}
# Compound symmetry
cat("=== Compound Symmetry ===\n")
cat("Residual sigma:", fit_cs$sigma, "\n")
cat("Rho:", coef(fit_cs$modelStruct$corStruct, unconstrained = FALSE),
    "\n")
cs_cor <- corMatrix(fit_cs$modelStruct$corStruct)[[1]]
cat("Correlation matrix:\n")
print(round(cs_cor, 4))

cat("\n=== Exponential ===\n")
cat("Residual sigma:", fit_exp$sigma, "\n")
cat("Range:", coef(fit_exp$modelStruct$corStruct, unconstrained = FALSE),
    "\n")
exp_cor <- corMatrix(fit_exp$modelStruct$corStruct)[[1]]
cat("Correlation matrix:\n")
print(round(exp_cor, 4))

cat("\n=== AR(1) ===\n")
cat("Residual sigma:", fit_ar1$sigma, "\n")
cat("Phi:", coef(fit_ar1$modelStruct$corStruct, unconstrained = FALSE),
    "\n")
ar1_cor <- corMatrix(fit_ar1$modelStruct$corStruct)[[1]]
cat("Correlation matrix:\n")
print(round(ar1_cor, 4))
```

**Summary of covariance parameters:**

For this dataset, the time points are equally spaced (ages 8, 10, 12,
14 with spacing 2), so the AR(1) and exponential models can produce identical correlation structures. In contrast,
compound symmetry assumes the same correlation regardless of time
separation, which may be overly restrictive if correlations decay with
increasing time lag.
