---
title: "P8139 Final Project Proposal: methylation-driven transcriptional silencing in TCGA-KIRC"
author: "Yongyan Liu (yl6107) — Homework #5"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: true
geometry: margin=0.9in
fontsize: 11pt
linestretch: 1.1
header-includes:
  - \usepackage{booktabs}
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup{font=small,labelfont=bf,skip=4pt}
  - \setlength{\parskip}{0.35em}
  - \setlength{\parindent}{0pt}
  - \setlength{\textfloatsep}{8pt}
  - \setlength{\intextsep}{8pt}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(knitr)
library(kableExtra)
```

# Cancer type and research question

**Cancer type.** Clear-cell renal cell carcinoma (ccRCC), TCGA project KIRC. ccRCC is chosen because it is the largest paired-design cohort in TCGA, its molecular biology is dominated by VHL inactivation (a well-characterized positive control), and its single dominant histology limits biological heterogeneity (TCGA Research Network 2013; Kaelin 2007).

**Motivating problem.** Paired tumor–adjacent-normal methylation analyses on the Illumina 450K platform routinely declare more than half of tested CpGs differentially methylated at FDR $<0.05$. The list is statistically valid but biologically uninformative because most differentially methylated CpGs are passenger or cell-composition changes, not regulators of transcription.

**Research question.** *Which CpG methylation changes drive transcriptional silencing of their cognate gene?* This is the epigenetic counterpart of Knudson's tumor-suppressor second hit and the operationally relevant question for cancer epigenetics. We address it by integrating paired DNA methylation with paired RNA-seq from the same patients, requiring concordant evidence from two statistical tests before declaring a CpG transcriptionally silencing.

# Data sources, samples, and study design

**Data acquisition.** DNA methylation 450K $\beta$-values, RNA-seq STAR-Counts (used for $\log_2(\mathrm{TPM}+1)$), and harmonized clinical metadata are retrieved from the GDC with `TCGAbiolinks::GDCquery` and `GDCdownload` (Bioconductor 3.22; Colaprico et al. 2016).

**Samples and design.** This is a **paired case–control design** at the methylation level (each patient contributes a primary-tumor `-01A` aliquot and a solid adjacent-normal `-11A` aliquot) combined with an **independent case-mostly design** at the integration level (most patients have only the tumor RNA-seq). One `-05A` "additional new primary" is excluded by pre-registration. Sample counts:

```{r samples}
samp <- data.frame(
  Modality = c("DNA methylation 450K (aliquots)",
               "DNA methylation 450K (patients)",
               "Paired methylation set (both -01A and -11A)",
               "RNA-seq STAR-Counts (patient × sample-type keys)",
               "Integration set: matched (patient, sample-type) keys"),
  N = c("484 (324 tumor + 160 normal)",
        "319",
        "160 patients",
        "605",
        "342 (318 tumor + 24 adjacent-normal)"),
  check.names = FALSE)
kable(samp, format = "latex", booktabs = TRUE, align = c("l","r"),
      caption = "TCGA-KIRC cohort sizes after pre-registered exclusions.",
      linesep = "", longtable = FALSE) |>
  kable_styling(latex_options = "HOLD_position", position = "center")
```

The DM test runs on the 160 paired patients; the methylation–expression integration test runs on the 342-key matched set.

# DNA methylation preprocessing steps

The 485,577-probe $\beta$-value matrix is filtered sequentially, with attrition recorded. Preprocessing is standard and conservative; no probes are added back at any later stage.

```{r preprocess}
attrition <- data.frame(
  Step = c("Baseline (450K array)",
           "Intersect with Illumina 450K annotation (drop unmapped)",
           "Drop chr X / chr Y (avoid sex-driven PC2 in EDA)",
           "Cross-reactive + SNP-overlapping probes (DMRcate; Chen 2013, Peters 2015)",
           "Drop probes with > 20% sample missingness",
           "Complete-case across all 484 samples (paired-test ready)"),
  `Probes retained` = format(c(485577, 472651, 461295, 433595, 395325, 320953),
                              big.mark = ","),
  check.names = FALSE)
kable(attrition, format = "latex", booktabs = TRUE, align = c("l","r"),
      caption = "Preprocessing attrition for DNA methylation.",
      linesep = "", longtable = FALSE) |>
  kable_styling(latex_options = "HOLD_position", position = "center")
```

$\beta$-values are clipped to $[0.001, 0.999]$ and transformed to $M = \log_2(\beta/(1-\beta))$ for variance stabilization in the downstream paired test (Du et al. 2010). RNA-seq counts are converted to TPM and log-transformed as $\log_2(\mathrm{TPM}+1)$; replicate aliquots within (patient, sample-type) are mean-aggregated.

# Analysis plan

The plan combines two independent statistical tests on the same patients; **a CpG is "silencing" only if both tests agree with concordant direction**. This operationalizes Knudson's tumor-suppressor second hit in the methylation–expression coupling regime: differential methylation alone cannot distinguish promoter silencing from gene-body activation, and methylation–expression coupling alone cannot distinguish a tumor-acquired event from a constitutive one.

**(1) Differential methylation (DM) test — paired moderated $t$.** For each CpG, within-pair M-value differences $D_i = M_i^{\text{tumor}} - M_i^{\text{normal}}$ ($i = 1, \dots, 160$) are tested against $H_0\!:\,\mathbb{E}[D] = 0$ using `limma::lmFit` with an intercept-only design and `eBayes` shrinkage (Smyth 2004; Ritchie et al. 2015). For balanced paired data this is equivalent to the moderated paired $t$-test, and patient-level confounders cancel in $D_i$.

**(2) Methylation–expression association (MEA) test — partial-correlation regression.** For each promoter (CpG, gene) pair (annotated as `TSS1500`, `TSS200`, `5'UTR`, or `1stExon`), we fit
$$
\log_2(\mathrm{TPM}_{\text{gene}} + 1) \;=\; \alpha + \beta\,\beta_{\text{CpG}} + \gamma^{\!\top} Z + \varepsilon,
\quad Z = (\text{sample\_type, age, sex, tissue source site}),
$$
across the 342 matched samples. Adjusting for `sample_type` is essential: without it, cross-sample correlation is dominated by the bulk tumor/normal mean shift (e.g., for VHL the cross-cohort $\rho = -0.03$ vs within-status partial $p = 2 \times 10^{-6}$). The test is computed as a partial correlation after QR residualization, vectorizing to all 137,451 promoter pairs.

**(3) Silencing definition.** A (CpG, gene) pair is **silencing** if (i) DM FDR $<0.05$, (ii) MEA FDR $<0.05$, and (iii) $\widehat{\mathrm{logFC}}_{\text{DM}} > 0$ and $\hat{\beta}_{\text{MEA}} < 0$. The reverse direction (hypomethylation $\cap$ positive coupling) is recorded as **activation** and reported descriptively.

**(4) Pattern partition by responder fraction.** Mean-based DM dilutes signals present in only a minority of tumors — VHL, the dominant ccRCC driver, is methylation-silenced in only ~15% of tumors and is therefore invisible to cohort-mean tests. Each silencing CpG is annotated by $\hat{\pi}_{\text{resp}}$, the fraction of tumors whose $\beta$ deviates from the adjacent-normal mean by $>0.2$ (the standard 450K biological-significance threshold). Silencing pairs split into **subgroup-driven** ($\hat{\pi}_{\text{resp}} < 0.3$) and **cohort-wide** ($\geq 0.3$).

**(5) Pathway enrichment and cell-composition control — central methodological claim.** Bulk tumor DM is confounded by cell composition: tumor tissue carries more infiltrating leukocytes than adjacent normal kidney, so any CpG distinguishing immune from epithelial cells appears differentially methylated for compositional reasons alone. The standard fix is reference-based deconvolution (EpiDISH, CIBERSORT). Our hypothesis is that silencing's directional concordance criterion automatically excises this signal: an immune CpG bulk-hypomethylated in tumor *by* infiltration also carries elevated bulk mRNA from those same cells — the opposite of silencing's hypermethylation + low-mRNA signature, so it fails the concordance test. The falsifiable prediction is that immune pathways dominant in raw-DM enrichment disappear under the silencing restriction. We compare GSEA on raw-DM rankings (`fgsea`) against ORA on the silencing gene list (hypergeometric on Hallmark + KEGG Legacy).

**(6) Pre-registered validation.** (a) An 18-gene literature TSG list (VHL, RASSF1, SFRP1/2/4/5, DKK3, GATA5, BNC1, COL14A1, PCDH17, SLIT2, CDKN2A, APAF1, TIMP3, NEFH, UCHL1, PITX2) is the concurrent-validity sanity check. (b) **VHL positive control**: at least one VHL-promoter CpG must flag as silencing. (c) **Calibration**: the DM test's $\lambda_{GC}$ is paired with a sign-flipping permutation null on 30,000 probes × 20 replicates. (d) **External replication**: the identical pipeline is run on TCGA-KIRP (papillary RCC, 45 paired patients) and silencing-gene overlap with KIRC is tested for hypergeometric enrichment. (e) **Threshold sensitivity**: silencing counts and ORA top-hits are reported across a $3\times3$ (DM FDR, MEA FDR) $\in \{0.01, 0.05, 0.10\}$ grid.

**(7) Multiple-testing correction.** BH FDR at $q<0.05$ is applied separately to each test; the silencing AND of two independently FDR-controlled sets is conservative relative to either single-test FDR. BH FDR is also used across gene-set families. Calibration is anchored by the permutation null in (6c) rather than by Bonferroni, which is too conservative given the pervasive real signal in this cohort.

# Figures, tables, and feasibility

**Planned outputs.** Six figures: (1) PCA on top-variance CpGs (PC1 $\approx$ tumor/normal, PC2 $\approx$ sex); (2) DM diagnostic panel (p-value histogram with permutation-null overlay, paired-$\Delta\beta$ histogram, $\Delta\beta$-vs-p volcano with anchor TSGs labelled); (3) MEA volcano + responder-fraction histogram; (4) six-locus methylation–expression scatter; (5) VHL/`cg13672843` subgroup-driven case study; (6) raw-DM GSEA vs silencing-restricted ORA. Four tables: preprocessing attrition, immune-pathway dropout, FDR sensitivity grid, 18-gene TSG recovery.

**Feasibility (pipeline implemented).** A first end-to-end run of the pipeline produces the expected qualitative pattern: raw DM enrichment is dominated by immune pathways (Inflammatory Response, Complement, Interferon $\gamma$/$\alpha$, TNF$\alpha$/NF-$\kappa$B), all of which fall out of significance under the silencing restriction, leaving recognizable cancer biology (EMT, TGF-$\beta$, KRAS, Pathways in Cancer). The VHL positive control flags as silencing, and the 18-gene literature TSG panel is enriched over baseline. External replication on TCGA-KIRP recovers a substantial silencing-gene overlap with KIRC. Quantitative results, threshold-sensitivity, and calibration are deferred to the final report.

**Expected conclusion.** Silencing — defined as the directional intersection of paired-design DM and within-status methylation–expression association — should be sufficient to identify transcriptionally silencing CpGs without an external cell-composition deconvolution panel. If confirmed, the immune-pathway confounding dominating raw DM collapses under the silencing restriction while cancer biology survives, and a complementary descriptive observation — that most methylation silencing is patient-subgroup-specific rather than cohort-wide — argues for routinely reporting per-tumor responder fractions in cancer methylation studies.

# References {-}

\footnotesize \setlength{\parskip}{0pt} \setlength{\parindent}{0pt}

\noindent
Colaprico, A., et al. (2016). TCGAbiolinks: integrative analysis of TCGA data. *Nucleic Acids Research* **44**, e71. \
Du, P., et al. (2010). Comparison of Beta-value and M-value methods for quantifying methylation. *BMC Bioinformatics* **11**, 587. \
Jones, P. A., Baylin, S. B. (2007). The epigenomics of cancer. *Cell* **128**, 683–692. \
Kaelin, W. G. Jr. (2007). The VHL tumour suppressor protein: O$_2$ sensing and cancer. *Nature Reviews Cancer* **8**, 865–873. \
Peters, T. J., et al. (2015). De novo identification of differentially methylated regions from Illumina 450K data (DMRcate). *Epigenetics & Chromatin* **8**, 6. \
Ritchie, M. E., et al. (2015). limma powers differential expression analyses for RNA-sequencing and microarray studies. *Nucleic Acids Research* **43**, e47. \
The Cancer Genome Atlas Research Network (2013). Comprehensive molecular characterization of clear cell RCC. *Nature* **499**, 43–49.
