---
title: "P8131 Homework 9"
author: "Yongyan Liu (yl6107)"
date: "2026-04-24"
output:
  pdf_document:
    latex_engine: xelatex
header-includes:
  - \usepackage{amsmath}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 6, fig.height = 4)
library(survival)
library(KMsurv)
```

# Problem 1

**Determine the survival and density functions for a continuous survival
time variable with hazard function**
$$
h(x) = \frac{2x}{1 + x^{2}}, \qquad x \ge 0.
$$
**Hint: consider the derivative of $\log(1 + x^{2})$.**

From Lecture 21 the survival function is recovered from the hazard via
$$
S(x) = \exp\!\bigl\{-H(x)\bigr\}, \qquad
H(x) = \int_{0}^{x} h(u)\,du,
$$
Using the hint, observe that $h$ is exactly the derivative of $\log(1 + u^{2})$.
$$
H(x) \;=\; \int_{0}^{x}\!\frac{2u}{1+u^{2}}\,du
       \;=\; \Bigl[\log(1 + u^{2})\Bigr]_{0}^{x}
       \;=\; \log(1 + x^{2}).
$$
$$
S(x) \;=\; \exp\!\bigl\{-H(x)\bigr\}
      \;=\; \exp\!\bigl\{-\log(1 + x^{2})\bigr\}
      \;=\; \frac{1}{1 + x^{2}}, \qquad x \ge 0.
$$
Using $f(x) = h(x)\,S(x)$,
$$
f(x) \;=\; \frac{2x}{1+x^{2}} \cdot \frac{1}{1+x^{2}}
       \;=\; \frac{2x}{(1 + x^{2})^{2}}, \qquad x \ge 0.
$$
# Problem 2

**For the data $1,\,2,\,2,\,4^{+},\,5^{+},\,6,\,7^{+},\,8^{+},\,9^{+},\,10^{+}$
(where $+$ denotes right censoring), write out the data table and
calculate (a) the Kaplan--Meier estimate of $S(t)$, (b) the
Nelson--Aalen estimate of $H(t)$, and (c) the Fleming--Harrington
estimate of $S(t)$.**

See the table below:

\begin{center}
\begin{tabular}{cccccccc}
\hline
$t_i$ & $d_i$ & $c_i$ & $n_i$ & $\hat\lambda_i = d_i/n_i$ &
$\widehat S_{\mathrm{KM}}(t_i)$ & $\widetilde H(t_i)$ &
$\widehat S_{\mathrm{FH}}(t_i)$ \\
\hline
1  & 1 & 0 & 10 & $1/10 = 0.1000$         & $0.9000$ & $0.1000$ & $0.9048$ \\
2  & 2 & 0 & 9  & $2/9  \approx 0.2222$   & $0.7000$ & $0.3222$ & $0.7245$ \\
4  & 0 & 1 & 7  & $0$                     & $0.7000$ & $0.3222$ & $0.7245$ \\
5  & 0 & 1 & 6  & $0$                     & $0.7000$ & $0.3222$ & $0.7245$ \\
6  & 1 & 0 & 5  & $1/5  = 0.2000$         & $0.5600$ & $0.5222$ & $0.5932$ \\
7  & 0 & 1 & 4  & $0$                     & $0.5600$ & $0.5222$ & $0.5932$ \\
8  & 0 & 1 & 3  & $0$                     & $0.5600$ & $0.5222$ & $0.5932$ \\
9  & 0 & 1 & 2  & $0$                     & $0.5600$ & $0.5222$ & $0.5932$ \\
10 & 0 & 1 & 1  & $0$                     & $0.5600$ & $0.5222$ & $0.5932$ \\
\hline
\end{tabular}
\end{center}

# Problem 3

**Use the `tongue` data in the R package `KMsurv`. For each tumor type
(aneuploid and diploid), plot the Kaplan--Meier curve of the survival
function and its pointwise 95% confidence intervals (using the log
transformation). What are the estimated 1-year survival rate and 95%
CI?**

The `tongue` data set records survival of 80 patients with cancer of
the tongue, classified by tumor DNA profile: `type = 1` is aneuploid
($n_1 = 52$) and `type = 2` is diploid ($n_2 = 28$); `time` is in
**weeks** and `delta` is the death indicator. One year therefore
corresponds to $t = 52$ weeks.

```{r}
data(tongue)
str(tongue)
table(tongue$type)

aneu <- subset(tongue, type == 1)
dipl <- subset(tongue, type == 2)

fit_aneu <- survfit(Surv(time, delta) ~ 1, data = aneu, conf.type = "log")
fit_dipl <- survfit(Surv(time, delta) ~ 1, data = dipl, conf.type = "log")
```

## KM curves with log-transform 95% CIs

```{r km-aneu, fig.cap="Kaplan--Meier survival for aneuploid tumors with pointwise 95\\% log-transform CIs. Tick marks indicate right-censored observations; the dotted vertical line marks $t = 52$ weeks (1 year)."}
plot(fit_aneu, conf.int = TRUE, mark.time = TRUE,
     xlab = "Time (weeks)", ylab = "Survival probability",
     main = "Aneuploid (type = 1)")
abline(v = 52, lty = 3, col = "grey50")
```

```{r km-dipl, fig.cap="Kaplan--Meier survival for diploid tumors with pointwise 95\\% log-transform CIs. Tick marks indicate right-censored observations; the dotted vertical line marks $t = 52$ weeks (1 year)."}
plot(fit_dipl, conf.int = TRUE, mark.time = TRUE,
     xlab = "Time (weeks)", ylab = "Survival probability",
     main = "Diploid (type = 2)")
abline(v = 52, lty = 3, col = "grey50")
```

## 1-year survival rate and 95% CI

```{r}
summary(fit_aneu, times = 52)
summary(fit_dipl, times = 52)
```

- **Aneuploid tumors (type = 1).** The estimated 1-year (52-week)
  survival rate is $\widehat S(52) = 0.654$, with 95\% log-transform
  CI $(0.537,\; 0.797)$.
- **Diploid tumors (type = 2).** The estimated 1-year (52-week)
  survival rate is $\widehat S(52) = 0.486$, with 95\% log-transform
  CI $(0.330,\; 0.716)$.

Patients with aneuploid tumors have a noticeably higher estimated
1-year survival probability than patients with diploid tumors,
although the diploid CI is wide (smaller subgroup, $n_2 = 28$) and
the two confidence bands overlap.
