---
title: "Data Science II Homework 4"
author: "Yongyan Liu (yl6107)"
date: "2026-04-14"
output:
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

```{r libraries}
library(tidyverse)
library(rsample)
library(caret)
library(rpart)
library(rpart.plot)
library(ranger)
library(gbm)
library(vip)
library(pROC)
```

# Problem 1

In this exercise, we will build tree-based models using the `College` data (see "College.csv" in Homework 2). The response variable is the out-of-state tuition (`Outstate`). Partition the dataset into two parts: training data (80%) and test data (20%).

```{r p1-data}
college <- read.csv("College.csv") %>%
  dplyr::select(-College)                # drop the college name column

# If the dataset happens to include the Private factor column, convert it.
if ("Private" %in% names(college)) college$Private <- factor(college$Private)

set.seed(2026)
split1 <- initial_split(college, prop = 0.8)
train1 <- training(split1)
test1  <- testing(split1)
```

## (a) Build a regression tree on the training data to predict the response. Create a plot of the tree.

We fit a regression tree with `caret::train(method = "rpart")` and tune the complexity parameter `cp` by 10-fold cross-validation, then plot the resulting tree.

```{r p1a}
ctrl1 <- trainControl(method = "cv", number = 10)

set.seed(2026)
rpart.fit <- train(Outstate ~ .,
                   data   = train1,
                   method = "rpart",
                   tuneGrid = data.frame(cp = exp(seq(-8, -2, length = 50))),
                   trControl = ctrl1)

ggplot(rpart.fit, highlight = TRUE)
rpart.fit$bestTune

rpart.plot(rpart.fit$finalModel)
```

The CV-selected tree splits first on `Expend`, consistent with instructional expenditure being the strongest single predictor of out-of-state tuition; subsequent splits involve `Room.Board`, `Grad.Rate`, and `Apps`.

## (b) Perform random forest on the training data. Report the variable importance and the test error.

We fit a random forest via `caret`+`ranger`, tuning `mtry` and `min.node.size` by 10-fold CV, and set `importance = "permutation"` to obtain permutation-based variable importance.

```{r p1b}
rf.grid <- expand.grid(mtry = seq(2, 16, by = 2),
                       splitrule = "variance",
                       min.node.size = c(2, 5, 10))

set.seed(2026)
rf.fit <- train(Outstate ~ .,
                data      = train1,
                method    = "ranger",
                tuneGrid  = rf.grid,
                trControl = ctrl1,
                importance = "permutation")

ggplot(rf.fit, highlight = TRUE)
rf.fit$bestTune

# Variable importance
vip(rf.fit$finalModel)

# Test error
rf.pred   <- predict(rf.fit, newdata = test1)
rf.rmse   <- RMSE(rf.pred, test1$Outstate)
rf.mse    <- rf.rmse^2
tibble(metric = c("Test RMSE", "Test MSE"), value = c(rf.rmse, rf.mse))
```

The most important predictors are `Expend`, `Room.Board`, and `Terminal`/`PhD`, consistent with the regression tree. The test RMSE is reported above.

## (c) Perform boosting on the training data. Report the variable importance and the test error.

We fit a gradient boosting model using `caret::train(method = "gbm")` with `distribution = "gaussian"` and tune over `n.trees`, `interaction.depth`, `shrinkage`, and `n.minobsinnode` using 10-fold CV.

```{r p1c}
gbm.grid <- expand.grid(n.trees = c(1000, 2000, 3000, 5000),
                        interaction.depth = 1:4,
                        shrinkage = c(0.005, 0.01),
                        n.minobsinnode = 10)

set.seed(2026)
gbm.fit <- train(Outstate ~ .,
                 data       = train1,
                 method     = "gbm",
                 tuneGrid   = gbm.grid,
                 trControl  = ctrl1,
                 distribution = "gaussian",
                 verbose    = FALSE)

ggplot(gbm.fit, highlight = TRUE)
gbm.fit$bestTune

# Variable importance
par(mar = c(4, 8, 2, 2))
summary(gbm.fit$finalModel, las = 1, cBars = 19, cex.names = 0.7)

# Test error
gbm.pred  <- predict(gbm.fit, newdata = test1)
gbm.rmse  <- RMSE(gbm.pred, test1$Outstate)
gbm.mse   <- gbm.rmse^2
tibble(metric = c("Test RMSE", "Test MSE"), value = c(gbm.rmse, gbm.mse))
```

Boosting identifies `Expend`, `Room.Board`, and `Grad.Rate` as the top predictors. The test RMSE is reported above and is generally comparable to the random forest.

\newpage

# Problem 2

This problem is based on the data `auto.csv` in Homework 3. Split the dataset into two parts: training data (70%) and test data (30%).

```{r p2-data}
auto <- read.csv("auto.csv") %>%
  mutate(mpg_cat = factor(mpg_cat, levels = c("low", "high")),
         origin  = factor(origin))

set.seed(2026)
split2 <- initial_split(auto, prop = 0.7, strata = mpg_cat)
train2 <- training(split2)
test2  <- testing(split2)
```

## (a) Build a classification tree using the training data, with `mpg_cat` as the response. Which tree size corresponds to the lowest cross-validation error? Is this the same as the tree size obtained using the 1 SE rule?

We grow a large classification tree with `rpart` and use its internal 10-fold cross-validation to pick the best `cp`. We then compare the tree size (number of terminal nodes) selected by (i) the minimum CV error and (ii) the 1-SE rule.

```{r p2a}
set.seed(2026)
tree2 <- rpart(mpg_cat ~ .,
               data   = train2,
               method = "class",
               control = rpart.control(cp = 0, xval = 10, minsplit = 10))

printcp(tree2)
plotcp(tree2)

cptab <- tree2$cptable

# Minimum CV-error tree
idx.min <- which.min(cptab[, "xerror"])
cp.min  <- cptab[idx.min, "CP"]
size.min <- cptab[idx.min, "nsplit"] + 1   # terminal nodes = nsplit + 1

# 1-SE rule: smallest tree with xerror <= min(xerror) + corresponding xstd
xerr.min <- cptab[idx.min, "xerror"]
xstd.min <- cptab[idx.min, "xstd"]
idx.1se  <- min(which(cptab[, "xerror"] <= xerr.min + xstd.min))
cp.1se   <- cptab[idx.1se, "CP"]
size.1se <- cptab[idx.1se, "nsplit"] + 1

tibble(rule = c("min CV", "1-SE"),
       cp   = c(cp.min, cp.1se),
       size = c(size.min, size.1se))

# Prune the two trees and plot the 1-SE tree
tree.min <- prune(tree2, cp = cp.min)
tree.1se <- prune(tree2, cp = cp.1se)

rpart.plot(tree.min, main = "Min CV-error tree")
rpart.plot(tree.1se, main = "1-SE rule tree")
```

```{r p2a-answer, echo=FALSE, results='asis'}
cat(sprintf(
  "The lowest CV error corresponds to a tree with **%d** terminal nodes, while the 1-SE rule selects a tree with **%d** terminal nodes. They are **%s**.",
  size.min, size.1se,
  ifelse(size.min == size.1se, "the same", "different")))
```

## (b) Perform boosting on the training data and report the variable importance. Report the test data performance.

We fit a binomial gradient boosting machine via `caret::train(method = "gbm")` with `twoClassSummary` so that ROC is used for CV tuning, following the L9 `caret` convention.

```{r p2b}
ctrl2 <- trainControl(method = "cv",
                      number = 10,
                      summaryFunction = twoClassSummary,
                      classProbs = TRUE)

gbm.grid2 <- expand.grid(n.trees = c(1000, 2000, 3000, 5000),
                         interaction.depth = 1:4,
                         shrinkage = c(0.005, 0.01),
                         n.minobsinnode = 10)

set.seed(2026)
gbm.fit2 <- train(mpg_cat ~ .,
                  data       = train2,
                  method     = "gbm",
                  tuneGrid   = gbm.grid2,
                  trControl  = ctrl2,
                  metric     = "ROC",
                  distribution = "bernoulli",
                  verbose    = FALSE)

ggplot(gbm.fit2, highlight = TRUE)
gbm.fit2$bestTune

# Variable importance
par(mar = c(4, 8, 2, 2))
summary(gbm.fit2$finalModel, las = 1, cBars = 8, cex.names = 0.8)
```

```{r p2b-test}
# Test data performance: classification error, confusion matrix, AUC
gbm.pred.class <- predict(gbm.fit2, newdata = test2)
gbm.pred.prob  <- predict(gbm.fit2, newdata = test2, type = "prob")[, "high"]

cm <- confusionMatrix(data      = gbm.pred.class,
                      reference = test2$mpg_cat,
                      positive  = "high")
cm

test.err <- mean(gbm.pred.class != test2$mpg_cat)

roc.gbm <- roc(test2$mpg_cat, gbm.pred.prob, levels = c("low", "high"))
plot(roc.gbm, legacy.axes = TRUE, print.auc = TRUE)

tibble(metric = c("Test classification error", "Test accuracy", "Test AUC"),
       value  = c(test.err, as.numeric(cm$overall["Accuracy"]), as.numeric(roc.gbm$auc)))
```

The most influential variables for classifying `mpg_cat` are `displacement`, `weight`, and `horsepower`. The test accuracy and test AUC are reported above; boosting achieves strong performance on the held-out 30% test set.
